Regulation of hippocampal mossy fiber-CA3 synapse function by a Bcl11b/C1ql2/Nrxn3(25b+) pathway

Version 1

Abstract

Abstract

The transcription factor Bcl11b has been linked to neurodevelopmental and neuropsychiatric disorders associated with synaptic dysfunction. Bcl11b is highly expressed in dentate gyrus granule neurons and is required for the structural and functional integrity of mossy fiber-CA3 synapses. The underlying molecular mechanisms, however, remained unclear. We show that the synaptic organizer molecule C1ql2 is a direct functional target of Bcl11b that regulates synaptic vesicle recruitment and long-term potentiation at mossy fiber-CA3 synapses in vivo and in vitro. Furthermore, we demonstrate C1ql2 to exert its functions through direct interaction with a specific splice variant of neurexin-3, Nrxn3(25b+). Interruption of C1ql2-Nrxn3(25b+) interaction by expression of a non-binding C1ql2 mutant or by deletion of Nrxn3 in the dentate gyrus granule neurons recapitulates major parts of the Bcl11b as well as C1ql2 mutant phenotype, and interferes with C1ql2 targeting to the synapse. Together, this study identifies a novel C1ql2-Nrxn3(25b+)-dependent signaling pathway through which Bcl11b controls mossy fiber-CA3 synapse function. Thus, our findings contribute to the mechanistic understanding of neurodevelopmental disorders accompanied by synaptic dysfunction.

Introduction

Disruptions in synaptic structure and function have been identified as a major determinant in the manifestation of various neurodevelopmental and neuropsychiatric disorders, such as autism spectrum disorder, schizophrenia and intellectual disability (Hayashi-Takagi, 2017; Lepeta et al., 2016; Zoghbi & Bear, 2012). The need to understand how synaptic function is compromised in these disorders has accentuated the importance of studying the regulatory mechanisms of physiological synaptic function. These mechanisms involve cell adhesion molecules at both the pre- and post-synaptic side that act as synaptic organizers, whose unique combination determines the structural and functional properties of the synapse. Many such proteins have already been identified and our understanding of their complex role in synapse assembly and function has significantly increased over the last years (de Wit & Ghosh, 2016; O’Rourke et al., 2012; Sudhof, 2017). Furthermore, recent advances in the genetics of neurodevelopmental and neuropsychiatric disorders have implicated genes encoding for several of the known synaptic proteins, supporting a role for these molecules in the pathogenesis of corresponding disorders (Sudhof, 2021; Torres et al., 2017; Wang et al., 2018).

The sensitivity of the functional specification of the synapse to the combination of distinct synaptic proteins and their relative expression levels suggests that genetically encoded programs define at least facets of the synaptic properties in a cell-type specific manner (Sudhof, 2017, 2021). In fact, several of the synaptic proteins have been shown to promote formation of functional pre- and postsynaptic assemblies when presented in non-neuronal cells (Dalva et al., 2000; Dean et al., 2003; Scheiffele et al., 2000), showing that their ability to specify synapses is in part independent of signaling processes and neuronal activity and supporting the idea that synaptic function is governed by cues linked to cellular origin (Gomez et al., 2021). Thus, the investigation for synaptic organizers and their function in health and disease should be expanded to the transcriptional programs that regulate their expression.

Bcl11b (also known as Ctip2) is a zinc finger transcription factor that has been implicated in various disorders of the nervous system including Alzheimer’s and Huntington’s disease, and schizophrenia (Kunkle et al., 2016; Song et al., 2022; Whitton et al., 2018; Whitton et al., 2016). Patients with BCL11B mutations present with neurodevelopmental delay, overall learning deficits as well as impaired speech acquisition and autistic features (Eto et al., 2022; Lessel et al., 2018; Punwani et al., 2016; Yang et al., 2020). Bcl11b is expressed in several neuron types, including the dentate gyrus granule neurons (DGN) of the hippocampus. Expression of Bcl11b in the DGN starts during embryonic development and persists into adulthood (Simon et al., 2020). We have previously demonstrated that Bcl11b plays a crucial role in the development of the hippocampal mossy fiber system, the adult hippocampal neurogenesis as well as hippocampal learning and memory (Simon et al., 2016; Simon et al., 2012). In the mature hippocampus, Bcl11b is critical for the structural and functional integrity of mossy fiber synapses (MFS), the connection between DGN and CA3 pyramidal neurons (De Bruyckere et al., 2018). MFS have a critical role in learning and memory stemming from their unique structural and functional properties, such as an enormous pool of releasable synaptic vesicles (SV), and reliable presynaptic short- and long-term plasticity (Nicoll & Schmitz, 2005; Rollenhagen & Lubke, 2010). Conditional ablation of Bcl11b in murine DGN impairs presynaptic recruitment of SV and abolishes mossy fiber long-term potentiation (MF-LTP) (De Bruyckere et al., 2018). The molecular mechanisms, however, through which the transcriptional regulator Bcl11b controls highly dynamic properties of the MFS remained elusive.

In the present study, we show that the secreted synaptic organizer molecule C1ql2, a member of the C1q-like protein family (Yuzaki, 2017), is a functional target of Bcl11b in murine DGN. Reintroduction of C1ql2 in Bcl11b mutant DGN rescued the localization and docking of SV to the active zone (AZ), as well as MF-LTP that was abolished upon Bcl11b ablation. Knock-down (KD) of C1ql2 in wildtype animals recapitulated major part of the MFS phenotype observed in Bcl11b mutants. Furthermore, we show that C1ql2 requires direct interaction with a specific neurexin-3 isoform, Nrxn3(25b+), a member of a polymorphic family of presynaptic cell adhesion molecules (Reissner et al., 2013; Sudhof, 2017), to recruit SV in vitro and in vivo. Finally, we demonstrate that the C1ql2-Nrxn3(25b+) interaction is necessary for the targeting of C1ql2 to the MFS. Taken together, this study identifies a novel Bcl11b/C1ql2/Nrxn3(25b+)-dependent regulatory mechanism that is essential for the control of MFS function. Recent genetic studies suggested its single components to be associated with neurodevelopmental and neuropsychiatric disorders characterized by synaptic dysfunction. Our data, for the first time, demonstrate these molecules to be interconnected in one regulatory pathway. Thus, our findings provide new mechanistic insight into the pathogenesis of corresponding human disorders.

Results

Reintroduction of C1ql2 into Bcl11b mutant dentate granule neurons restores synaptic vesicle recruitment at the mossy fiber-CA3 synapse

We demonstrated before that Bcl11b is critical for the structural and functional integrity of adult excitatory hippocampal MFS (De Bruyckere et al., 2018; Simon et al., 2012). The downstream regulatory mechanisms, however, through which Bcl11b exerts its complex functions at the MFS remained unclear. Using differential transcriptomic analyses on Bcl11b conditional knock-out (cKO) and wildtype (WT) DGN we screened for candidate transcriptional targets of this transcription factor. Among these, we identified the synaptic organizer molecule C1ql2, previously implicated in modulating MFS functions (Matsuda et al., 2016). C1ql2 transcript and protein levels are massively downregulated in Bcl11b mutant DGN (Supplementary figure 1a-b), and the Bcl11b protein directly binds to consensus sequences within the C1ql2 promotor (De Bruyckere et al., 2018), suggesting Bcl11b to act on MFS through C1ql2. To directly test this, we stereotaxically injected a C1ql2-expressing AAV (Figure 1a-b) into the dentate gyrus (DG) of Bcl11b cKO mice 2 weeks after induction of the mutation. AAV-mediated re-expression of C1ql2 in the Bcl11b mutant DGN completely restored C1ql2 protein expression to control levels (Figure 1c-d; Control+EGFP: 1±0.216, Bcl11b cKO+EGFP: 0.2±0.023, Bcl11b cKO+EGFP-2A-C1ql2: 2.44±0.745, mean±SEM). Furthermore, the spatial distribution of the exogenous C1ql2 protein in mutants was indistinguishable from controls (Figure 1e). Using vGlut1 and Homer1 as pre- and postsynaptic markers, respectively, we observed exogenous C1ql2 protein to precisely localize at glutamatergic synapses within the stratum lucidum (SL) of CA3, confirming that reintroduced C1ql2 is correctly targeted to the MFS (Figure 1f).

Stereotaxic injection of C1ql2-expressing AAV into Bcl11b cKO DGN restores C1ql2 levels

a Experimental design to analyze the functions of C1ql2 in the MFS as a downstream target of Bcl11b. b AAV constructs injected in the DG of Bcl11b cKO and control littermates. c Western blot and d relative C1ql2 protein levels in mouse hippocampal homogenates. n=3. All data are presented as means; error bars indicate SEM. Two-way ANOVA and Tuckey’s PHC. Bcl11b cKO+EGFP-2A-C1ql2 vs. Bcl11b cKO+EGFP: *p=0.015; Bcl11b cKO+EGFP-2A-C1ql2 vs. Bcl11b cKO+EGFP-2A-C1ql3: *p=0.019; ns, not significant. e Immunohistochemistry of C1ql2 (cyan), GFP (green) and Bcl11b (red) on hippocampal sections. Scale bar: 200 μm. f Immunohistochemistry of C1ql2 (cyan), vGlut1 (magenta) and Homer1 (yellow) in the SL of CA3. White arrowheads indicate co-localizing puncta of all three proteins. Scale bar: 15 μm

MFS of Bcl11b cKO animals were characterized by a misdistribution of SV in relation to the AZ, with fewer SV being present in the vicinity of AZ (Figure 2a), as reflected by a lower average synapse score (Figure 2b; c.f. De Bruyckere et al., 2018). Reintroduction of C1ql2 fully recovered the synapse score to control values (Figure 2a-b; Control+EGFP: 3.4±0.012, Bcl11b cKO+EGFP: 2.96±0.037, Bcl11b cKO+EGFP-2A-C1ql2: 3.47±0.043, mean±SEM). As revealed by the relative frequency of the individual synapse scores, C1ql2 did not only reduce the number of inactive synapses, characterized by a synapse score of 0 but also improved the synapse score of active synapses (Figure 2c). To test for the specificity of the C1ql2 effects, we overexpressed C1ql3 in the DGN of Bcl11b cKO (Figure 1a-b). C1ql3, a different member of the C1ql subfamily, is co-expressed with C1ql2 in DGN and the two proteins have been shown to form functional heteromers (Matsuda et al., 2016). C1ql3 expression is unchanged in Bcl11b cKO (Supplementary figure 2a-b; Control+EGFP: 1±0.022, Bcl11b cKO+EGFP: 1.09±0.126, Bcl11b cKO+EGFP-2A-C1ql2: 0.87±0.146, mean±SEM). Overexpression of C1ql3 in the DGN of Bcl11b cKO neither interfered with C1ql2 expression levels (Figure 1c-e; Bcl11b cKO+EGFP-2A-C1ql3: 0.29±0.042, mean ± SEM) nor was it able to rescue the synapse score of Bcl11b mutants (Figure 2a-c; Bcl11b cKO+EGFP-2A-C1ql3: 2.97±0.062, mean±SEM).

C1ql2 reintroduction in Bcl11b cKO DGN rescues SV recruitment in MFS

a Electron microscope images of MFS and proximal SV. White bars mark synapse length from the postsynaptic side. Scale bar: 100 nm. b Average synapse score. Control+EGFP, n=3; Bcl11b cKO+EGFP, Bcl11b cKO+EGFP-2A-C1ql2, Bcl11b cKO+EGFP-2A-C1ql3, n=4. Two-way ANOVA and Tuckey’s PHC. Control+EGFP vs. Bcl11b cKO+EGFP: ***p=0.0002, and vs. Bcl11b cKO+EGFP-2A-C1ql3: ***p=0.0003; Bcl11b cKO+EGFP-2A-C1ql2 vs. Bcl11b cKO+EGFP and vs. Bcl11b cKO+EGFP-2A-C1ql3: ***p<0.0001; ns, not significant. c Relative frequency of synapse scores. d Number of docked vesicles per 100 nm AZ profile length. Control+EGFP, Bcl11b cKO+EGFP-2A-C1ql3, n=3; Bcl11b cKO+EGFP, Bcl11b cKO+EGFP-2A-C1ql2, n=4. All data are presented as means; error bars indicate SEM. Points represent the individual examined AZ and SV, respectively. Two-way ANOVA and Tuckey’s PHC. Control+EGFP vs. Bcl11b cKO+EGFP: *p=0.024, and vs. Bcl11b cKO+EGFP-2A-C1ql3: *p=0.045; Bcl11b cKO+EGFP-2A-C1ql2 vs. Bcl11b cKO+EGFP: *p=0.026, and vs. Bcl11b cKO+EGFP-2A-C1ql3: *p=0.049; ns, not significant. e Diameter of docked vesicles. Control+EGFP, Bcl11b cKO+EGFP-2A-C1ql3, n=3; Bcl11b cKO+EGFP, Bcl11b cKO+EGFP-2A-C1ql2, n=4; Two-way ANOVA. ns, not significant. f Cumulative and g relative frequency of the number of docked vesicles per 100 nm AZ profile length.

To analyze the C1ql2-dependent functions of Bcl11b on SV distribution in more detail, we quantified the number of SV docked on AZ in control animals, Bcl11b mutants, and upon the reintroduction of C1ql2. SV with a ≤ 5 nm distance from the plasma membrane were considered docked (Kusick et al., 2022; Vandael et al., 2020). Bcl11b mutant animals had significantly fewer docked vesicles per 100 nm of AZ profile length compared to control animals and more AZ with no docked vesicles at all. Rescue of C1ql2 expression restored the number of docked SV to control levels, while the overexpression of C1ql3 did not affect the number of docked vesicles (Figure 2d, f-g; Control+EGFP: 0.53±0.098, Bcl11b cKO+EGFP: 0.24±0.038, Bcl11b cKO+EGFP-2A-C1ql2: 0.51±0.049, Bcl11b cKO+EGFP-2A-C1ql3: 0.26±0.041, mean±SEM). In contrast to vesicle distribution, Bcl11b and C1ql2 did not affect the size of the docked SV (Figure 2e; Control+EGFP: 36.39±1.75, Bcl11b cKO+EGFP: 35.18±1.13, Bcl11b cKO+EGFP-2A-C1ql2: 36.35±1.01, Bcl11b cKO+EGFP-2A-C1ql3: 36.65±0.1, mean±SEM). Thus, our data suggest that C1ql2 specifically controls SV recruitment downstream of Bcl11b.

Conditional deletion of Bcl11b in the adult hippocampus also leads to a loss of MFS, as well as reduced ultrastructural complexity of the remaining mossy fiber boutons (MFB) (De Bruyckere et al., 2018), posing the question of whether these phenotypic features also depend on C1ql2. Interestingly, C1ql2 reintroduction in the DGN of Bcl11b cKO neither restored the loss of glutamatergic synapses, as quantified by the colocalization of pre- and postsynaptic markers, vGlut1 and Homer1 (Supplementary figure 2e; Control+EGFP: 90.65±8.25, Bcl11b cKO+EGFP: 60.68±4.62, Bcl11b cKO+EGFP-2A-C1ql2: 56.84±6.99, mean±SEM), nor the reduced ultrastructural complexity of MFB, as quantified by the MFB perimeter/area ratio (Supplementary figure 2c-d; Control+EGFP: 0.0051±0.00031, Bcl11b cKO+EGFP: 0.0042±0.00014, Bcl11b cKO+EGFP-2A-C1ql2: 0.0037±0.00021, mean±SEM). This suggests that Bcl11b acts on MFS through C1ql2-dependent as well as -independent signaling pathways.

Reintroduction of C1ql2 into Bcl11b mutant dentate granule neurons rescues mossy fiber synapse long-term potentiation

The ultrastructural changes at the MFS point towards potential alterations in synaptic function. Indeed, adult-induced Bcl11b cKO was previously found to result in a loss of MF-LTP (De Bruyckere et al., 2018). We therefore tested whether the reintroduction of C1ql2 in Bcl11b cKO DGN can rescue LTP at the mutant MFS, similarly to SV recruitment. We stimulated mossy fibers in acute slices and measured the resulting field potentials in the SL of CA3. Field responses were carefully validated for the specificity of mossy fiber signals by the presence of strong paired-pulse facilitation and block by the mGluR antagonist DCG-IV. Under these conditions, input-output curves of fEPSP slopes versus axonal fiber volleys revealed no significant differences between control and Bcl11b cKO mice (Supplementary figure 2f-g; Control+EGFP: 1.95±0.09, Bcl11b cKO+EGFP: 2.02±0.14, mean±SEM), indicating that basal synaptic transmission was unaltered in the Bcl11b mutants. We then induced LTP by high-frequency stimulation (HFS) of mossy fibers in control and Bcl11b cKO animals with or without AAV-mediated expression of C1ql2. Compared to controls, Bcl11b mutants injected with the control AAV displayed a strong reduction of LTP at 20-30 and 30-40 min after induction (Figure 3a-c; 0-10 min: Control+EGFP: 90.4±7.2, Bcl11b cKO+EGFP: 106.1±10.8, 10-20 min: Control+EGFP: 42.7±3.6, Bcl11b cKO+EGFP: 39.5±4.6, 20-30 min: Control+EGFP: 52.5±7.6, Bcl11b cKO+EGFP: 24.8±3.2, 30-40 min: Control+EGFP: 50.1±7.3, Bcl11b cKO+EGFP: 20.3±3.7, mean±SEM), consistent with our previous data (De Bruyckere et al., 2018). The loss of LTP was completely reversed upon the re-expression of C1ql2, with the rescue mice exhibiting comparable LTP to controls at all time intervals (Figure 3a-c; 0-10 min: Control+EGFP: 90.4±7.2, Bcl11b cKO+EGFP-2A-C1ql2: 86.5±7.4, 10-20 min: Control+EGFP: 42.7±3.6, Bcl11b cKO+EGFP-2A-C1ql2: 49.4±5.9, 20-30 min: Control+EGFP: 52.5±7.6, Bcl11b cKO+EGFP-2A-C1ql2: 47.2±5.7, 30-40 min: Control+EGFP: 50.1±7.3, Bcl11b cKO+EGFP-2A-C1ql2: 44.9±5.3, mean±SEM). Importantly, this rescue effect was specific to C1ql2 as the overexpression of C1ql3 failed to reverse the Bcl11b cKO phenotype (Figure 3a-c; 0-10 min: Control+EGFP: 90.4±7.2, Bcl11b cKO+EGFP-2A-C1ql3: 104.2±9.9, 10-20 min: Control+EGFP: 42.7±3.6, Bcl11b cKO+EGFP-2A-C1ql3: 44.4±5.7, 20-30 min: Control+EGFP: 52.5±7.6, Bcl11b cKO+EGFP-2A-C1ql3: 29.0±2.3, 30-40 min: Control+EGFP: 50.1±7.3, Bcl11b cKO+EGFP-2A-C1ql3: 22.6±2.4, mean±SEM).

C1ql2 reintroduction in Bcl11b cKO DGN rescues mossy fiber LTP

a Representative fEPSP traces showing baselines before HFS (black), fEPSP changes 30-40 min after HFS (cyan) and following the application of 3 µM DCG-IV (red). b Time course of fEPSP slopes. The black arrow indicates HFS and the dashed line the baseline level. c Quantification of fEPSP facilitation at four different time intervals after HFS. Changes in fEPSP slope are shown as percentage of the mean baseline fEPSP. Control+EGFP, n=7; Bcl11b cKO+EGFP, Bcl11b cKO+EGFP-2A-C1ql3, n=8; Bcl11b cKO+EGFP-2A-C1ql2, n=6; All data are presented as means; error bars indicate SEM. One-way ANOVA followed by Bonferroni’s PHC for each time interval. 20-30 min: Control+EGFP vs. Bcl11b cKO+EGFP: **p=0.002, and vs. Bcl11b cKO+EGFP-2A-C1ql3: *p=0.011; Bcl11b cKO+EGFP-2A-C1ql2 vs. Bcl11b cKO+EGFP: *p=0.023; 30-40 min: Control+EGFP vs. Bcl11b cKO+EGFP: ***p<0.001, and vs. Bcl11b cKO+EGFP-2A-C1ql3: **p=0.002; Bcl11b cKO+EGFP-2A-C1ql2 vs. Bcl11b cKO+EGFP: *p=0.01 and vs. Bcl11b cKO+EGFP-2A-C1ql3: *p=0.0523; ns, not significant. d Representative fEPSP traces showing baselines before forskolin application (black), fEPSP changes 105-120 min after the start of application (cyan) and following the addition of 3 µM DCG-IV (red). e Time course of fEPSP slopes. The black solid line indicates forskolin perfusion and dashed line the baseline level. f Quantification of fEPSP facilitation at two different time intervals after the start of forskolin application. Changes in fEPSP slope are shown as percentage of the mean baseline fEPSP. n=8. All data are presented as means; error bars indicate SEM. Unpaired t-test for both time intervals. 15-30 min: **p=0.005; 105-120 min: **p=0.0025.

MF-LTP is known to be mediated by a cAMP-dependent signaling cascade. To test whether Bcl11b acts on LTP through the cAMP-dependent pathway, we directly activated the cAMP pathway in slices from control and Bcl11b cKO mice by applying the adenylyl cyclase (AC) activator forskolin (Weisskopf et al., 1994). Compared to slices from control animals, forskolin-induced LTP in the mutants had a significantly lower peak and remained significantly weaker throughout the recording (Figure 3d-f; 15-30 min: Control: 342.4±28.0, Bcl11b cKO: 232.3±17.7, 105-120 min: Control: 88.0±7.4, Bcl11b cKO: 47.2±8.2, mean±SEM). This suggests that the regulation of MF-LTP by Bcl11b involves the cAMP-dependent signaling pathway.

Knock-down of C1ql2 in dentate granule neurons perturbs synaptic vesicle recruitment and long-term potentiation at the mossy fiber-CA3 synapse

To further corroborate the observation that Bcl11b acts on MFS specifically through C1ql2, we knocked down C1ql2 expression in DGN of adult WT mice by stereotaxically injecting an AAV carrying an shRNA cassette against C1ql2 (Figure 4a). Quantitative PCR (Figure 4b), western blot analysis (Figure 4c) as well as immunohistochemistry using C1ql2 antibodies on hippocampal tissue (Figure 4d) revealed that the shRNA-mediated KD resulted in a strong reduction of the C1ql2 transcript (Figure 4b; +shNS-EGFP: 1±0.07, +shC1ql2-EGFP: 0.23±0.059, mean±SEM) as well as protein levels (Figure 4c-d), as compared to animals injected with the control AAV. The shRNA-mediated KD of C1ql2 had no effect on the expression of C1ql3, demonstrating the specificity of this approach (Supplementary figure 3a-b; +shNS-EGFP: 1±0.09, +shC1ql2-EGFP: 0.986±0.035, mean±SEM). Compared to controls, C1ql2 KD was sufficient to reduce the average synapse score to similar levels as observed in Bcl11b cKO (Figure 4e, Supplementary figure 3c; +shNS-EGFP: 3.38±0.069, +shC1ql2-EGFP: 3.15±0.031, mean±SEM), as well as the number of docked vesicles per 100 nm of AZ profile length (Figure 4f, Supplementary figure 3d-e; +shNS-EGFP: 0.48±0.04, +shC1ql2-EGFP: 0.31±0.02, mean±SEM). At the same time, the diameter of the docked vesicles remained unchanged (Supplementary figure 3f; +shNS-EGFP: 34.28±0.84, +shC1ql2-EGFP: 35.37±0.21, mean±SEM). To test whether shRNA-mediated KD of C1ql2 expression also affects MF-LTP, we performed LTP recordings in C1ql2 KD and control mice. Compared to controls, slices from C1ql2 KD mice exhibited a significant reduction of LTP at 20-30 and 30-40 min time intervals, similarly to Bcl11b cKO animals (Figure 4g-i; 0-10 min: +shNS-EGFP: 105.0±4.0, +shC1ql2-EGFP: 94.3±4.5, 10-20 min: +shNS-EGFP: 56.3±4.5, +shC1ql2-EGFP: 35.1±2.8, 20-30 min: +shNS-EGFP: 50.2±4.5, +shC1ql2-EGFP: 23.4±3.5, 30-40 min: +shNS-EGFP: 44.6±4.3, +shC1ql2-EGFP: 20.1±4.1, mean±SEM). Thus, KD of C1ql2 in WT DGN recapitulates major phenotypes observed upon Bcl11b cKO, supporting that Bcl11b controls SV recruitment and LTP in hippocampal MFS specifically through C1ql2.

KD of C1ql2 in DGN of WT mice impairs SV recruitment and LTP

a Experimental design to analyze the MFS after AAV-mediated KD of C1ql2 in WT DGN. b Relative C1ql2 mRNA levels in DGN. n=4 (3). All data are presented as means; error bars indicate SEM. Unpaired t-test: ***p=0.0002. c Western blot of mouse hippocampal homogenates. d Immunohistochemistry of C1ql2 (red) and GFP (green) on hippocampal sections. Scale bar: 200 μm. e Average synapse score. n=4. Unpaired t-test. *p=0.025. f Number of docked vesicles per 100 nm AZ profile length. n=3. All data are presented as means; error bars indicate SEM. Points represent the individual examined AZ. Unpaired t-test. *p=0.018. g Representative fEPSP traces showing baselines before HFS (black), fEPSP changes 30-40 min after HFS (cyan) and following the application of 3 µM DCG-IV (red). h Time course of fEPSP slopes. The black arrow indicates HFS and the dashed line the baseline level. i Quantification of fEPSP facilitation at four different time intervals after HFS. Changes in fEPSP slope are shown as percentage of the mean baseline fEPSP. +shNS-EGFP, n=6; +shC1ql2-EGFP, n=7. All data are presented as means; error bars indicate SEM. Mann-Whitney U-test for each time interval. 10-20 min: **p=0.0012; 20-30 min: **p=0.0023; 30-40 min: **p=0.0023; ns, not significant.

C1ql2-Nrxn3(25b+) interaction recruits presynaptic vesicles in vitro and in vivo

C1ql2 was previously shown to interact with a particular splice variant of Nrxn3β containing exon 25b sequences, Nrxn3(25b+), which was recombinantly expressed in HEK293 cells (Matsuda et al., 2016). To explore whether the Bcl11b/C1ql2-dependent regulation of MFS involves interaction with neuronal Nrxn3, we co-cultured myc-C1ql2 expressing HEK293 cells with primary hippocampal neurons transfected with GFP-Nrxn3α(25b+). We used here the extracellularly longer Nrxn3α isoform because it is more strongly expressed in the murine DG compared to Nrxn3β (Uchigashima et al., 2019). C1ql2-dependent recruitment of Nrxn3α(25b+) was quantified by the surface area of HEK293 cells covered by GFP-Nrxn3α(25b+)-positive neuronal profiles (Figure 5a). HEK293 cells expressing myc-C1ql2 had a significantly larger surface area covered by neuronal Nrxn3α(25b+) in comparison to HEK293 cells expressing myc-tag only (Figure 5b; myc-C1ql2: 39.97±3.99, myc-tag: 17.29±2.27, mean±SEM). Using vGlut1 immunoreactivity as a proxy for SV localization (Aoto et al., 2007; Fremeau et al., 2004), we examined in the same system whether C1ql2-expressing HEK293 cells were able to cluster vGlut1 in surrounding GFP-Nrxn3α(25b+)-positive neurons (Figure 5c). Interestingly, vGlut1 accumulation was significantly increased in GFP-Nrxn3α(25b+)-positive neurons contacting myc-C1ql2 expressing HEK293 cells as compared to C1ql2-negative HEK293 cells (Figure 5d; myc-C1ql2: 40.88±3.25, myc-tag: 24.78±4.99, mean±SEM). To specifically analyze whether endogenous Nrxn3 is required for C1ql2-mediated vGlut1 accumulation in neurons, we co-cultured myc-C1ql2 expressing HEK293 cells with primary hippocampal neurons derived from Nrxn123flox/flox mice, in which all three Nrxn genes are floxed (Chen et al., 2017), and which we transfected with either active Cre recombinase or an inactive Cre (Klatt et al., 2021) (Figure 5e). Neurons lacking all Nrxn variants showed a significantly lower accumulation of endogenous vGlut1 when contacting the C1ql2-expressing HEK293 cells compared to control neurons. Strikingly, selective reintroduction of the Nrxn3α(25b+) isoform into the null-mutant Nrxn cKO neurons was sufficient to normalize vGlut1 accumulation in vitro (Figure 5f; inactive Cre: 51.66±5.97, Cre: 27.83±2.83, Cre+ Nrxn3α(25b+): 39.23±4.3, mean±SEM). Collectively, our data strongly suggest that the C1ql2-mediated recruitment of vGlut1-positive SV in hippocampal neurons depends on the presence of Nrxn3α(25b+).

C1ql2-Nrxn3 interaction recruits vGlut1 in vitro

a Immunocytochemistry of myc-tagged C1ql2, C1ql2.K262E or myc-tag (magenta) expressing HEK293 cells and GFP-Nrxn3α(25b+) (cyan) from contacting hippocampal neurons. Scale bar: 5 μm. b Nrxn3α(25b+) recruitment by differentially transfected HEK293 cells. n=3. All data are presented as means; error bars indicate SEM. One-way ANOVA and Tuckey’s PHC. myc-C1ql2 vs. myc-tag: *p=0.016, and vs. myc-K262E: *p=0.022; ns, not significant. c Immunocytochemistry of myc-tagged C1ql2, C1ql2.K262E or myc-tag (magenta) expressing HEK293 cells and vGlut1 (cyan) from contacting hippocampal neurons. Scale bar: 5 μm. d vGlut1 recruitment by differentially transfected HEK293 cells. n = 3. All data are presented as means; error bars indicate SEM. One-way ANOVA and Tuckey’s PHC. myc-C1ql2 vs. myc-tag: *p=0.04, and vs. myc-K262E: **p=0.007; ns: non-significant. e Immunocytochemistry of myc-tagged C1ql2 (magenta) expressing HEK293 cells and vGlut1 (cyan) from contacting control, Nrxn123 KO or Nrxn123 KO with Nrxn3α(25+) rescued hippocampal neurons. Scale bar: 5 μm. f vGlut1 recruitment by HEK293 cells in presence or absence of neuronal Nrxns. n = 3. All data are presented as means; error bars indicate SEM. One-way ANOVA and Tuckey’s PHC. inactive Cre vs. Cre: *p=0.023, and vs. Cre+ Nrxn3α(25+): p=0.21; ns: non-significant. g Trimeric structures of C1ql2 (PDB_ID: 4QPY, upper panels) and the variant C1ql2.K262E (lower panels). Residue 262 is the central residue (red, left & middle panels) of a larger area underneath the C1ql2-specific calcium and receptor binding loops (magenta, middle panel). The mutation K262E alters the charge of that surface area negative (yellow-circled area, right panels) and makes it potentially repulsive to bind Nrxn3(25b+).

To explore the relevant epitope that mediates binding of C1ql2 to Nrxn3(25b+) proteins, we analyzed the solvent accessible electrostatic surface properties of the C1ql-domain trimeric structure of C1ql2 (Ressl et al., 2015) (PDB_ID: 4QPY) and found that a change of lysine262 (K262) to glutamic acid renders a large area underneath the C1ql2-specific calcium and receptor binding loops negative (Figure 5g) and hypothesized that this would repel binding to Nrxn3(25b+). We generated a C1ql2.K262E variant, expressed it in HEK293 cells as before, and tested it for its ability to cluster Nrxn3α(25b+) as well as vGlut1 in contacting primary neurons (Figure 5a-d). In the presence of C1ql2.K262E, recruitment of Nrxn3α(25b+) was significantly lower compared to WT C1ql2 and indistinguishable from myc-tag control levels. (Figure 5a-b; myc-K262E: 18.84±5.15). Moreover, the expression of C1ql2.K262E in HEK293 cells was unable to accumulate vGlut1 in contacting neurons expressing GFP-Nrxn3α(25b+) (Figure 5c-d; myc-K262E: 16.9±1.2, mean±SEM). Together, these results provide in vitro evidence that the clustering of vGlut1 depends on an intact C1ql2-Nrxn3(25b+) interaction, and that a single point mutation that creates a negative charge of that surface area underneath the C1ql2-specific calcium and receptor binding loops abolishes this binding activity and, thereby, the regulation of SV clustering.

To validate our identification of K262 as a key residue for the C1ql2-Nrxn3(25b+) interaction in vivo, we expressed C1ql2.K262E in Bcl11b cKO DGN, in which endogenous C1ql2 expression is downregulated by the ablation of Bcl11b (Figure 6a). AAV-mediated introduction of C1ql2.K262E in Bcl11b cKO DGN resulted in strong over all expression of the mutant protein (Figure 6b-c; Control+EGFP: 1±0.42, Bcl11b cKO+EGFP-2A-K262E: 9.68±4.75, mean±SEM). However, the spatial distribution of C1ql2.K262E was notably different from the WT protein in the SL of CA3 where most of the MFS are located (Figure 6d). In the SL, expression levels of C1ql2.K262E were significantly lower compared to WT C1ql2 as quantified by the integrated fluorescence density (Figure 6d-e; Bcl11b cKO+EGFP-2A-C1ql2: 9.75±0.57 x104, Bcl11b cKO+EGFP-2A-K262E: 5.89±0.55 x104, mean±SEM). As C1ql3 has been shown to form functional heteromers with C1ql2 at the MFS, we examined the spatial distribution of the C1ql3 protein upon AAV-mediated introduction of C1ql2.K262E for potential expression pattern changes but observed no difference (Supplementary figure 4a).

C1ql2-Nrxn3(25b+) interaction is important for C1ql2 localization at the MFS and SV recruitment

a Experimental design to analyze the MFS after AAV-mediated expression of C1ql2.K262E in Bcl11b cKO DGN. b Western blot and c relative C1ql2.K262E protein levels in mouse hippocampal homogenates. n=3. All data are presented as means; error bars indicate SEM. Mann-Whitney U-test. ns: non-significant. d Immunohistochemistry of C1ql2 (red) and GFP (green) in hippocampal sections. Scale bar: 200 μm. Upper right corner of C1ql2 panels depicts close-ups from the SL of CA3. Scale bar: 15 μm. e Integrated density of C1ql2 fluorescence in the SL of CA3. n=3. All data are presented as means; error bars indicate SEM. Unpaired t-test. *p=0.008. f Electron microscope images of MFS and proximal SV. White bars mark synapse length from postsynaptic side. Scale bar: 100 nm. g Average synapse score. n=3. Unpaired t-test. **p=0.0012. h Number of docked vesicles per 100 nm AZ profile length. n=3. All data are presented as means; error bars indicate SEM. Points represent the individual examined AZ. Unpaired t-test. *p=0.018.

To investigate how the misdistribution of C1ql2.K262E in the SL affected the structure of MFS terminals, we determined if C1ql2.K262E was able to recover the average synapse score in the Bcl11b cKO background. As expected, we found an ineffective activity of the mutant molecule (Figure 6f-g, Supplementary figure 4b; Control+EGFP: 3.39±0.048, Bcl11b cKO+EGFP-2A-K262E: 2.87±0.043, mean±SEM), consistent with a role for C1ql2 in organizing MFS structure. Furthermore, the number of docked vesicles per 100 nm of AZ profile length in the MFB of animals receiving the C1ql2.K262E AAV was significantly lower compared to control animals (Figure 6h, Supplementary figure 4c-d; Control+EGFP: 0.41±0.069, Bcl11b cKO+EGFP-2A-K262E: 0.19±0.035, mean±SEM), indicating that C1ql2-dependent signaling also affected the docking of SV within these synapses. In contrast, the diameter of docked vesicles was unchanged (Supplementary figure 4e; Control+EGFP: 39.06±1.22, Bcl11b cKO+EGFP-2A-K262E: 36.38±2.19, mean±SEM). Unexpectedly, however, C1ql2.K262E was able to rescue the loss of MF-LTP observed in Bcl11b cKO (Supplementary figure 4f-h; 0-10 min: Control+EGFP: 90.4±7.2, Bcl11b cKO+EGFP-2A-K262E: 155.8±30.3, 10-20 min: Control+EGFP: 42.7±3.6, Bcl11b cKO+EGFP-2A-K262E: 68.7±17.3, 20-30 min: Control+EGFP: 52.5±7.6, Bcl11b cKO+EGFP-2A-K262E: 55.9±13.8, 30-40 min: Control+EGFP: 50.1±7.3, Bcl11b cKO+EGFP-2A-K262E: 47.8±9.3, mean±SEM). Thus, our data suggest that Bcl11b regulates MFS function through divergent C1ql2-dependent downstream signaling pathways: while SV recruitment depends on a direct interaction of C1ql2 with Nrxn3(25b+), C1ql2 appears to regulate MF-LTP through Nrxn3(25b+)-independent mechanisms.

To explore if the binding to Nrxn3α(25b+) is required for correct C1ql2 targeting to the SL in vivo, we stereotaxically injected an AAV expressing GFP-tagged Cre or inactive Cre into the DG of 2 month old Nrxn123flox/flox mice (Figure 7a), which resulted in strong reduction of Nrxn3 mRNA levels 2 months later, (Figure 7b; inactive Cre: 1±0.127, Cre: 0.38±0.09, mean±SEM). Strikingly, endogenous C1ql2 expression along the MF axons in the SL of CA3 was significantly reduced in Nrxn123 cKO animals compared to Nrxn123flox/flox animals expressing inactive Cre (Figure 7c-d; inactive Cre: 11.65±1.75, Cre: 4.71±0.93, mean±SEM). To control for the specificity of this effect, we also determined the level of C1ql3 expression and found it unchanged in Nrxn123 cKO (Supplementary figure 5a). To exclude that the reduced C1ql2 expression was simply a consequence of an overall loss of MFB, we used ZnT3 as a marker of MFB and found it unchanged in Nrxn123 cKO compared to controls (Figure 7e). Remarkably, disruption of the C1ql2-Nrxn3(25b+) binding by ablation of Nrxn3 in Nrxn123 cKO mutants not only led to the reduced C1ql2 expression (Figure 7c) but in fact recapitulated the phenotype observed upon Bcl11b ablation or by KD of C1ql2 as evidenced by a large reduction of the average synapse score in Nrxn123 cKO (Figure 7f-g, inactive Cre: 3.11±0.06; Cre: 2.67±0.074, mean±SEM). Also, similarly to the Bcl11b and C1ql2 mutant phenotypes, we observed the number of docked vesicles per 100 nm of AZ profile length in Nrxn123 cKO to be diminished compared to controls (Figure 7h, Supplementary figure 5c-d; inactive Cre: 0.404±0.035, Cre: 0.197±0.019, mean±SEM), whereas the diameter of docked vesicles remained unchanged (Supplementary figure 5e; inactive Cre: 38.38±0.44; Cre: 37.12±0.8, mean±SEM). Thus, our results provide compelling evidence that Bcl11b controls MFS organization through C1ql2/Nrxn3(25b+)-dependent signaling, explicating how Bcl11b, a transcription factor with a broad range of functions, can regulate highly specific processes in the brain.

Nrxn KO perturbs C1ql2 localization at the MFS and SV

a Experimental design to analyze the MFS after AAV-mediated Nrxn KO. b Relative Nrxn3 mRNA levels. n=4. All data are presented as means; error bars indicate SEM. Unpaired t-test. **p=0.007. c Immunohistochemistry of C1ql2 (red) and GFP (green) in hippocampal sections. Scale bar: 200 μm. Upper right corner of C1ql2 panels depicts close-ups from the SL of CA3. Scale bar: 15 μm. d Integrated density of C1ql2 fluorescence in the SL of CA3. n=3. All data are presented as means; error bars indicate SEM. Unpaired t-test. *p=0.02. e Immunohistochemistry of ZnT3 (red) and GFP (green) in hippocampal sections. Scale bar: 200 μm. Upper right corner of ZnT3 panels depicts close-ups from the SL of CA3. Scale bar: 15 μm. f Electron microscope images of MFS and proximal SVs. White bars mark synapse length from postsynaptic side. Scale bar: 100 nm. g Average synapse score. n=3. Unpaired t-test. **p=0.009. h Number of docked vesicles per 100 nm AZ profile length. n=3. All data are presented as means; error bars indicate SEM. Points represent the individual examined AZ. Unpaired t-test. **p=0.007.

Discussion

There is emerging evidence that the zinc finger transcription factor Bcl11b is involved in the pathogenesis of neurodevelopmental as well as neuropsychiatric disorders that are frequently associated with synaptic dysfunction. Previous work from our group demonstrated Bcl11b to be essential for synapse function in the mossy fiber circuit of the adult murine hippocampus. The underlying molecular mechanisms downstream of Bcl11b, however, remained elusive. In the present study, we uncover a novel C1ql2-dependent regulatory pathway through which Bcl11b controls the structural as well as functional integrity of hippocampal MFS in adult mice. We show that SV recruitment to the AZ of MFS, as well as the expression of MF-LTP, depend on C1ql2, which is a direct functional target of Bcl11b. Reintroduction of C1ql2 into Bcl11b mutant DGN restores defective SV recruitment and LTP expression. KD of C1ql2 in DGN recapitulates the impaired SV recruitment and loss of LTP observed in Bcl11b mutants. Finally, we show that C1ql2 controls SV recruitment through a direct interaction with presynaptic Nrxn3(25b+), while LTP potentially depends on a Nrxn3-independent pathway. Recent studies suggested Nrxn3, as well as C1ql2, to be associated with neuropsychiatric disorders. Thus, our study for the first time identifies a Bcl11b/C1ql2/Nrxn3-dependent signaling pathway in the control of basic structural and functional properties of MFS. Analysis of this regulatory pathway in mice may provide important novel insights into the pathogenesis of neurodevelopmental and neuropsychiatric disorders.

We have previously shown that conditional ablation of Bcl11b in the adult hippocampus leads to structural and functional changes of MFS characterized by an overall reduction in synapse numbers, loss of bouton complexity, misdistribution of SV as well as loss of MF-LTP (De Bruyckere et al., 2018). Here, we found that AAV-mediated reintroduction of the synaptic organizer protein C1ql2, which is a direct transcriptional target of Bcl11b and is downregulated in Bcl11b mutant DGN (De Bruyckere et al., 2018), was able to rescue major part of the Bcl11b mutant phenotype at the MFS. Restoring C1ql2 expression in Bcl11b cKO DGN led to a complete rescue of the SV distribution and docking as well as LTP at the MFS, while synapse numbers and ultrastructural complexity of boutons remained unchanged. Furthermore, KD of C1ql2 in WT DGN recapitulated the Bcl11b phenotype with impaired SV recruitment and loss of LTP, supporting the specificity of C1ql2 function. MF-LTP, which manifests as a long-term increase in presynaptic vesicle release probability (Pr) (Shahoha et al., 2022), directly depends on the distribution of SV at AZ. Recent studies have shown that the increase in Pr involves the recruitment of new AZ and an increase in the number of docked and tethered vesicles, corresponding to the readily releasable pool of SV (Orlando et al., 2021; Vandael et al., 2020). The perturbed SV recruitment in both Bcl11b cKO and C1ql2 KD mice could thus explain the loss of LTP in both conditions. Indeed, reintroduction of C1ql2 in Bcl11b cKO DGN specifically rescued SV recruitment and LTP, while synapse numbers and ultrastructural complexity of MFB remained unchanged. It is important to note, however, that additional factors have been suggested to contribute to the increase in Pr, including a tighter coupling between Ca2+ channels and SV (Midorikawa & Sakaba, 2017) and the accumulation of Ca2+ channels near release sites (Fukaya et al., 2021). Therefore, it cannot be excluded that C1ql2 regulates MF-LTP through one of these alternative mechanisms. Aiming to narrow in on the nature of the mechanism through which Bcl11b regulates MF-LTP, we used forskolin to induce LTP in Bcl11b cKO. MF-LTP relies on presynaptic mechanisms (Castillo, 2012; Zalutsky & Nicoll, 1990), with a central role for the second messenger cAMP. This mediator is produced by AC in response to Ca2+ influx through voltage-gated Ca2+ channels (Li et al., 2007) and kainate receptors (KAR) (Lauri et al., 2001; Schmitz et al., 2003). By using forskolin to directly activate AC we bypassed these initial steps, and still found a reduction of LTP in Bcl11b cKO mice, similarly to HFS. These results confirm that the loss of LTP is caused by a process downstream of the initial presynaptic Ca2+ influx following stimulation. We note that, in the present experiments, we did not observe the decrease in input-output relation as reported in De Bruyckere et al. (2018). After excluding technical differences, e.g. different methods of data analysis, we conclude that the discrepancy is best explained by differences in the population of presynaptic fibers. In the present study, mossy fiber responses were specifically identified by testing for frequency facilitation and sensitivity to mGluR antagonists, whereas in the previous study this purification was not done (De Bruyckere et al., 2018). Together, our results suggest Bcl11b to be an important synaptic regulator that controls the structure and function of adult MFS through both C1ql2-dependent, as well as -independent transcriptional programs.

C1ql proteins are complement-related factors that are synthesized by the presynapse and secreted into the synaptic cleft. Within the hippocampus, C1ql2 and −3 protein expression overlaps and is highly restricted to DGN (Iijima et al., 2010) and the corresponding mossy fiber system, including MF-CA3 synapses. C1ql2 and −3 were previously suggested to form functional heteromers that are able to cluster postsynaptic KAR on MFS. Selective deletion of either C1ql2 or −3 in mice was reported to have no overt mutant hippocampal phenotype, suggesting functional compensation for both proteins (Matsuda et al., 2016). Using shRNA-mediated selective KD of C1ql2 in DGN as well as rescue of the Bcl11b mutation by the reintroduction of C1ql2 into mutant DGN, we observed a novel presynaptic function for C1ql2 in the recruitment of SV and the expression of LTP in MFS. This function was specific to C1ql2 since overexpression of C1ql3 in Bcl11b mutant DGN was unable to rescue the synapse phenotype. Previous in vitro studies suggested that C1ql2 function at the MFS involves interaction with Nrxn3 isoforms containing the SS525b sequence (Matsuda et al., 2016). Nrxns are synaptic cell adhesion molecules that mediate various synaptic properties (Reissner et al., 2013; Sudhof, 2017), including the recruitment of SV and dense-core vesicles (Dean et al., 2003; Ferdos et al., 2021; Quinn et al., 2017; Rui et al., 2017). This prompted us to analyze, whether C1ql2-dependent SV recruitment in MFS requires a direct interaction with Nrxn3(25b+) in vitro and in vivo. Expression of C1ql2 in HEK293 cells co-cultured with GFP-Nrxn3α(25+)-expressing hippocampal neurons was able to recruit Nrxn3α(25b+) and vGlut1 at contact points, while C1ql2.K262E, a C1ql2 variant with an amino-acid replacement that perturbs the interaction with Nrxn3(25b+), was no longer able to recruit neuronal vGlut1. Furthermore, clustering of vGlut1 by C1ql2-expressing HEK293 cells was reduced in neurons harboring a pan-neurexin mutation, a phenotype that was rescued by the selective reintroduction of Nrxn3α(25b+). Finally, the introduction of C1ql2.K262E in Bcl11b cKO DGN in vivo was unable to rescue the SV recruitment, while the silencing of Nrxns in DGN in vivo perturbed the SV recruitment to a similar extent as in Bcl11b cKO and C1ql2 KD. Based on these findings, we anticipated the overexpression of C1ql2.K262E in Bcl11b cKO DGN to be unable to rescue MF-LTP. Unexpectedly, the introduction of C1ql2.K262E into Bcl11b cKO fully rescued MF-LTP. This raises the possibility that C1ql2 can influence MF-LTP through additional, yet uncharacterized mechanisms, independent of SV recruitment or direct interaction with Nrxn3(25b+). We cannot exclude, however, that the expression of a mutant C1ql2 variant created an additional gain-of-function effect that circumvented SV recruitment and allowed the rescue of MF-LTP in our experimental system. The latter is supported by the fact that within the first 10 minutes after HFS, fEPSP slopes for C1ql2.K262E were significantly elevated compared to controls, an effect which was not seen after C1ql2 re-expression. Together, our data provide comprehensive experimental evidence that the direct interaction of C1ql2 with Nrxn3(25b+) is essential for SV recruitment at the MFS. The C1q domain is able to form stable, higher-order oligomers (Ressl et al., 2015). Neurexins, on the other hand, are highly mobile outside and inside of synaptic terminals Error! Reference source not found.(Klatt et al., 2021; Neupert et al., 2015). We observed that an abolished interaction between C1ql2 and Nrxn3(25b+) was accompanied by impaired targeting of the C1ql2 protein to the MFS. Thus, the interaction of C1ql2 with Nrxn3(25b+) may reciprocally augment the accumulation of both proteins at synaptic sites.

Perturbations in synaptic structure and function are major determinants of various neuropsychiatric and neurodevelopmental disorders (Hayashi-Takagi, 2017; Lepeta et al., 2016; Zoghbi & Bear, 2012). Emerging evidence from recent genetic studies suggests such disorders to be linked to various genes encoding for synaptic proteins (Sudhof, 2021; Torres et al., 2017; Wang et al., 2018). Decoding the molecular mechanisms of synaptic organization and stability and their transcriptional regulation would therefore be expected to contribute to the mechanistic understanding of neuropsychiatric and neurodevelopmental disorders. The transcription factor Bcl11b has been linked to neurodevelopmental (Lessel et al., 2018), neurodegenerative (Kunkle et al., 2016; Song et al., 2022) and neuropsychiatric disorders (Whitton et al., 2018; Whitton et al., 2016). BCL11B mutations in humans are associated with neurodevelopmental delay, overall learning deficits as well as impaired speech acquisition and autistic features (Eto et al., 2022; Lessel et al., 2018; Punwani et al., 2016; Yang et al., 2020). Moreover, conditional ablation of Bcl11b selectively in the adult murine hippocampus results in impaired learning and memory behavior (Simon et al., 2016). NRXN3 single-nucleotide polymorphisms (SNP) have been implicated in schizophrenia (Hu et al., 2013) and addiction (Hishimoto et al., 2007), with one recorded SNP located close to SS5 altering the expression of Nrxn3(25b+). Interestingly, recent studies have also associated C1QL2 with schizophrenia (Marballi et al., 2022) as well as cocaine addiction (Huggett & Stallings, 2020a, 2020b). In this study we demonstrate that Bcl11b, through its transcriptional target C1ql2, modulates the synaptic organization of MFS by controlling the recruitment of SV at AZ. This regulatory mechanism depends on a direct interaction of C1ql2 with Nrxn3(25b+). Importantly, SV trafficking and altered release probability have been implicated in neurological and neuropsychiatric disorders (Egbujo et al., 2016; Lepeta et al., 2016; Zhu et al., 2021). Thus, the identification of the Bcl11b/C1ql2/Nrxn3(25b+)-dependent signaling module in this study provides a new entry point for future mechanistic analyses of synaptopathies. Moreover, the existence of such cell type-specific signaling modules reveals how a fundamental transcription factor with diverse functions such as Bcl11b can be implicated in the pathomegenesis of brain disorders characterized by synaptic dysfunction.

Materials and Methods

Animals

Bcl11b inducible mutants were generated as previously described (De Bruyckere et al., 2018). Bcl11bflox/flox; CaMKIIa-CreERT2 (Bcl11b cKO) and Bcl11b+/+; CaMKIIa-CreERT2 (control) littermates were used. The Bcl11b mutation was induced by intraperitoneal injection of 2 mg tamoxifen for five consecutive days. C57BL/6JRj mice were obtained from Janvier-Labs. For the pan-neurexin KO, Nrxn123flox mice (Chen et al., 2017) were used. Animals were kept in a 12:12-h light–dark cycle at a constant temperature (22 ± 1 °C) in IVC cages. All mouse experiments were carried out in compliance with the German law and approved by the respective government offices in Tübingen and Karlsruhe, Germany.

Stereotaxic Injections

For the expression of C1ql2 and C1ql3, the DG of 80 days old Bcl11b cKO mice were injected with AAV vectors expressing EGFP-2A-C1ql2 and EGFP-2A-C1ql3, respectively. As control, Bcl11b cKO and control mice were injected with an AAV expressing EGFP. For the KD of C1ql2, the DG of 60 days old C57BL/6JRj mice were injected with AAV 4xshC1ql2-EGFP, expressing 4 shRNAs against C1ql2 or control AAV 4xshNS-EGFP, expressing 4 non-sense shRNAs. For pan-neurexin KO, the DG of 60 days old Nrxn123flox/flox mice were injected with an AAV expressing EGFP-Cre or a control AAV expressing EGFP-Cre.Y324F, an inactive Cre. All AAV were produced by the Viral Vector Facility of the Neuroscience Center Zurich on request. The mice were anesthetized with 5% isoflurane and placed in a mouse stereotaxic apparatus. During the entire procedure, anesthesia was maintained by constant administration of 2.2% isoflurane. Eye ointment was applied to prevent eyes from drying. For electrophysiological experiments, mice were subcutaneously injected with buprenorphine hydrochloride (0.1 mg/kg, Temgesic, Indivior) 30 minutes before and 3 hours after each surgery. For all other experiments, Butorphanol (Livisto) and Meloxicam (Boehringer-Ingelheim; 5 μg/g) were injected subcutaneously and the local anesthetic Bupivacaine (Puren; 5 μg/g) was injected subcutaneously at the incision site. After 10 min the head of the mouse was shaved and disinfected and an incision was made in the skin. Targeted injection sites were identified and a small craniotomy was performed for each site. The injector was placed at the individual sites and the viral solution was injected at 100 nL/min, with a 5-10 min recovery before removing the injector. After injections at all sites the incision was sutured and the animal was monitored for recovery from anesthesia, after which it was returned to its home cage. For histological and EM analyses of MFS, AAV were injected at three sites per hemisphere with the following coordinates (Bregma: AP 0; ML: 0; DV:0): AP −2 mm; ML ±1 mm; DV −2 mm. AP −2.5 mm; ML ±1.5 mm; DV −1.8 mm. AP −3.1 mm; ML ±2, DV −2.2 mm. For electrophysiological analyses, AAV were injected at two dorsoventral coordinates per hemisphere: AP −3.0 mm; ML ±3.25 mm; DV −2.4 and −2.8 mm. 200-300 nL of AAV (1e12 vg/mL) were injected in each location.

RNA isolation and quantitative real-time PCR

All procedures were performed in an RNase-free environment. Animals were sacrificed under deep CO2-induced anesthesia, brains were quickly dissected in ice-cold PBS, cryopreserved in 20% sucrose overnight, frozen in OCT compound (Polysciences), and stored at −80 °C. 20 μm thick coronal sections were collected on UV-treated and 0.05% poly-L-lysine coated membrane-covered PEN slides (Zeiss), fixed for 1 min in ice-cold 70% EtOH, incubated for 45 sec in 1% cresyl violet acetate solution (Waldeck) and washed for 1 min each in 70% EtOH and 100% EtOH. Sections were briefly dried on a 37 °C warming plate and immediately processed. The granule cell layer of the DG was isolated by laser capture microdissection using a PALM MicroBeam Rel.4.2 (Zeiss). RNA was isolated from the collected tissue using RNeasy Micro Kit (Qiagen) and reverse transcribed using the SensiFast cDNA Synthesis Kit (Bioline). Quantitative real-time PCR was performed in triplets for each sample using the LightCycler DNA Master SYBR Green I Kit in a LightCycler 480 System (Roche). The relative copy number of Gapdh RNA was used for normalization. Data were analyzed using the comparative CT method (Schmittgen & Livak, 2008).

Western blots

Briefly, hippocampi from freshly removed brains were dissected in ice-cold PBS, collected in Lysis Buffer (50mM Tris pH 7.5, 150mM NaCl, 0.5% sodium deoxycholate, 1% triton-X100, 0.1% SDS) and manually homogenated. Samples were centrifuged for 25 min at 13200 rpm at 4 °C and the supernatant was collected. Protein concentration was calculated with Bradford assay. Protein suspension containing 40 μg of protein was mixed 1:1 with 2x SDS loading dye (62.5mM Tris, 10% Glycerol, 5% β-mercaptoethanol, 80mM SDS, 1.5mM bromophenol blue), boiled at 95 °C for 5 min, separated by SDS-PAGE and electrophoretically transferred onto PVDF membranes (Merck). Membranes were blocked with 5% non-fat milk (Sigma-Aldrich), incubated with mouse anti-β-actin (1:5000; Sigma-Aldrich) and rabbit anti-C1ql2 (1:500; Sigma-Aldrich), followed by Peroxidase-conjugated secondary antibodies (Jackson ImmunoResearch) and developed with Pierce ECL Western Blotting Substrate (ThermoScientific). Signal was detected with ChemiDoc Imaging System (BioRad) and analyzed with Image Lab Software (BioRad). Protein signal was normalized with the signal of β-actin.

Immunohistochemistry and RNA in situ Hybridization

Animals were sacrificed under deep CO2-induced anesthesia, brains were dissected in ice-cold 1x PBS, and either fixed for 4 h in 4% PFA in PBS at 4 °C and cryopreserved in 20% sucrose in PBS overnight at 4 °C or directly cryopreserved and then frozen in OCT compound (Polysciences). Sections were prepared at 14 μm. The unfixed sections were postfixed with 4% PFA in 1x PBS for 20 min. Heat-induced antigen retrieval in 10mM citrate buffer (pH 6.0) was performed for fixed sections. Sections were blocked at RT for 1 h in 1x PBS containing 0.1% TritonX-100 and 10% horse serum, and incubated overnight at 4 °C with primary antibodies, followed by a 90 min incubation with secondary antibodies. Sections were counterstained with DAPI (Molecular Probes). The following primary antibodies were used on fixed sections: guinea pig anti-Bcl11b (1:1000; Simon et al., 2012), rabbit anti-C1ql2 (1:1000; Invitrogen), rabbit anti-C1ql2 (1:500; Sigma-Aldrich), chicken anti-GFP (1:2000; Abcam) and rabbit anti-C1ql3 (1:500; Biozol). Primary antibodies used on unfixed sections: mouse anti-vGlut1 (1:100; Synaptic Systems), guinea pig anti-Homer1 (1:250; Synaptic Systems), rabbit anti-C1ql2 (1:1000; Invitrogen) and rabbit anti-ZnT3 (1:200; Synaptic Systems). All fluorescent secondary antibodies were purchased from Jackson ImmunoResearch and used at 1:500 dilution. Hybridizations were performed with DIG-labelled riboprobes on 14 μm thick sections.

Transmission electron microscopy

Animals were sacrificed through CO2-inhalation and immediately perfused transcardially with 0.9% NaCl for 1 min, followed by a fixative solution of 1.5% glutaraldehyde (Carl Roth) and 4% PFA in 0.1M PB pH 7.2 for 13 min. Brains were dissected and postfixed in the fixative solution for 4 h at 4 °C. Ultrathin sections (60 nm) were prepared and stained with lead citrate. Images were acquired using a transmission electron microscope LEO 906 (Zeiss) with a sharp-eye 2k CCD camera and processed with ImageSP (Tröndle). Synapse score (De Bruyckere et al., 2018) was calculated according to the following criteria: 0-5 vesicles above the active zone = 0; 5-20 vesicles = 1; small group of vesicles (≤ 200000 nm2) with distance between density and closest vesicle > 100 nm = 2; small group of vesicles (≤ 200000 nm2) with distance between density and closest vesicle ≤ 100 nm = 3; big group of vesicles (> 200000 nm2) with distance between density and closest vesicle (> 100 nm) = 4; big group of vesicles (> 200000 nm2) with distance between density and closest vesicle ≤ 100 nm = 5. Synapses from approximately 30 MFB per animal were analyzed. Vesicles with a distance ≤ 5 nm from the plasma membrane were considered docked (Kusick et al., 2022; Vandael et al., 2020). Approximately 100 AZ per animal were analyzed.

Electrophysiological recordings and data analysis

Animals were sacrificed under deep CO2-induced anesthesia at 4 months. Brains were quickly removed and placed in ice-cold modified ACSF containing (in mM) 92 N-methyl-D-glucamine (NMDG), 2.5 KCl, 1.2 NaH2PO4, 30 NaHCO3, 20 HEPES, 25 glucose, 5 Na-ascorbate, 2 thiourea, 3 Na-pyruvate, 10 MgSO4, 0.5 CaCl2, 6 N-acetyl-L-cysteine (NAC), saturated with carbogen gas (95% O2 and 5% CO2, pH 7.4) (Ting et al., 2014). 450 µm thick horizontal slices were cut using a vibratome slicer (Leica) at a defined angle to improve the preservation of mossy fibers (Bischofberger et al., 2006). After cutting, slices were transferred to a “Haas”-type interface chamber (Haas et al., 1979), where they were perfused with carbogen-saturated ACSF containing (in mM) 124 NaCl, 3 KCl, 2.3 CaCl2, 1.8 MgSO4, 10 glucose, 1.25 NaH2PO4, 26 NaHCO3 (pH 7.4 at 34°C) at a rate of 1.5 mL/min at 34 ± 1 °C. Slices were allowed to recover for a minimum of 1 h before the start of recordings.

Recordings were carried out by placing a glass micropipette (tip diameter 3–5 μm) filled with ACSF in the SL of the CA3b area. To induce MF field excitatory post-synaptic potentials (fEPSP), a bipolar electrode (Science Products) was placed within the hilus region of the DG. 0.1 ms pulses were delivered with an Iso-Flex stimulus isolator (AMPI) at 20 sec intervals. Putative mossy fiber signals were preliminarily identified using a 25 Hz train of five pulses. Input-output relationships were obtained by measuring the fiber volley amplitude and fEPSP slope in response to stimulations with intensities ranging from 3 to 40 V. For LTP recordings, stimulation intensity for each slice was adjusted to obtain a slope value of 20% (30% in the case of forskolin (Biomol) experiments) of the maximum fEPSP slope. LTP was induced by three trains of 100 stimulation pulses at 100 Hz (high frequency stimulation, HFS), repeated every 8 sec. 3 μM DCG-IV (Tocris Bioscience) was applied after each experiment, and only recordings displaying >70% reduction in putative MF-fEPSP slopes were used for analysis. fEPSPs were amplified 100x with an EXT 10-2F amplifier (npi electronics). Signals were low-pass filtered at 2 kHz and high-pass filtered at 0.3 Hz, digitized at 20 kHz with an analog-to-digital converter (Cambridge Electronic Design (CED)) and stored for offline analysis using Spike2 (v7) software (CED). Offline data analysis was performed on raw traces using Spike2. Slope values were measured from the linear part of the fEPSP rising phase by manually placing vertical cursors. Changes in fEPSP slopes were calculated as a percentage of the average baseline fEPSP ((average fEPSP slope in a given time interval after HFS - average fEPSP slope before HFS)/ (average fEPSP slope before HFS)).

DNA constructs

For expression, C1ql2 was cloned from mouse cDNA. A 6xHis-myc tag was attached to the N-terminus and the construct was cloned into the pSecTag2A vector (Invitrogen) in frame with the N-terminal IgK signal sequence. A stop codon was introduced directly after C1ql2. The K262E point mutation was introduced with the Q5 Site-Directed Mutagenesis Kit (New England Biolabs). pSecTag2A was used for control experiments. Rat Nrxn3α(25b+) cDNA (Ushkaryov & Sudhof, 1993) was inserted into an pSyn5 vector with human Synapsin promoter (Neupert et al., 2015) using BamHI and BglII. For the pan-neurexin KO and the control experiments, vectors with NLS-GFP-Cre or NLS-GFP-Cre.Y324F were used (Klatt et al., 2021; Wang et al., 2016). All vectors were validated by sequencing (Eurofins Genomics).

Primary Hippocampal Cultures

Hippocampi were dissected from P0 mice in HBSS media, digested for 15 min with HBSS containing 0.1% Trypsin (Gibco) at 37 °C, dissociated in plating media (MEM supplemented with 0.6% glucose, 10% FBS, 1% penicillin/streptomycin, DNaseI 4 U/mL) and seeded on poly-L-Lysin precoated coverslips placed inside 12-well plates at 1.5×105 cells/mL. After 3 h the plating media was replaced with neuronal growth media (Neurobasal A supplemented with 2% B27, 2mM L-Glutamine, 1% penicillin/streptomycin, 1% N2 and 0.005% NGF). Cultures were kept at 37 °C under 5% CO2 atmosphere. The day of plating was considered as 0 days in vitro (DIV). At DIV3 and DIV7 80% of the medium was exchanged with fresh growth medium. At DIV9 the medium was exchanged with penicillin/streptomycin-free growth medium and at DIV10 neurons were transfected using Lipofectamine 2000 (Invitrogen). Briefly, a total of 200 μL transfection mix per well was prepared by first mixing 100 μL Opti-MEM with 4 μL Lipofectamine 2000 in one tube and 100 μL Opti-MEM with 3 μg DNA in a different tube. After 5 min both volumes were combined and the mixture was incubated for 20 min at RT. The transfection mix was then added dropwise to the neurons. After 3 h of incubation the medium was exchanged with fresh growth medium.

HEK293 Cell Culture

Human embryonic Kidney (HEK) 293 cells were obtained from ATCC and were maintained in DMEM supplemented with 10% fetal calf serum and 1% penicillin/streptomycin at 37 °C under 5% CO2 atmosphere. Cells were transfected using Lipofectamine 2000 according to the manufacturer’s instructions on the same day the neurons were transfected. Cells were incubated for at least 24 h before being used in co-culture experiments.

Neuronal and HEK293 co-culture and immunostaining

Transfected HEK293 cells were washed, dissociated and resuspended in neuronal growth medium. 15×103 cells were added in each well containing DIV11 transfected neurons. HEK293 cells were co-cultured with the hippocampal primary neurons 2 days (DIV13 for neurons) before proceeding with immunostaining. Coverslips with cultured neurons and HEK293 cells were first fixed with 4% PFA in 1x PBS for 10 min at 4 °C, then washed 3x with 1 mL 1x PBS and blocked with 1x PBS containing 0.1% Triton X-100 and 10% horse serum for 1 h at RT. Primary antibodies were incubated overnight at 4 °C, followed by a 90 min incubation with secondary antibodies. Cells were counterstained with DAPI (Molecular Probes). The following primary antibodies were used: rabbit anti-myc-tag (1:2000; Abcam), guinea pig anti-vGlut1 (1:250; Synaptic Systems), chicken anti-GFP (1:2000; Abcam). All fluorescent secondary antibodies were purchased from Jackson ImmunoResearch and used at 1:500 dilution. For each condition, 25 cells per experiment were analyzed.

Structural protein modelling

The crystal structure of trimeric C1q-domains of mouse C1ql2 (Ressl et al., 2015) was used to predict a potential electrostatic binding site to splice insert 25 of Nrxn3α. An electrostatic surface map of the trimer was calculated using APBS (Jurrus et al., 2018). The K262E mutation was introduced using FoldX (foldxsuite.crg.eu) and was chosen in order to generate a negatively charged surface that would potentially be repulsive to Nrxn3α binding. Final models were visualized with PyMOL (pymol.org).

Image acquisition and analysis

All fluorescent images of sectioned hippocampal tissue were examined on a TCS SP5II confocal microscope (Leica) using LAS-X software and processed with Fiji (Schindelin et al., 2012). Overview images were acquired with a 20x objective. Synapse numbers were quantified in the SL imaged with a 40x objective at x2 zoom. C1ql2 fluorescence intensity was quantified in the SL imaged with a 40x objective. Acquisition settings were kept constant for every sample and condition. All fluorescent images of co-cultured HEK293 cells were examined on a TCS SP8 confocal microscope (Leica) using LAS-X software and processed with Fiji. Images were acquired with a 40x objective at 4x zoom. As before, acquisition settings were kept constant for every sample and condition. Images were analyzed by masking transfected HEK293 cells and measuring the area of each mask covered by the chosen stain.

Quantification and statistical analysis

Data analysis was conducted blind to the experimental treatment. Statistical analysis and graph generation was done using Python 3. If samples met the criteria for normality, we used two-tailed unpaired t-test to compare two groups and one-way ANOVA for more than two groups. For non-normally distributed data Mann-Whitney u-test was used. Two-way ANOVA was used for examining the influence of two different categorical independent variables. If ANOVAs were significant, we used a post hoc Tukey’s multiple-comparisons test to compare groups (structural and expression data) or a post hoc Bonferonni’s comparison test (electrophysiological data). Data are presented as mean±SEM. Significance levels were set as indicated in figures: *p<0.05, **p<0.01, ***p<0.001.

Data availability

All relevant data are available from the authors.

eLife Assessment

The authors identify a new role for C1ql2 at moss fiber synapses in the hippocampus and find that C1ql2, whose expression is controlled by Bcl11b, controls the recruitment of synaptic vesicles to active zones and is necessary for synaptic plasticity. The data implicating C1ql2 involvement, using numerous viral/genetic rescue approaches, are largely convincing, while the experimental evidence supporting the role of the specific Nrxn3 splice variant is less complete. These data are valuable, building on prior discoveries of how Bcl11b, a disease-relevant molecule, contributes to our understanding of mossy-fiber synaptic development.

Reviewer #1 (Public Review):

Koumoundourou et al., identify a pathway downstream of Bcl11b that controls synapse morphology and plasticity of hippocampal mossy fiber synapses. Using an elegant combination of in vivo, ex vivo, and in vitro approaches, the authors build on their previous work that indicated C1ql2 as a functional target of Bcl11b (De Bruyckere et al., 2018). Here, they examine the functional implications of C1ql2 at MF synapses in Bcl11b cKO mice and following C1ql2 shRNA. The authors find that Bcl11b KO and shRNA against C1ql2 significantly reduces the recruitment of synaptic vesicles and impairs LTP at MF synapses. Importantly, the authors test a role for the previously identified C1ql2 binding partner, exon 25b-containing Nrxn3 (Matsuda et al., 2016), as relevant at MF synapses to maintain synaptic vesicle recruitment. To test this, the authors developed a K262E C1ql2 mutant that disrupts binding to Nrxn3. Curiously, while Bcl11b KO and C1ql2 KD largely phenocopy (reduced vesicle recruitment and impaired LTP), only vesicle recruitment is dependent on C1ql2-Nrxn3 interactions. These findings provide new insight into the functional role of C1ql2 at MF synapses. While the authors convincingly demonstrate a role for C1ql2-Nrxn3(25b+) interaction for vesicle recruitment and a Nrxn3(25b+)-independent role for C1ql2 in LTP, the underlying mechanisms remain inconclusive. Additionally, a discussion of how these findings relate to previous work on C1ql2 at mossy fiber synapses and how the findings contribute to the biology of Nrxn3 would increase the interpretability of this work.

Reviewer #2 (Public Review):

This manuscript describes experiments that further investigate the actions of the transcription factor Bcl11b in regulating mossy fiber (MF) synapses in the hippocampus. Prior work from the same group had demonstrated that loss of Bcl11b results in loss of MF synapses as well as a decrease in LTP. Here the authors focus on a target of Bcl11b a secreted synaptic organizer C1ql2 which is almost completely lost in Bcl11b KO. Viral reintroduction of C1ql2 rescues the synaptic phenotypes, whereas direct KD of C1ql2 recapitulates the Bcl1 phenotype. C1ql2 itself interacts directly with Nrxn3 and replacement with a binding deficient mutant C1q was not able to rescue the Bcl11b KO phenotype. Overall there are some interesting observations in the study, however there are also some concerns about the measures and interpretation of data.

The authors state that they used a differential transcriptomic analysis to screen for candidate targets of Bcl11b, yet they do not present any details of this screen. This should be included and at the very least a table of all DE genes included. It is likely that many other genes are also regulated by Bcl11b so it would be important to the reader to see the rationale for focusing attention on C1ql2 in this study.

All viral-mediated expression uses AAVs which are known to ablate neurogenesis in the DG (Johnston DOI: 10.7554/eLife.59291) through the ITR regions and leads to hyperexcitability of the dentate. While it is not clear how this would impact the measurements the authors make in MF-CA3 synapses, this should be acknowledged as a potential caveat in this study.

The authors claim that the viral re-introduction "restored C1ql2 protein expression to control levels. This is misleading given that the mean of the data is 2.5x the control (Figure 1d and also see Figure 6c). The low n and large variance are a problem for these data. Moreover, they are marked ns but the authors should report p values for these. At the least, this likely large overexpression and variability should be acknowledged. In addition, the use of clipped bands on Western blots should be avoided. Please show the complete protein gel in primary figures of supplemental information.

Measurement of EM micrographs: As prior work suggested that MF synapse structure is disrupted the authors should report active zone length as this may itself affect "synapse score" defined by the number of vesicles docked. More concerning is that the example KO micrographs seem to have lost all the densely clustered synaptic vesicles that are away from the AZ in normal MF synapses e.g. compare control and KO terminals in Fig 2a or 6f or 7f. These terminals look aberrant and suggest that the important measure is not what is docked but what is present in the terminal cytoplasm that normally makes up the reserve pool. This needs to be addressed with further analysis and modifications to the manuscript.

The study also presents correlated changes in MF LTP in Bcl11b KO which are rescued by C1ql2 expression. It is not clear whether the structural and functional deficits are causally linked and this should be made clearer in the manuscript. It is also not apparent why this functional measure was chosen as it is unlikely that C1ql2 plays a direct role in presynaptic plasticity mechanisms that are through a cAMP/ PKA pathway and likely disrupted LTP is due to dysfunctional synapses rather than a specific LTP effect. The authors should consider measures that might support the role of Bcl11b targets in SV recruitment during the depletion of synapses or measurements of the readily releasable pool size that would complement their findings in structural studies.

Bcl11b KO reduces the number of synapses, yet the I-O curve reported in Supp Fig 2 is not changed. How is that possible? This should be explained.

Matsuda et al DOI: 10.1016/j.neuron.2016.04.001 previously reported that C1ql2 organizes MF synapses by aligning postsynaptic kainate receptors with presynaptic elements. As this may have consequences for the functional properties of MF synapses including their plasticity, the authors should report whether they see deficient postsynaptic glutamate receptor signaling in the Bcl11b KO and rescue in the C1ql2 re-expression.

Reviewer #3 (Public Review):

Overall, this is a strong manuscript that uses multiple current techniques to provide specific mechanistic insight into prior discoveries of the contributions of the Bcl11b transcription factor to mossy fiber synapses of dentate gyrus granule cells. The authors employ an adult deletion of Bcl11b via Tamoxifen-inducible Cre and use immunohistochemical, electron microscopy, and electrophysiological studies of synaptic plasticity, together with viral rescue of C1ql2, a direct transcriptional target of Bcl11b or Nrxn3, to construct a molecular cascade downstream of Bcl11b for DG mossy fiber synapse development. They find that C1ql2 re-expression in Bcl11b cKOs can rescue the synaptic vesicle docking phenotype and the impairments in MF-LTP of these mutants. They also show that C1ql2 knockdown in DG neurons can phenocopy the vesicle docking and plasticity phenotypes of the Bcl11b cKO. They also use artificial synapse formation assays to suggest that C1ql2 functions together with a specific Nrxn3 splice isoform in mediating MF axon development, extending these data with a C1ql2-K262E mutant that purports to specifically disrupt interactions with Nrxn3. All of the molecules involved in this cascade are disease-associated and this study provides an excellent blueprint for uncovering downstream mediators of transcription factor disruption. Together this makes this work of great interest to the field. Strengths are the sophisticated use of viral replacement and multi-level phenotypic analysis while weaknesses include the linkage of C1ql2 with a specific Nrxn3 splice variant in mediating these effects.

Here is an appraisal of the main claims and conclusions:

1. C1ql2 is a downstream target of Bcl11b which mediates the synaptic vesicle recruitment and synaptic plasticity phenotypes seen in these cKOs. This is supported by the clear rescue phenotypes of synapse anatomy (Fig.2) and MF synaptic plasticity (Fig.3). One weakness here is the absence of a control assessing over-expression phenotypes of C1ql2. It's clear from Fig.1D that viral rescue is often greater than WT expression (totally expected). In the case where you are trying to suppress a LoF phenotype, it is important to make sure that enhanced expression of C1ql2 in a WT background does not cause your rescue phenotype. A strong overexpression phenotype in WT would weaken the claim that C1ql2 is the main mediator of the Bcl11b phenotype for MF synapse phenotypes.

2. Knockdown of C1ql2 via 4 shRNAs is sufficient to produce the synaptic vesicle recruitment and MF-LTP phenotypes. This is supported by clear effects in the shRNA-C1ql2 groups as compared to nonsense-EGFP controls. One concern (particularly given the use of 4 distinct shRNAs) is the potential for off-target effects, which is best controlled for by a rescue experiment with RNA-insensitive C1ql2 cDNA as opposed to nonsense sequences, which may not elicit the same off-target effects.

3. C1ql2 interacts with Nrxn3(25b+) to facilitate MF terminal SV clustering. This claim is theoretically supported by the HEK cell artificial synapse formation assay (Fig.5), the inability of the K262-C1ql2 mutation to rescue the Bcl11b phenotype (Fig.6), and the altered localization of C1ql2 in the Nrxn1-3 deletion mice (Fig.7). Each of these lines of experimental evidence has caveats that should be acknowledged and addressed. Given the hypothesis that C1ql2 and Nrxn3b(25b) are expressed in DG neurons and work together, the heterologous co-culture experiment seems strange. Up till now, the authors are looking at pre-synaptic function of C1ql2 since they are re-expressing it in DGNs. The phenotypes they are seeing are also pre-synaptic and/or consistent with pre-synaptic dysfunction. In Fig.5, they are testing whether C1ql2 can induce pre-synaptic differentiation in trans, i.e. theoretically being released from the 293 cells "post-synaptically". But the post-synaptic ligands (Nlgn1 and and GluKs) are not present in the 293 cells, so a heterologous synapse assay doesn't really make sense here. The effect that the authors are seeing likely reflects the fact that C1ql2 and Nrxn3 do bind to each other, so C1ql2 is acting as an artificial post-synaptic ligand, in that it can cluster Nrxn3 which in turn clusters synaptic vesicles. But this does not test the model that the authors propose (i.e. C1ql2 and Nrxn3 are both expressed in MF terminals). Perhaps a heterologous assay where GluK2 is put into HEK cells and the C1ql2 and Nrxn3 are simultaneously or individually manipulated in DG neurons?

4. K262-C1ql2 mutation blocks the normal rescue through a Nrxn3(25b) mechanism (Fig.6). The strength of this experiment rests upon the specificity of this mutation for disrupting Nrxn3b binding (presynaptic) as opposed to any of the known postsynaptic C1ql2 ligands such as GluK2. While this is not relevant for interpreting the heterologous assay (Fig.5), it is relevant for the in vivo phenotypes in Fig.6. Similar approaches as employed in this paper can test whether binding to other known postsynaptic targets is altered by this point mutation.

5. Altered localization of C1ql2 in Nrxn1-3 cKOs. These data are presented to suggest that Nrx3(25b) is important for localizing C1ql2 to the SL of CA3. Weaknesses of this data include both the lack of Nrxn specificity in the triple a/b KOs as well as the profound effects of Nrxn LoF on the total levels of C1ql2 protein. Some measure that isn't biased by this large difference in C1ql2 levels should be attempted (something like in Fig.1F).

Author Response

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Koumoundourou et al., identify a pathway downstream of Bcl11b that controls synapse morphology and plasticity of hippocampal mossy fiber synapses. Using an elegant combination of in vivo, ex vivo, and in vitro approaches, the authors build on their previous work that indicated C1ql2 as a functional target of Bcl11b (De Bruyckere et al., 2018). Here, they examine the functional implications of C1ql2 at MF synapses in Bcl11b cKO mice and following C1ql2 shRNA. The authors find that Bcl11b KO and shRNA against C1ql2 significantly reduces the recruitment of synaptic vesicles and impairs LTP at MF synapses. Importantly, the authors test a role for the previously identified C1ql2 binding partner, exon 25b-containing Nrxn3 (Matsuda et al., 2016), as relevant at MF synapses to maintain synaptic vesicle recruitment. To test this, the authors developed a K262E C1ql2 mutant that disrupts binding to Nrxn3. Curiously, while Bcl11b KO and C1ql2 KD largely phenocopy (reduced vesicle recruitment and impaired LTP), only vesicle recruitment is dependent on C1ql2-Nrxn3 interactions. These findings provide new insight into the functional role of C1ql2 at MF synapses. While the authors convincingly demonstrate a role for C1ql2-Nrxn3(25b+) interaction for vesicle recruitment and a Nrxn3(25b+)independent role for C1ql2 in LTP, the underlying mechanisms remain inconclusive. Additionally, a discussion of how these findings relate to previous work on C1ql2 at mossy fiber synapses and how the findings contribute to the biology of Nrxn3 would increase the interpretability of this work.

As suggested by reviewer #1, we extended our discussion of previous work on C1ql2 and additionally discussed the biology of Nrxn3 and how our work relates to it. Moreover, we extended our mechanistic analysis of how Bcl11b/C1ql2/Nrxn3 pathway controls synaptic vesicle recruitment as well as LTP (please see also response to reviewer #2 points 5 and 8 and reviewer #3 point 4 of public reviews below for detailed discussion).

Reviewer #2 (Public Review):

This manuscript describes experiments that further investigate the actions of the transcription factor Bcl11b in regulating mossy fiber (MF) synapses in the hippocampus. Prior work from the same group had demonstrated that loss of Bcl11b results in loss of MF synapses as well as a decrease in LTP. Here the authors focus on a target of Bcl11b a secreted synaptic organizer C1ql2 which is almost completely lost in Bcl11b KO. Viral reintroduction of C1ql2 rescues the synaptic phenotypes, whereas direct KD of C1ql2 recapitulates the Bcl1 phenotype. C1ql2 itself interacts directly with Nrxn3 and replacement with a binding deficient mutant C1q was not able to rescue the Bcl11b KO phenotype. Overall there are some interesting observations in the study, however there are also some concerns about the measures and interpretation of data.

The authors state that they used a differential transcriptomic analysis to screen for candidate targets of Bcl11b, yet they do not present any details of this screen. This should be included and at the very least a table of all DE genes included. It is likely that many other genes are also regulated by Bcl11b so it would be important to the reader to see the rationale for focusing attention on C1ql2 in this study.

The transcriptome analysis mentioned in our manuscript was published in detail in our previous study (De Bruyckere et al., 2018), including chromatin-immunoprecipitation that revealed C1ql2 as a direct transcriptional target of Bcl11b. Upon revision of the manuscript, we made sure that this was clearly stated within the main text module to avoid future confusion. In the same publication (De Bruyckere et al., 2018), we discuss in detail several identified candidate genes such as Sema5b, Ptgs2, Pdyn and Penk as putative effectors of Bcl11b in the structural and functional integrity of MFS. C1ql2 has been previously demonstrated to be almost exclusively expressed in DG neurons and localized to the MFS.

There it bridges the pre- and post-synaptic sides through interaction with Nrxn3 and KAR subunits, respectively, and regulates synaptic function (Matsuda et al., 2016). Taken together, C1ql2 was a very good candidate to study as a potential effector downstream of Bcl11b in the maintenance of MFS structure and function. However, as our data reveal, not all Bcl11b mutant phenotypes were rescued by C1ql2 (see supplementary figures 2d-f of revised manuscript). We expect additional candidate genes, identified in our transcriptomic screen, to act downstream of Bcl11b in the control of MFS.

All viral-mediated expression uses AAVs which are known to ablate neurogenesis in the DG (Johnston DOI: 10.7554/eLife.59291) through the ITR regions and leads to hyperexcitability of the dentate. While it is not clear how this would impact the measurements the authors make in MF-CA3 synapses, this should be acknowledged as a potential caveat in this study.

We agree with reviewer #2 and are aware that it has been demonstrated that AAV-mediated gene expression ablates neurogenesis in the DG. To avoid potential interference of the AAVs with the interpretability of our phenotypes, we made sure during the design of the study that all of our control groups were treated in the same way as our groups of interest, and were, thus, injected with control AAVs. Moreover, the observed phenotypes were first described in Bcl11b mutants that were not injected with AVVs (De Bruyckere et al., 2018). Finally, we thoroughly examined the individual components of the proposed mechanism (rescue of C1ql2 expression, over-expression of C1ql3 and introduction of mutant C1ql2 in Bcl11b cKOs, KD of C1ql2 in WT mice, and Nrxn123 cKO) and reached similar conclusions. Together, this strongly supports that the observed phenotypes occur as a result of the physiological function of the proteins involved in the described mechanism and not due to interference of the AAVs with these biological processes. We have now addressed this point in the main text module of the revised ms.

The authors claim that the viral re-introduction "restored C1ql2 protein expression to control levels. This is misleading given that the mean of the data is 2.5x the control (Figure 1d and also see Figure 6c). The low n and large variance are a problem for these data. Moreover, they are marked ns but the authors should report p values for these. At the least, this likely large overexpression and variability should be acknowledged. In addition, the use of clipped bands on Western blots should be avoided. Please show the complete protein gel in primary figures of supplemental information.

We agree with reviewer #2 that C1ql2 expression after its re-introduction in Bcl11b cKO mice was higher compared to controls and that this should be taken into consideration for proper interpretation of the data. To address this, based also on the suggestion of reviewer #3 point 1 below, we overexpressed C1ql2 in DG neurons of control animals. We found no changes in synaptic vesicle organization upon C1ql2 over-expression compared to controls. This further supports that the observed effect upon rescue of C1ql2 expression in Bcl11b cKOs is due to the physiological function of C1ql2 and not as result of the overexpression. These data are included in supplementary figure 2g-j and are described in detail in the results part of the revised manuscript.

Additionally, we looked at the effects of C1ql2 overexpression in Bcl11b cKO DGN on basal synaptic transmission. We plotted fEPSP slopes versus fiber volley amplitudes, measured in slices from rescue animals, as we had previously done for the control and Bcl11b cKO (Author response image 1a). Although regression analysis revealed a trend towards steeper slopes in the rescue mice (Author response image 1a and b), the observation did not prove to be statistically significant, indicating that C1ql2 overexpression in Bcl11b cKO animals does not strongly alter basal synaptic transmission at MFS. Overall, our previous and new findings support that the observed effects of the C1ql2 rescue are not caused by the artificially elevated levels of C1ql2, as compared to controls, but are rather a result of the physiological function of C1ql2.

Following the suggestion of reviewer #2 all western blot clipped bands were exchanged for images of the full blot. This includes figures 1c, 4c, 6b and supplementary figure 2g of the revised manuscript. P-value for Figure 1d has now been included.

C1ql2 reintroduction in Bcl11b cKO DGN does not significantly alter basal synaptic transmission at mossy fiber-CA3 synapses.

a Input-output curves generated by plotting fEPSP slope against fiber volley amplitude at increasing stimulation intensities. b Quantification of regression line slopes for input-output curves for all three conditions. Control+EGFP, 35 slices from 16 mice; Bcl11b cKO+EGFP, 32 slices from 14 mice; Bcl11b cKO+EGFP-2A-C1ql2, 22 slices from 11 mice. The data are presented as means, error bars represent SEM. Kruskal-Wallis test (non-parametric ANOVA) followed by Dunn’s post hoc pairwise comparisons. p=0.106; ns, not significant.

Measurement of EM micrographs: As prior work suggested that MF synapse structure is disrupted the authors should report active zone length as this may itself affect "synapse score" defined by the number of vesicles docked. More concerning is that the example KO micrographs seem to have lost all the densely clustered synaptic vesicles that are away from the AZ in normal MF synapses e.g. compare control and KO terminals in Fig 2a or 6f or 7f. These terminals look aberrant and suggest that the important measure is not what is docked but what is present in the terminal cytoplasm that normally makes up the reserve pool. This needs to be addressed with further analysis and modifications to the manuscript.

As requested by reviewer #2 we analyzed and reported in the revised manuscript the active zone length. We found that the active zone length remained unchanged in all conditions (control/Bcl11b cKO/C1ql2 rescue, WT/C1ql2 KD, control/K262E and control/Nrxn123 cKO), strengthening our results that the described Bcl11b/C1ql2/Nrxn3 mechanism is involved in the recruitment of synaptic vesicles. These data have been included in supplementary figures 2c, 4h, 5f and 6g and are described in the results part of the revised manuscript.

We want to clarify that the synapse score is not defined by the number of docked vesicles to the plasma membrane. The synapse score, which is described in great detail in our materials and methods part and has been previously published (De Bruyckere et al., 2018), rates MFS based on the number of synaptic vesicles and their distance from the active zone and was designed according to previously described properties of the vesicle pools at the MFS. The EM micrographs refer to the general misdistribution of SV in the proximity of MFS. Upon revision of the manuscript, we made sure that this was clearly stated in the main text module to avoid further confusion.

The study also presents correlated changes in MF LTP in Bcl11b KO which are rescued by C1ql2 expression. It is not clear whether the structural and functional deficits are causally linked and this should be made clearer in the manuscript. It is also not apparent why this functional measure was chosen as it is unlikely that C1ql2 plays a direct role in presynaptic plasticity mechanisms that are through a cAMP/ PKA pathway and likely disrupted LTP is due to dysfunctional synapses rather than a specific LTP effect.

The inclusion of functional experiments in this and our previous study (de Bruyckere et al., 2018) was first and foremost intended to determine whether the structural alterations observed at MFB disrupt MFS signaling. From the signaling properties we tested, basal synaptic transmission (this study) and short-term potentiation (de Bruyckere et al., 2018) were unaltered by Bcl11b KO, whereas MF LTP was found to be abolished (de Bruyckere et al., 2018). Indeed, because MF LTP largely depends on presynaptic mechanisms, including the redistribution of the readily releasable pool and recruitment of new active zones (Orlando et al., 2021; Vandael et al., 2020), it appears to be particularly sensitive to the specific structural changes we observed. We therefore believe that it is valuable information that MF LTP is affected in Bcl11b cKO animals - it conveys a direct proof for the functional importance of the observed morphological alterations, while basic transmission remains largely normal. Furthermore, it subsequently provided a functional marker for testing whether the reintroduction of C1ql2 in Bcl11b cKO animals or the KD of C1ql2 in WT animals can functionally recapitulate the control or the Bcl11b KO phenotype, respectively.

We fully agree with the reviewer that C1ql2 is unlikely to directly participate in the cAMP/PKA pathway and that the ablation of C1ql2 likely disrupts MF LTP through an alternative mode of action. Our original wording in the paragraph describing the results of the forskolin-induced LTP experiment might have overstressed the importance of the cAMP pathway. We have now rephrased that paragraph to better describe the main idea behind the forskolin experiment, namely to circumvent the initial Ca2+ influx in order to test whether deficient presynaptic Ca2+ channel/KAR signaling might be responsible for the loss of LTP in Bcl11b cKO. The results are strongly indicative of a downstream mechanism and further investigation is needed to determine the specific mechanisms by which C1ql2 regulates MFLTP, especially in light of the result that C1ql2.K262E rescued LTP, while it was unable to rescue the SV recruitment at the MF presynapse. This raises the possibility that C1ql2 can influence MF-LTP through additional, yet uncharacterized mechanisms, independent of SV recruitment. As such, a causal link between the structural and functional deficits remains tentative and we have now emphasized that point by adding a respective sentence to the discussion of our revised manuscript. Nevertheless, we again want to stress that the main rationale behind the LTP experiments was to assess the functional significance of structural changes at MFS and not to elucidate the mechanisms by which MF LTP is established.

The authors should consider measures that might support the role of Bcl11b targets in SV recruitment during the depletion of synapses or measurements of the readily releasable pool size that would complement their findings in structural studies.

We fully agree that functional measurements of the readily releasable pool (RRP) size would be a valuable addition to the reported redistribution of SV in structural studies. We have, in fact, attempted to use high-frequency stimulus trains in both field and single-cell recordings (details on single-cell experiments are described in the response to point 8) to evaluate potential differences in RRP size between the control and Bcl11b KO (Figure for reviewers 2a and b). Under both recording conditions we see a trend towards lower values of the intersection between a regression line of late responses and the y-axis. This could be taken as an indication of slightly smaller RRP size in Bcl11b mutant animals compared to controls. However, due to several technical reasons we are extremely cautious about drawing such far-reaching conclusions based on these data. At most, they suffice to conclude that the availability of release-ready vesicles in the KO is likely not dramatically smaller than in the control.

The primary issue with using high-frequency stimulus trains for RRP measurements at MFS is the particularly low initial release probability (Pr) at these synapses. This means that a large number of stimulations is required to deplete the RRP. As the RRP is constantly replenished, it remains unclear when steady state responses are reached (reviewed by Kaeser and Regehr, 2017). This is clearly visible in our single-cell recordings (Author response image 2b), which were additionally complicated by prominent asynchronous release at later stages of the stimulus train and by a large variability in the shapes of cumulative amplitude curves between cells. In contrast, while the cumulative amplitude curves for field potential recordings do reach a steady state (Author response image 2a), field potential recordings in this context are not a reliable substitute for single cell or, in the case of MFB, singlebouton recordings. Postsynaptic cells in field potential recordings are not clamped, meaning that the massive release of glutamate due to continuous stimulation depolarizes the postsynaptic cells and reduces the driving force for Na+, irrespective of depletion of the RRP. This is supported by the fact that we consistently observed a recovery of fEPSP amplitudes later in the trains where RRP had presumably been maximally depleted. In summary, high-frequency stimulus trains at the field potential level are not a valid and established technique for estimating RRP size at MFS.

Specialized laboratories have used highly advanced techniques, such as paired recordings between individual MFB and postsynaptic CA3 pyramidal cells, to estimate the RRP size of MFB (Vandael et al., 2020). These approaches are outside the scope of our present study which, while elucidating functional changes following Bcl11b depletion and C1ql2 rescue, does not aim to provide a high-end biophysical analysis of the presynaptic mechanisms involved.

Estimation of RRP size using high-frequency stimulus trains at mossy fiber-CA3 synapses.

a Results from field potential recordings. Cumulative fEPSP amplitude in response to a train of 40 stimuli at 100 Hz. All subsequent peak amplitudes were normalized to the amplitude of the first peak. Data points corresponding to putative steady state responses were fit with linear regression (RRP size is indirectly reflected by the intersection of the regression line with the yaxis). Control+EGFP, 6 slices from 5 mice; Bcl11b cKO+EGFP, 6 slices from 3 mice. b Results from single-cell recordings. Cumulative EPSC amplitude in response to a train of 15 stimuli at 50 Hz. The last four stimuli were fit with linear regression. Control, 5 cells from 4 mice; Bcl11b cKO, 3 cells from 3 mice. Note the shallow onset of response amplitudes and the subsequent frequency potentiation. Due to the resulting increase in slope at higher stimulus numbers, intersection with the y-axis occurs at negative values. The differences shown were not found to be statistically significant; unpaired t-test or Mann-Whitney U-test.

Bcl11b KO reduces the number of synapses, yet the I-O curve reported in Supp Fig 2 is not changed. How is that possible? This should be explained.

We agree with reviewer #2– this apparent discrepancy has indeed struck us as a counterintuitive result. It might be that synapses that are preferentially eliminated in Bcl11b cKO are predominantly silent or have weak coupling strength, such that their loss has only a minimal effect on basal synaptic transmission. Although perplexing, the result is fully supported by our single-cell data which shows no significant differences in MF EPSC amplitudes recorded from CA3 pyramidal cells between controls and Bcl11b mutants (Author response image 3; please see the response below for details and also our response to Reviewer #1 question 2).

Matsuda et al DOI: 10.1016/j.neuron.2016.04.001 previously reported that C1ql2 organizes MF synapses by aligning postsynaptic kainate receptors with presynaptic elements. As this may have consequences for the functional properties of MF synapses including their plasticity, the authors should report whether they see deficient postsynaptic glutamate receptor signaling in the Bcl11b KO and rescue in the C1ql2 re-expression.

We agree that the study by Matsuda et al. is of key importance for our present work. Although MF LTP is governed by presynaptic mechanisms and we previously did not see differences in short-term plasticity between the control and Bcl11b cKO (De Bruyckere et al., 2018), the clustering of postsynaptic kainate receptors by C1ql2 is indeed an important detail that could potentially alter synaptic signaling at MFS in Bcl11b KO. We, therefore, re-analyzed previously recorded single-cell data by performing a kinetic analysis on MF EPSCs recorded from CA3 pyramidal cells in control and Bcl11b cKO mice (Figure for reviewers 3a) to evaluate postsynaptic AMPA and kainate receptor responses in both conditions. We took advantage of the fact that AMPA receptors deactivate roughly 10 times faster than kainate receptors, allowing the contributions of the two receptors to mossy fiber EPSCs to be separated (Castillo et al., 1997 and reviewed by Lerma, 2003). We fit the decay phase of the second (larger) EPSC evoked by paired-pulse stimulation with a double exponential function, yielding a fast and a slow component, which roughly correspond to the fractional currents evoked by AMPA and kainate receptors, respectively. Analysis of both fast and slow time constants and the corresponding fractional amplitudes revealed no significant differences between controls and Bcl11b mutants (Figure for reviewers 3e-h), indicating that both AMPA and kainate receptor signaling is unaffected by the ablation of C1ql2 following Bcl11b KO.

Importantly, MF EPSC amplitudes evoked by the first and the second pulse (Author response image 3b), paired-pulse facilitation (Author response image 3c) and failure rates (Author response image 3d) were all comparable between controls and Bcl11b mutants. These results further corroborate our observations from field recordings that basal synaptic transmission at MFS is unaltered by Bcl11b KO.

We note that the results from single cell recordings regarding basal synaptic transmission merely confirm the observations from field potential recordings, and that the attempted measurement of RRP size at the single cell level was not successful. Thus, our single-cell data do not add new information about the mechanisms underlying the effects of Bcl11b-deficiency and we therefore decided not to report these data in the manuscript.

Basal synaptic transmission at mossy fiber-CA3 synapses is unaltered in Bcl11b cKO mice.

a Representative average trace (20 sweeps) recorded from CA3 pyramidal cells in control and Bcl11b cKO mice at minimal stimulation conditions, showing EPSCs in response to paired-pulse stimulation (PPS) at an interstimulus interval of 40 ms. The signal is almost entirely blocked by the application of 2 μM DCG-IV (red). b Quantification of MF EPSC amplitudes in response to PPS for both the first and the second pulse. c Ratio between the amplitude of the second over the first EPSC. d Percentage of stimulation events resulting in no detectable EPSCs for the first pulse. Events <5 pA were considered as noise. e Fast decay time constant obtained by fitting the average second EPSC with the following double exponential function: I(t)=Afaste−t/τfast+Aslowe−t/τslow+C, where I is the recorded current amplitude after time t, Afast and Aslow represent fractional current amplitudes decaying with the fast (τfast) and slow (τslow) time constant, respectively, and C is the offset. Starting from the peak of the EPSC, the first 200 ms of the decaying trace were used for fitting. f Fractional current amplitude decaying with the fast time constant. g-h Slow decay time constant and fractional current amplitude decaying with the slow time constant. For all figures: Control, 8 cells from 4 mice; Bcl11b cKO, 8 cells from 6 mice. All data are presented as means, error bars indicate SEM. None of the differences shown were found to be statistically significant; Mann-Whitney U-test for nonnormally and unpaired t-test for normally distributed data.

Reviewer #3 (Public Review):

Overall, this is a strong manuscript that uses multiple current techniques to provide specific mechanistic insight into prior discoveries of the contributions of the Bcl11b transcription factor to mossy fiber synapses of dentate gyrus granule cells. The authors employ an adult deletion of Bcl11b via Tamoxifen-inducible Cre and use immunohistochemical, electron microscopy, and electrophysiological studies of synaptic plasticity, together with viral rescue of C1ql2, a direct transcriptional target of Bcl11b or Nrxn3, to construct a molecular cascade downstream of Bcl11b for DG mossy fiber synapse development. They find that C1ql2 re-expression in Bcl11b cKOs can rescue the synaptic vesicle docking phenotype and the impairments in MF-LTP of these mutants. They also show that C1ql2 knockdown in DG neurons can phenocopy the vesicle docking and plasticity phenotypes of the Bcl11b cKO. They also use artificial synapse formation assays to suggest that C1ql2 functions together with a specific Nrxn3 splice isoform in mediating MF axon development, extending these data with a C1ql2-K262E mutant that purports to specifically disrupt interactions with Nrxn3. All of the molecules involved in this cascade are disease-associated and this study provides an excellent blueprint for uncovering downstream mediators of transcription factor disruption. Together this makes this work of great interest to the field. Strengths are the sophisticated use of viral replacement and multi-level phenotypic analysis while weaknesses include the linkage of C1ql2 with a specific Nrxn3 splice variant in mediating these effects.

Here is an appraisal of the main claims and conclusions:

1. C1ql2 is a downstream target of Bcl11b which mediates the synaptic vesicle recruitment and synaptic plasticity phenotypes seen in these cKOs. This is supported by the clear rescue phenotypes of synapse anatomy (Fig.2) and MF synaptic plasticity (Fig.3). One weakness here is the absence of a control assessing over-expression phenotypes of C1ql2. It's clear from Fig.1D that viral rescue is often greater than WT expression (totally expected). In the case where you are trying to suppress a LoF phenotype, it is important to make sure that enhanced expression of C1ql2 in a WT background does not cause your rescue phenotype. A strong overexpression phenotype in WT would weaken the claim that C1ql2 is the main mediator of the Bcl11b phenotype for MF synapse phenotypes.

As suggested by reviewer #3, we carried out C1ql2 over-expression experiments in control animals. We show that the over-expression of C1ql2 in the DG of control animals had no effect on the synaptic vesicle organization in the proximity of MFS. This further supports that the observed effect upon rescue of C1ql2 expression in Bcl11b cKOs is due to the physiological function of C1ql2 and not a result of the artificial overexpression. These data are now included in supplementary figure 2g-j and are described in detail in the results part of the revised manuscript. Please also see response to point 3 of reviewer #2.

1. Knockdown of C1ql2 via 4 shRNAs is sufficient to produce the synaptic vesicle recruitment and MFLTP phenotypes. This is supported by clear effects in the shRNA-C1ql2 groups as compared to nonsense-EGFP controls. One concern (particularly given the use of 4 distinct shRNAs) is the potential for off-target effects, which is best controlled for by a rescue experiment with RNA insensitive C1ql2 cDNA as opposed to nonsense sequences, which may not elicit the same off-target effects.

We agree with reviewer #3 that the usage of shRNAs could potentially create unexpected off-target effects and that the introduction of a shRNA-insensitive C1ql2 in parallel to the expression on the shRNA cassette would be a very effective control experiment. However, the suggested experiment would require an additional 6 months (2 months for AAV production, 2-3 months from animal injection to sacrifice and 1-2 months for EM imaging/analysis and LTP measurements) and a high number of additional animals (minimum 8 for EM and 8 for LTP measurements). We note here, that before the production of the shRNA-C1ql2 and the shRNA-NS, the individual sequences were systematically checked for off-target bindings on the murine exome with up to two mismatches and presented with no other target except the proposed (C1ql2 for shRNA-C1ql2 and no target for shRNA-NS). Taking into consideration our in-silico analysis, we feel that the interpretation of our findings is valid without this (very reasonable) additional control experiment.

1. C1ql2 interacts with Nrxn3(25b+) to facilitate MF terminal SV clustering. This claim is theoretically supported by the HEK cell artificial synapse formation assay (Fig.5), the inability of the K262-C1ql2 mutation to rescue the Bcl11b phenotype (Fig.6), and the altered localization of C1ql2 in the Nrxn1-3 deletion mice (Fig.7). Each of these lines of experimental evidence has caveats that should be acknowledged and addressed. Given the hypothesis that C1ql2 and Nrxn3b(25b) are expressed in DG neurons and work together, the heterologous co-culture experiment seems strange. Up till now, the authors are looking at pre-synaptic function of C1ql2 since they are re-expressing it in DGNs. The phenotypes they are seeing are also pre-synaptic and/or consistent with pre-synaptic dysfunction. In Fig.5, they are testing whether C1ql2 can induce pre-synaptic differentiation in trans, i.e. theoretically being released from the 293 cells "post-synaptically". But the post-synaptic ligands (Nlgn1 and and GluKs) are not present in the 293 cells, so a heterologous synapse assay doesn't really make sense here. The effect that the authors are seeing likely reflects the fact that C1ql2 and Nrxn3 do bind to each other, so C1ql2 is acting as an artificial post-synaptic ligand, in that it can cluster Nrxn3 which in turn clusters synaptic vesicles. But this does not test the model that the authors propose (i.e. C1ql2 and Nrxn3 are both expressed in MF terminals). Perhaps a heterologous assay where GluK2 is put into HEK cells and the C1ql2 and Nrxn3 are simultaneously or individually manipulated in DG neurons?

C1ql2 is expressed by DG neurons and is then secreted in the MFS synaptic cleft, while Nrxn3, that is also expressed by DG neurons, is anchored at the presynaptic side. In our work we used the well established co-culture system assay and cultured HEK293 cells secreting C1ql2 (an IgK secretion sequence was inserted at the N-terminus of C1ql2) together with hippocampal neurons expressing Nrxn3(25b+). We used the HEK293 cells as a delivery system of secreted C1ql2 to the neurons to create regions of high concentration of C1ql2. By interfering with the C1ql2-Nrxn3 interaction in this system either by expression of the non-binding mutant C1ql2 variant in the HEK cells or by manipulating Nrxn expression in the neurons, we could show that C1ql2 binding to Nrxn3(25b+) is necessary for the accumulation of vGlut1. However, we did not examine and do not claim within our manuscript that the interaction between C1ql2 and Nrxn3(25b+) induces presynaptic differentiation. Our experiment only aimed to analyze the ability of C1ql2 to cluster SV through interaction with Nrxn3. Moreover, by not expressing potential postsynaptic interaction partners of C1ql2 in our system, we could show that C1ql2 controls SV recruitment through a purely presynaptic mechanism. Co-culturing GluK2-expressing HEK cells with simultaneous manipulation of C1ql2 and/or Nrxn3 in neurons would not allow us to appropriately answer our scientific question, but rather focus on the potential synaptogenic function of the Nrxn3/C1ql2/GluK2 complex and the role of the postsynaptic ligand in it. Thus, we feel that the proposed experiment, while very interesting in characterization of additional putative functions of C1ql2, may not provide additional information for the point we were addressing. In the revised manuscript we tried to make the aim and methodological approach of this set of experiments more clear.

1. K262-C1ql2 mutation blocks the normal rescue through a Nrxn3(25b) mechanism (Fig.6). The strength of this experiment rests upon the specificity of this mutation for disrupting Nrxn3b binding (presynaptic) as opposed to any of the known postsynaptic C1ql2 ligands such as GluK2. While this is not relevant for interpreting the heterologous assay (Fig.5), it is relevant for the in vivo phenotypes in Fig.6. Similar approaches as employed in this paper can test whether binding to other known postsynaptic targets is altered by this point mutation.

It has been previously shown that C1ql2 together with C1ql3 recruit postsynaptic GluK2 at the MFS. However, loss of just C1ql2 did not affect the recruitment of GluK2, which was disrupted only upon loss of both C1ql2 and C1ql3 (Matsuda et al., 2018). In our study we demonstrate a purely presynaptic function of C1ql2 through Nrxn3 in the synaptic vesicle recruitment. This function is independent of C1ql3, as C1ql3 expression is unchanged in all of our models and its over-expression did not compensate for C1ql2 functions (Fig. 2, 3a-c). Our in vitro experiments also reveal that C1ql2 can recruit both Nrxn3 and vGlut1 in the absence of any known postsynaptic C1ql2 partner (KARs and BAI3; Fig.5; please also see response above). Furthermore, we have now performed a kinetic analysis on single-cell data which we had previously collected to evaluate postsynaptic AMPA and kainate receptor responses in both the control and Bcl11b KO. Our analysis reveals no significant differences in postsynaptic current kinetics, making it unlikely that AMPA and kainate receptor signaling is altered upon the loss of C1ql2 following Bcl11b cKO (Author response image 3e-h; please also see our response to reviewer #2 point 8). Thus, we have no experimental evidence supporting the idea that a loss of interaction between C1ql2.K262E and GluK2 would interfere with the examined phenotype. However, to exclude that the K262E mutation disrupts interaction between C1ql2 and GluK2, we performed co-immunoprecipitation from protein lysate of HEK293 cells expressing GluK2myc-flag and GFP-C1ql2 or GluK2-myc-flag and GFP-K262E and could show that both C1ql2 and K262E had GluK2 bound when precipitated. These data are included in supplementary figure 5k of the revised manuscript.

1. Altered localization of C1ql2 in Nrxn1-3 cKOs. These data are presented to suggest that Nrx3(25b) is important for localizing C1ql2 to the SL of CA3. Weaknesses of this data include both the lack of Nrxn specificity in the triple a/b KOs as well as the profound effects of Nrxn LoF on the total levels of C1ql2 protein. Some measure that isn't biased by this large difference in C1ql2 levels should be attempted (something like in Fig.1F).

We acknowledge that the lack of specificity in the Nrxn123 model makes it difficult to interpret our data. We have now examined the mRNA levels of Nrxn1 and Nrxn2 upon stereotaxic injection of Cre in the DG of Nrxn123flox/flox animals and found that Nrxn1 was only mildly reduced. At the same time Nrxn2 showed a tendency for reduction that was not significant (data included in supplementary figure 6a of revised manuscript). Only Nrxn3 expression was strongly suppressed. Of course, this does not exclude that the mild reduction of Nrxn1 and Nrxn2 interferes with the C1ql2 localization at the MFS. We further examined the mRNA levels of C1ql2 in control and Nrxn123 mutants to ensure that the observed changes in C1ql2 protein levels at the MFS are not due to reduced mRNA expression and found no changes (data are included in supplementary figure 6b of the revised manuscript), suggesting that overall protein C1ql2 expression is normal.

The reduced C1ql2 fluorescence intensity at the MFS was first observed when non-binding C1ql2 variant K262E was introduced to Bcl11b cKO mice that lack endogenous C1ql2 (Fig.6). In these experiments, we found that despite the overall high protein levels of C1ql2.K262E in the hippocampus (Fig. 6c), its fluorescence intensity at the SL was significantly reduced compared to WT C1ql2 (Fig. 6d-e). The remaining signal of the C1ql2.K262E at the SL was equally distributed and in a punctate form, similar to WT C1ql2. Together, this suggests that loss of C1ql2-Nrxn3 interaction interferes with the localization of C1ql2 at the MFS, but not with the expression of C1ql2. Of course, this does not exclude that other mechanisms are involved in the synaptic localization of C1ql2, beyond the interaction with Nrxn3, as both the mutant C1ql2 in Bcl11b cKO and the endogenous C1ql2 in Nrxn123 cKOs show residual immunofluorescence at the SL. Further studies are required to determine how C1ql2-Nrxn3 interaction regulates C1ql2 localization at the MFS.

Reviewer #1 (Recommendations For The Authors):

In addition to addressing the comments below, this study would benefit significantly from providing insight and discussion into the relevant potential postsynaptic signaling components controlled exclusively by C1ql2 (postsynaptic kainate receptors and the BAI family of proteins).

We have now performed a kinetic analysis on single-cell data that we had previously collected to evaluate postsynaptic AMPA and kainate receptor responses in both the control and Bcl11b cKO. Our analysis reveals no significant differences in postsynaptic current kinetics, making it unlikely that AMPA and kainate receptor signaling differ between controls and upon the loss of C1ql2 following Bcl11b cKO (Author response image 3e-h; please also see our response to Reviewer #2 point 8). This agrees with previous findings that C1ql2 regulates postsynaptic GluK2 recruitment together with C1ql3 and only loss of both C1ql2 and C1ql3 results in a disruption of KAR signaling (Matsuda et al., 2018). In our study we demonstrate a purely presynaptic function of C1ql2 through Nrxn3 in the synaptic vesicle recruitment. This function is independent of C1ql3, as C1ql3 expression is unchanged in all of our models and its over-expression did not compensate for C1ql2 functions (Fig. 2, 3a-c). Our in vitro experiments also reveal that C1ql2 can recruit both Nrxn3 and vGlut1 in the absence of any known postsynaptic C1ql2 partner (KARs and BAI3; Fig.5; please also see our response to reviewer #3 point 4 above). We believe that further studies are needed to fully understand both the pre- and the postsynaptic functions of C1ql2. Because the focus of this manuscript was on the role of the C1ql2-Nrxn3 interaction and our investigation on postsynaptic functions of C1ql2 was incomplete, we did not include our findings on postsynaptic current kinetics in our revised manuscript. However, we increased the discussion on the known postsynaptic partners of C1ql2 in the revised manuscript to increase the interpretability of our results.

Major Comments:

The authors demonstrate that the ultrastructural properties of presynaptic boutons are altered after Bcl11b KO and C1ql2 KD. However, whether C1ql2 functions as part of a tripartite complex and the identity of the postsynaptic receptor (BAI, KAR) should be examined.

Matsuda and colleagues have nicely demonstrated in their 2016 (Neuron) study that C1ql2 is part of a tripartite complex with presynaptic Nrxn3 and postsynaptic KARs. Moreover, they demonstrated that C1ql2, together with C1ql3, recruit postsynaptic KARs at the MFS, while the KO of just C1ql2 did not affect the KAR localization. In our study we demonstrate a purely presynaptic function of C1ql2 through Nrxn3 in the synaptic vesicle recruitment. This function is independent of C1ql3, as C1ql3 expression is unchanged in all of our models and its over-expression did not compensate for C1ql2 functions (Fig. 2, 3a-c). Our in vitro experiments also reveal that C1ql2 is able to recruit both Nrxn3 and vGlut1 in the absence of any known postsynaptic C1ql2 partner (Fig. 5; please also see our response to reviewer #3 point 4 above). Moreover, we were able to show that the SV recruitment depends on C1ql2 interaction with Nrxn3 through the expression of a non-binding C1ql2 (Fig. 6) that retains the ability to interact with GluK2 (supplementary figure 5k of revised manuscript) or by KO of Nrxns (Fig. 7). Furthermore, we have now performed a kinetic analysis on single-cell data which we had previously collected to evaluate postsynaptic AMPA and kainate receptor responses in both the control and Bcl11b cKO. Our analysis reveals no significant differences in postsynaptic current kinetics, making it unlikely that AMPA and kainate receptor signaling differ between controls and Bcl11b mutants (Author response image 3e-h; please also see our response to Reviewer #2 question 8). Together, we have no experimental evidence so far that would support that the postsynaptic partners of C1ql2 are involved in the observed phenotype. While it would be very interesting to characterize the postsynaptic partners of C1ql2 in depth, we feel this would be beyond the scope of the present study.

Figure 1f: For a more comprehensive understanding of the Bcl11b KO phenotype and the potential role for C1ql2 on MF synapse number, a complete quantification of vGlut1 and Homer1 for all conditions (Supplement Figure 2e) should be included in the main text.

In our study we focused on the role of C1ql2 in the structural and functional integrity of the MFS downstream of Bcl11b. Bcl11b ablation leads to several phenotypes in the MFS that have been thoroughly described in our previous study (De Bruyckere et al., 2018). As expected, re-expression of C1ql2 only partially rescued these phenotypes, with full recovery of the SV recruitment (Fig. 2) and of the LTP (Fig. 3), but had no effect on the reduced numbers of MFS nor the structural complexity of the MFB created by the Bcl11b KO (supplementary figure 2d-f of revised manuscript). We understand that including the quantification of vGlut1 and Homer1 co-localization in the main figures would help with a better understanding of the Bcl11b mutant phenotype. However, in our manuscript we investigate C1ql2 as an effector of Bcl11b and thus we focus on its functions in SV recruitment and LTP. As we did not find a link between C1ql2 and the number of MFS/MFB upon re-expression of C1ql2 in Bcl11b cKO or now also in C1ql2 KD (see response to comment #4 below), we believe it is more suitable to present these data in the supplement.

Figure 3/4: Given the striking reduction in the numbers of synapses (Supplement Figure 2e) and docked vesicles (Figure 2d) in the Bcl11b KO and C1ql2 KD (Figure 4e-f), it is extremely surprising that basal synaptic transmission is unaffected (Supplement Figure 2g). The authors should determine the EPSP input-output relationship following C1ql2 KD and measure EPSPs following trains of stimuli at various high frequencies.

We fully acknowledge that this is an unexpected result. It is, however, well feasible that the modest displacement of SV fails to noticeably influence basal synaptic transmission. This would be the case, for example, if only a low number of vesicles are released by single stimuli, in line with the very low initial Pr at MFS. In contrast, the reduction in synapse numbers in the Bcl11b mutant might indeed be expected to reflect in the input-output relationship. It is possible, however, that synapses that are preferentially eliminated in Bcl11b cKO are predominantly silent or have weak coupling strength, such that their loss has only a minimal effect on basal synaptic transmission. Finally, we cannot exclude compensatory mechanisms (homeostatic plasticity) at the remaining synapses. A detailed analysis of these potential mechanisms would be a whole project in its own right.

As additional information, we can say that the largely unchanged input-output-relation in Bcl11b cKO is also present in the single-cell level data (Author response image 3; details on single-cell experiments are described in the response to Reviewer #2 point 8).

As suggested by the reviewer, we have now additionally analyzed the input-output relationship following C1ql2 KD and again did not observe any significant difference between control and KD animals. We have incorporated the respective input-output curves into the revised manuscript under Supplementary figure 3c-d.

Figure 4: Does C1ql2 shRNA also reduce the number of MFBs? This should be tested to further identify C1ql2-dependent and independent functions.

As requested by reviewer #1 we quantified the number of MFBs upon C1ql2 KD. We show that C1ql2 KD in WT animals does not alter the number of MFBs. The data are presented in supplementary figure 4d of the revised manuscript. Re-expression of C1ql2 in Bcl11b cKO did not rescue the loss of MFS created by the Bcl11b mutation. Moreover, C1ql2 re-expression did not rescue the complexity of the MFB ultrastructure perturbed by the Bcl11b ablation. Together, this suggests that Bcl11b regulates MFs maintenance through additional C1ql2-independent pathways. In our previously published work (De Bruyckere et al., 2018) we identified and discussed in detail several candidate genes such as Sema5b, Ptgs2, Pdyn and Penk as putative effectors of Bcl11b in the structural and functional integrity of MFS (please also see response to reviewer #2- point 1 of public reviews).

Figure 5: Clarification is required regarding the experimental design of the HEK/Neuron co-culture: 1. C1ql2 is a secreted soluble protein - how is the protein anchored to the HEK cell membrane to recruit Nrxn3(25b+) binding and, subsequently, vGlut1?

C1ql2 was secreted by the HEK293 cells through an IgK signaling peptide at the N-terminus of C1ql2. The high concentration of C1ql2 close to the secretion site together with the sparse coculturing of the HEK293 cells on the neurons allows for the quantification of accumulation of neuronal proteins. We have now described the experimental conditions in greater detail in the main text module of the revised manuscript

1. Why are the neurons transfected and not infected? Transfection efficiency of neurons with lipofectamine is usually poor (1-5%; Karra et al., 2010), while infection of neurons with lentiviruses or AAVs encoding cDNAs routinely are >90% efficient. Thus, interpretation of the recruitment assays may be influenced by the density of neurons transfected near a HEK cell.

We agree with reviewer #1 that viral infection of the neurons would have been a more effective way of expressing our constructs. However, due to safety allowances in the used facility and time limitation at the time of conception of this set of experiments, a lipofectamine transfection was chosen.

However, as all of our examined groups were handled in the same way and multiple cells from three independent experiments were examined for each experimental set, we believe that possible biases introduced by the transfection efficiency have been eliminated and thus have trust in our interpretation of these results.

1. Surface labeling of HEK cells for wild-type C1ql2 and K262 C1ql2 would be helpful to assess the trafficking of the mutant.

We recognize that potential changes to the trafficking of C1ql2 caused by the K262E mutation would be important to characterize, in light of the reduced localization of the mutant protein at the SL in the in vivo experiments (Fig. 6e). In our culture system, C1ql2 and K262E were secreted by the HEK cells through insertion of an IgK signaling peptide at the N-terminus of the myc-tagged C1ql2/K262E. Thus, trafficking analysis on this system would not be informative, as the system is highly artificial compared to the in vivo model. Further studies are needed to characterize C1ql2 trafficking in neurons to understand how C1ql2-Nrxn3 interaction regulates the localization of C1ql2. However, labeling of the myc-tag in C1ql2 or K262E expressing HEK cells of the co-culture model reveals a similar signal for the two proteins (Fig. 5a,c). Nrxn-null mutation in neurons co-cultured with C1ql2-expressing HEK cells disrupted C1ql2 mediated vGlut1 accumulation in the neurons. Selective expression of Nrxn3(25b) in the Nrxn-null neurons restored vGlut1 clustering was (Fig. 5e-f). Together, these data suggest that it is the interaction between C1ql2 and Nrxn3 that drives the accumulation of vGlut1.

Figure 6: Bcl11b KO should also be included in 6f-h.

As suggested by reviewer #1, we included the Bcl11b cKO in figures 6f-h and in corresponding supplementary figures 5c-j.

Figure 7b: What is the abundance of mRNA for Nrxn1 and Nrxn2 as well as the abundance of Nrxns after EGFP-Cre injection into DG?

We addressed this point raised by reviewer #1 by quantifying the relative mRNA levels of Nrxn1 and Nrxn2 via qPCR upon Nrxn123 mutation induction with EGFP-Cre injection. We have now examined the mRNA levels of Nrxn1 and Nrxn2 upon stereotaxic injection of Cre in the DG of Nrxn123flox/flox animals and found that Nrxn1 was only mildly reduced. At the same time Nrxn2 showed a tendency for reduction that was not significant. The data are presented in supplementary figure 6a of the revised maunscript.

Minor Comments for readability:

Synapse score is referred to frequently in the text and should be defined within the text for clarification.

'n' numbers should be better defined in the figure legends. For example, for protein expression analysis in 1c, n=3. Is this a biological or technical triplicate? For electrophysiology (e.g. 3c), does "n=7" reflect the number of animals or the number of slices? n/N (slices/animals) should be presented.

Figure 7a: Should the diagrams of the cre viruses be EGFP-Inactive or active Cre and not CRE-EGFP as shown in the diagram?

Figure 7b: the region used for the inset should be identified in the larger image.

All minor points have been fixed in the revised manuscript according to the suggestions.

Reviewer #3 (Recommendations For The Authors):

-Please describe the 'synapse score' somewhere in the text - it is too prominently featured to not have a clear description of what it is.

The description of the synapse score has been included in the main text module of the revised manuscript.

-The claim that Bcl11b controls SV recruitment "specifically" through C1ql2 is a bit stronger than is warranted by the data. Particularly given that C1ql2 is expressed at 2.5X control levels in their rescue experiments. See pt.2

Please see response to reviewer #3 point 1 of public reviews. To address this, we over-expressed C1ql2 in control animals and found no changes in the synaptic vesicle distribution (supplementary figure 2g-j of revised manuscript). This supports that the observed rescue of synaptic vesicle recruitment by re-expression of C1ql2 is due to its physiological function and not due to the artificially elevated protein levels. Of course, we cannot exclude the possibility that other, C1ql2-independent, mechanisms also contribute to the SV recruitment downstream of Bcl11b. Our data from the C1ql2 rescue, C1ql2 KD, the in vitro experiments and the interruption of C1ql2-Nrxn3 in vivo, strongly suggest C1ql2 to be an important regulator of SV recruitment.

-Does Bcl11b regulate Nrxn3 expression? Considering the apparent loss of C1ql2 expression in the Nrxn KO mice, this is an important detail.

We agree with reviewer #3 that this is an important point. We have previously done differential transcriptomics from DG neurons of Bcl11b cKOs compared to controls and did not find Nrxn3 among the differentially expressed genes. To further validate this, we now quantified the Nrxn3 mRNA levels via qPCR in Bcl11b cKOs compared to controls and found no differences. These data are included in supplementary figure 5a of the revised manuscript.

-It appears that C1ql2 expression is much lower in the Nrxn123 KO mice. Since the authors are trying to test whether Nrxn3 is required for the correct targeting of C1ql2, this is a confounding factor. We can't really tell if what we are seeing is a "mistargeting" of C1ql2, loss of expression, or both. If the authors did a similar analysis to what they did in Figure 1 where they looked at the synaptic localization of C1ql2 (and quantified it) that could provide more evidence to support or refute the "mistargeting" claim.

Please also see response to reviewer #3 point 5 of public reviews. To exclude that reduction of fluorescence intensity of C1ql2 at the SL in Nrxn123 KO mice is due to loss of C1ql2 expression, we examined the mRNA levels of C1ql2 in control and Nrxn123 mutants and found no changes (data are included in supplementary figure 6b of the revised manuscript), suggesting that C1ql2 gene expression is normal. The reduced C1ql2 fluorescence intensity at the MFS was first observed when non-binding C1ql2 variant K262E was introduced to Bcl11b cKO mice that lack endogenous C1ql2 (Fig.6). In these experiments, we found that despite the overall high protein levels of C1ql2.K262E in the hippocampus (Fig. 6c), its fluorescence intensity at the SL was significantly reduced compared to WT C1ql2 (Fig. 6d-e). The remaining C1ql2.K262E signal in the SL was equally distributed and in a punctate form, similar to WT C1ql2. Together, this indicates that the loss of C1ql2-Nrxn3 interaction interferes with the localization of C1ql2 along the MFS, but not with expression of C1ql2. Of course, this does not exclude that additional mechanisms regulate C1ql2 localization at the synapse, as both the mutant C1ql2 in Bcl11b cKO and the endogenous C1ql2 in Nrxn123 cKO show residual immunofluorescence at the SL.

We note here that we have not previously quantified the co-localization of C1ql2 with individual synapses. C1ql2 is a secreted molecule that localizes at the MFS synaptic cleft. However, not much is known about the number of MFS that are positive for C1ql2 nor about the mechanisms regulating C1ql2 targeting, transport, and secretion to the MFS. Whether C1ql2 interaction with Nrxn3 is necessary for the protection of C1ql2 from degradation, its surface presentation and transport or stabilization to the synapse is currently unclear. Upon revision of our manuscript, we realized that we might have overstated this particular finding and have now rephrased the specific parts within the results to appropriately describe the observation and have also included a sentence in the discussion referring to the lack of understanding of the mechanism behind this observation.

-Title of Figure S5 is "Nrxn KO perturbs C1ql2 localization and SV recruitment at the MFS", but there is no data on C1ql2 localization.

This issue has been fixed in the revised manusript.

-S5 should be labeled more clearly than just Cre+/-

This issue has been fixed in the revised manuscript.

References

Castillo, P.E., Malenka, R.C., Nicoll, R.A., 1997. Kainate receptors mediate a slow postsynaptic current in hippocampal CA3 neurons. Nature 388, 182–186. https://doi.org/10.1038/40645

De Bruyckere, E., Simon, R., Nestel, S., Heimrich, B., Kätzel, D., Egorov, A.V., Liu, P., Jenkins, N.A., Copeland, N.G., Schwegler, H., Draguhn, A., Britsch, S., 2018. Stability and Function of Hippocampal Mossy Fiber Synapses Depend on Bcl11b/Ctip2. Front. Mol. Neurosci. 11. https://doi.org/10.3389/fnmol.2018.00103

Kaeser, P.S., Regehr, W.G., 2017. The readily releasable pool of synaptic vesicles. Curr. Opin. Neurobiol. 43, 63–70. https://doi.org/10.1016/j.conb.2016.12.012

Lerma, J., 2003. Roles and rules of kainate receptors in synaptic transmission. Nat. Rev. Neurosci. 4, 481–495. https://doi.org/10.1038/nrn1118

Orlando, M., Dvorzhak, A., Bruentgens, F., Maglione, M., Rost, B.R., Sigrist, S.J., Breustedt, J., Schmitz, D., 2021. Recruitment of release sites underlies chemical presynaptic potentiation at hippocampal mossy fiber boutons. PLoS Biol. 19, e3001149. https://doi.org/10.1371/journal.pbio.3001149

Vandael, D., Borges-Merjane, C., Zhang, X., Jonas, P., 2020. Short-Term Plasticity at Hippocampal Mossy Fiber Synapses Is Induced by Natural Activity Patterns and Associated with Vesicle Pool Engram Formation. Neuron 107, 509-521.e7. https://doi.org/10.1016/j.neuron.2020.05.013

Version 2

Abstract

Abstract

The transcription factor Bcl11b has been linked to neurodevelopmental and neuropsychiatric disorders associated with synaptic dysfunction. Bcl11b is highly expressed in dentate gyrus granule neurons and is required for the structural and functional integrity of mossy fiber-CA3 synapses. The underlying molecular mechanisms, however, remained unclear. We show that the synaptic organizer molecule C1ql2 is a direct functional target of Bcl11b that regulates synaptic vesicle recruitment and long-term potentiation at mossy fiber-CA3 synapses in vivo and in vitro. Furthermore, we demonstrate C1ql2 to exert its functions through direct interaction with a specific splice variant of neurexin-3, Nrxn3(25b+). Interruption of C1ql2-Nrxn3(25b+) interaction by expression of a non-binding C1ql2 mutant or by deletion of Nrxn3 in the dentate gyrus granule neurons recapitulates major parts of the Bcl11b as well as C1ql2 mutant phenotype. Together, this study identifies a novel C1ql2-Nrxn3(25b+)-dependent signaling pathway through which Bcl11b controls mossy fiber-CA3 synapse function. Thus, our findings contribute to the mechanistic understanding of neurodevelopmental disorders accompanied by synaptic dysfunction.

Introduction

Disruptions in synaptic structure and function have been identified as a major determinant in the manifestation of various neurodevelopmental and neuropsychiatric disorders, such as autism spectrum disorder, schizophrenia, and intellectual disability (Hayashi-Takagi, 2017; Lepeta et al., 2016; Zoghbi & Bear, 2012). The need to understand how synaptic function is compromised in these disorders has accentuated the importance of studying the regulatory mechanisms of physiological synaptic function. These mechanisms involve cell adhesion molecules at both the pre– and post-synaptic side that act as synaptic organizers, whose unique combination determines the structural and functional properties of the synapse. Many such proteins have already been identified and our understanding of their complex role in synapse assembly and function has significantly increased over the last years (de Wit & Ghosh, 2016; O’Rourke et al., 2012; Sudhof, 2017). Furthermore, recent advances in the genetics of neurodevelopmental and neuropsychiatric disorders have implicated genes encoding for several of the known synaptic proteins, supporting a role for these molecules in the pathogenesis of corresponding disorders (Sudhof, 2021; Torres et al., 2017; Wang et al., 2018).

The sensitivity of the functional specification of the synapse to the combination of distinct synaptic proteins and their relative expression levels suggests that genetically encoded programs define at least facets of the synaptic properties in a cell-type-specific manner (Sudhof, 2017, 2021). Several of the synaptic proteins have been shown to promote formation of functional pre– and postsynaptic assemblies when presented in non-neuronal cells (Dalva et al., 2000; Dean et al., 2003; Scheiffele et al., 2000), showing that their ability to specify synapses is in part independent of signaling processes and neuronal activity and supporting the idea that synaptic function is governed by cues linked to cellular origin (Gomez et al., 2021). Thus, the investigation of synaptic organizers and their function in health and disease should be expanded to the transcriptional programs that regulate their expression.

Bcl11b (also known as Ctip2) is a zinc finger transcription factor that has been implicated in various disorders of the nervous system including Alzheimer’s and Huntington’s disease, and schizophrenia (Kunkle et al., 2016; Song et al., 2022; Whitton et al., 2018; Whitton et al., 2016). Patients with BCL11B mutations present with neurodevelopmental delay, overall learning deficits as well as impaired speech acquisition and autistic features (Eto et al., 2022; Lessel et al., 2018; Punwani et al., 2016; Yang et al., 2020). Bcl11b is expressed in several neuron types, including the dentate gyrus granule neurons (DGN) of the hippocampus. Expression of Bcl11b in the DGN starts during embryonic development and persists into adulthood (Simon et al., 2020). We have previously demonstrated that Bcl11b plays a crucial role in the development of the hippocampal mossy fiber system, adult hippocampal neurogenesis as well as hippocampal learning and memory (Simon et al., 2016; Simon et al., 2012). In the mature hippocampus, Bcl11b is critical for the structural and functional integrity of mossy fiber synapses (MFS), the connections between DGN and CA3 pyramidal neurons (De Bruyckere et al., 2018). MFS have a critical role in learning and memory stemming from their unique structural and functional properties, such as an enormous pool of releasable synaptic vesicles (SV), and reliable presynaptic short– and long-term plasticity (Nicoll & Schmitz, 2005; Rollenhagen & Lubke, 2010). Conditional ablation of Bcl11b in murine DGN impairs presynaptic recruitment of SV and abolishes mossy fiber long-term potentiation (MF-LTP) (De Bruyckere et al., 2018). The molecular mechanisms, however, through which the transcriptional regulator Bcl11b controls highly dynamic properties of the MFS remained elusive.

In the present study, we show that the secreted synaptic organizer molecule C1ql2, a member of the C1q-like protein family (Yuzaki, 2017), is a functional target of Bcl11b in murine DGN. Reintroduction of C1ql2 in Bcl11b mutant DGN rescued the localization and docking of SV to the active zone (AZ), as well as MF-LTP that was abolished upon Bcl11b ablation. Knock-down (KD) of C1ql2 in wildtype animals recapitulated a major part of the MFS phenotype observed in Bcl11b mutants. Furthermore, we show that C1ql2 requires direct interaction with a specific neurexin-3 isoform, Nrxn3(25b+), a member of a polymorphic family of presynaptic cell adhesion molecules (Reissner et al., 2013; Sudhof, 2017), to recruit SV in vitro and in vivo. Finally, we observe that localization of C1ql2 along the mossy fiber tract depends on C1ql2-Nrxn3(25b+) interaction. Taken together, this study identifies a novel Bcl11b/C1ql2/Nrxn3(25b+)-dependent regulatory mechanism that is essential for the control of MFS function. Recent genetic studies suggested its single components to be associated with neurodevelopmental and neuropsychiatric disorders characterized by synaptic dysfunction. Our data, for the first time, demonstrate these molecules to be interconnected in one regulatory pathway. Thus, our findings provide new mechanistic insight into the pathogenesis of corresponding human disorders.

Results

Reintroduction of C1ql2 into Bcl11b mutant dentate granule neurons restores synaptic vesicle recruitment at the mossy fiber-CA3 synapse

We demonstrated before that Bcl11b is critical for the structural and functional integrity of adult excitatory hippocampal MFS (De Bruyckere et al., 2018; Simon et al., 2012). The downstream regulatory mechanisms, however, through which Bcl11b exerts its complex functions at the MFS remained unclear. In a previous study, we carried out differential transcriptomic analyses on Bcl11b conditional knock-out (cKO) and wildtype (WT) DGN, to systematically screen for candidate transcriptional targets of Bcl11b (De Bruyckere et al., 2018). Among the differentially expressed candidate genes, we identified the synaptic organizer molecule C1ql2 (De Bruyckere et al., 2018), previously implicated in modulating MFS functions (Matsuda et al., 2016). C1ql2 transcript and protein levels are massively downregulated in Bcl11b mutant DGN (Supplementary figure 1a-b; De Bruyckere et al., 2018), and the Bcl11b protein directly binds to consensus sequences within the C1ql2 promotor (De Bruyckere et al., 2018), suggesting Bcl11b to act on MFS through C1ql2. To directly test this, we stereotaxically injected a C1ql2-expressing AAV (Figure 1a-b) into the dentate gyrus (DG) of Bcl11b cKO mice 2 weeks after induction of the mutation and compared them to control animals. To avoid potential interference of the AAV-mediated gene expression with the interpretability of observed phenotypes, we stereotaxically injected the DG of control animals as well, with AAV expressing EGFP only. AAV-mediated re-expression of C1ql2 in the Bcl11b mutant DGN completely restored C1ql2 protein expression (Figure 1c-d; Control+EGFP: 1±0.216, Bcl11b cKO+EGFP: 0.2±0.023, Bcl11b cKO+EGFP-2A-C1ql2: 2.44±0.745, mean±SEM). Furthermore, the spatial distribution of the exogenous C1ql2 protein in mutants was indistinguishable from controls (Figure 1e). Using vGlut1 and Homer1 as pre– and postsynaptic markers, respectively, we observed exogenous C1ql2 protein to precisely localize at glutamatergic synapses within the stratum lucidum (SL) of CA3, confirming that reintroduced C1ql2 is correctly targeted to the MFS (Figure 1f).

Stereotaxic injection of C1ql2-expressing AAV into Bcl11b cKO DGN restores C1ql2 levels.

a Experimental design to analyze the functions of C1ql2 in the MFS as a downstream target of Bcl11b. b AAV constructs injected in the DG of Bcl11b cKO and control littermates. c Western blot and d relative C1ql2 protein levels in mouse hippocampal homogenates. N=3. All data are presented as means; error bars indicate SEM. Two-way ANOVA and Tuckey’s PHC. Control+EGFP vs. Bcl11b cKO+EGFP-2A-C1ql2: ns, p=0.11; Bcl11b cKO+EGFP-2A-C1ql2 vs. Bcl11b cKO+EGFP: *p=0.015; Bcl11b cKO+EGFP-2A-C1ql2 vs. Bcl11b cKO+EGFP-2A-C1ql3: *p=0.019; ns, not significant. e Immunohistochemistry of C1ql2 (cyan), GFP (green), and Bcl11b (red) on hippocampal sections. Scale bar: 200 μm. f Immunohistochemistry of C1ql2 (cyan), vGlut1 (magenta) and Homer1 (yellow) in the SL of CA3. White arrowheads indicate co-localizing puncta of all three proteins. Scale bar: 15 μm

MFS of Bcl11b cKO animals were characterized by a misdistribution of SV in relation to the AZ, with fewer SV being present in the vicinity of AZ, as reflected by a lower average synapse score (Figure 2a-b). The scoring system used in this study rates MFS based on the number of SV and their distance from the AZ, is described in detail in the Materials and Methods part, and has been published previously (De Bruyckere et al., 2018). Reintroduction of C1ql2 fully recovered the synapse score to control values (Figure 2a-b; Control+EGFP: 3.4±0.012, Bcl11b cKO+EGFP: 2.96±0.037, Bcl11b cKO+EGFP-2A-C1ql2: 3.47±0.043, mean±SEM). As revealed by the relative frequency of the individual synapse scores, C1ql2 not only reduced the number of inactive synapses, characterized by a synapse score of 0, but also improved the synapse score of active synapses (Figure 2c). To test for the specificity of the C1ql2 effects, we overexpressed C1ql3 in the DGN of Bcl11b cKO (Figure 1a-b). C1ql3, a different member of the C1ql subfamily, is co-expressed with C1ql2 in DGN and the two proteins have been shown to form functional heteromers (Matsuda et al., 2016). C1ql3 expression is unchanged in Bcl11b cKO (Supplementary figure 2a-b; Control+EGFP: 1±0.022, Bcl11b cKO+EGFP: 1.09±0.126, Bcl11b cKO+EGFP-2A-C1ql2: 0.87±0.146, mean±SEM). Overexpression of C1ql3 in the DGN of Bcl11b cKO neither interfered with C1ql2 expression levels (Figure 1c-e; Bcl11b cKO+EGFP-2A-C1ql3: 0.29±0.042, mean ± SEM) nor was it able to rescue the synapse score of Bcl11b mutants (Figure 2a-c; Bcl11b cKO+EGFP-2A-C1ql3: 2.97±0.062, mean±SEM). While not significant, AAV-mediated re-expression of C1ql2 in Bcl11b cKO led to artificially elevated C1ql2 protein levels compared to controls (Figure 1c-d). To exclude that the observed effects were influenced by the elevated C1ql2 expression in the Bcl11b cKO background above physiological levels, we over-expressed C1ql2 as well in control animals, which resulted in a strong increase of C1ql2 (Figure 2h). However, this did not affect the average synapse score (Figure 2i-j; Control+EGFP: 3.4±0.012; Control+EGFP-2A-C1ql2: 3.41±0.031; Bcl11b cKO+EGFP-2A-C1ql2: 3.47±0.043, mean±SEM).

C1ql2 reintroduction in Bcl11b cKO DGN rescues SV recruitment in MFS.

a Electron microscope images of MFS and proximal SV. White bars mark synapse length from the postsynaptic side. Scale bar: 100 nm. b Average synapse score. Control+EGFP, N=3; Bcl11b cKO+EGFP, Bcl11b cKO+EGFP-2A-C1ql2, Bcl11b cKO+EGFP-2A-C1ql3, N=4. Two-way ANOVA and Tuckey’s PHC. Control+EGFP vs. Bcl11b cKO+EGFP: ***p=0.0002, and vs. Bcl11b cKO+EGFP-2A-C1ql3: ***p=0.0003; Bcl11b cKO+EGFP-2A-C1ql2 vs. Bcl11b cKO+EGFP and vs. Bcl11b cKO+EGFP-2A-C1ql3: ***p<0.0001; ns, not significant. c Relative frequency of synapse scores. d Number of docked vesicles per 100 nm AZ profile length. Control+EGFP, Bcl11b cKO+EGFP-2A-C1ql3, N=3; Bcl11b cKO+EGFP, Bcl11b cKO+EGFP-2A-C1ql2, N=4. All data are presented as means; error bars indicate SEM. Points represent the individual examined AZ and SV, respectively. Two-way ANOVA and Tuckey’s PHC. Control+EGFP vs. Bcl11b cKO+EGFP: *p=0.024, and vs. Bcl11b cKO+EGFP-2A-C1ql3: *p=0.045; Bcl11b cKO+EGFP-2A-C1ql2 vs. Bcl11b cKO+EGFP: *p=0.026, and vs. Bcl11b cKO+EGFP-2A-C1ql3: *p=0.049; ns, not significant. e Diameter of docked vesicles. Control+EGFP, Bcl11b cKO+EGFP-2A-C1ql3, N=3; Bcl11b cKO+EGFP, Bcl11b cKO+EGFP-2A-C1ql2, N=4; Two-way ANOVA. ns, not significant. f Cumulative and g relative frequency of the number of docked vesicles per 100 nm AZ profile length. h Western blot of mouse hippocampal homogenates. i Average synapse score. Control+EGFP, N=3; Control+EGFP-2A-C1ql2, N=6; Bcl11b cKO+EGFP-2A-C1ql2, N=4. Two-way ANOVA. ns, not significant. j Relative frequency of synapse scores. Data for Control+ EGFP-2A-C1ql2 from i-j in this figure are compared with Control+EGFP and Bcl11b cKO+EGFP-2A-C1ql2 data from Figure 2.

To analyze the C1ql2-dependent functions of Bcl11b on SV distribution in more detail, we quantified the number of SV docked on AZ in control animals, Bcl11b mutants, and upon the reintroduction of C1ql2. SV with a ≤ 5 nm distance from the plasma membrane were considered docked (Kusick et al., 2022; Vandael et al., 2020). Bcl11b mutant animals had significantly fewer docked vesicles per 100 nm of AZ profile length compared to control animals and more AZ with no docked vesicles at all. Rescue of C1ql2 expression restored the number of docked SV to control levels, while the overexpression of C1ql3 did not affect the number of docked vesicles (Figure 2d, f-g; Control+EGFP: 0.53±0.098, Bcl11b cKO+EGFP: 0.24±0.038, Bcl11b cKO+EGFP-2A-C1ql2: 0.51±0.049, Bcl11b cKO+EGFP-2A-C1ql3: 0.26±0.041, mean±SEM). The AZ length remained unchanged in all conditions (Supplementary Figure 2c; Control+EGFP: 168.3±4,94, Bcl11b cKO+EGFP: 161.9±5.56, Bcl11b cKO+EGFP-2A-C1ql2: 161.91±7.14, Bcl11b cKO+EGFP-2A-C1ql3: 171.87±6.74, mean±SEM). In contrast to vesicle docking, Bcl11b and C1ql2 did not affect the size of the docked SV (Figure 2e; Control+EGFP: 36.30±1.67, Bcl11b cKO+EGFP: 35.18±1.13, Bcl11b cKO+EGFP-2A-C1ql2: 36.35±1.01, Bcl11b cKO+EGFP-2A-C1ql3: 36.65±0.1, mean±SEM). Thus, our data suggest that C1ql2 specifically controls SV recruitment downstream of Bcl11b.

Conditional deletion of Bcl11b in the adult hippocampus also leads to a loss of MFS, as well as reduced ultrastructural complexity of the remaining mossy fiber boutons (MFB) (De Bruyckere et al., 2018), posing the question of whether these phenotypic features also depend on C1ql2. Interestingly, C1ql2 reintroduction in the DGN of Bcl11b cKO neither restored the loss of glutamatergic synapses, as quantified by the colocalization of pre– and postsynaptic markers, vGlut1 and Homer1 (Supplementary figure 2d; Control+EGFP: 90.65±8.25, Bcl11b cKO+EGFP: 60.68±4.62, Bcl11b cKO+EGFP-2A-C1ql2: 56.84±6.99, mean±SEM), nor the reduced ultrastructural complexity of MFB, as quantified by the MFB perimeter/area ratio (Supplementary figure 2e-f; Control+EGFP: 0.0051±0.00031, Bcl11b cKO+EGFP: 0.0042±0.00014, Bcl11b cKO+EGFP-2A-C1ql2: 0.0037±0.00021, mean±SEM). This suggests that Bcl11b acts on MFS through C1ql2-dependent as well as –independent signaling pathways.

Reintroduction of C1ql2 into Bcl11b mutant dentate granule neurons rescues mossy fiber synapse long-term potentiation

The ultrastructural changes at the MFS point towards potential alterations in synaptic function. Indeed, adult-induced Bcl11b cKO was previously found to result in a loss of MF-LTP (De Bruyckere et al., 2018). We therefore tested whether the reintroduction of C1ql2 in Bcl11b cKO DGN can rescue LTP at the mutant MFS, similarly to SV recruitment. We stimulated mossy fibers in acute slices and measured the resulting field potentials in the SL of CA3. Field responses were carefully validated for the specificity of mossy fiber signals by the presence of strong paired-pulse facilitation and block by the mGluR antagonist DCG-IV. Under these conditions, input-output curves of fEPSP slopes versus axonal fiber volleys revealed no significant differences between control and Bcl11b cKO mice (Supplementary figure 3a-b; Control+EGFP: 1.95±0.09, Bcl11b cKO+EGFP: 2.02±0.14, mean±SEM), indicating that basal synaptic transmission was unaltered in the Bcl11b mutants. We then induced LTP by high-frequency stimulation (HFS) of mossy fibers in control and Bcl11b cKO animals with or without AAV-mediated expression of C1ql2. Compared to controls, Bcl11b mutants injected with the control AAV displayed a strong reduction of LTP at 20-30 and 30-40 min after induction (Figure 3a-c; 0-10 min: Control+EGFP: 90.4±7.2, Bcl11b cKO+EGFP: 106.1±10.8, 10-20 min: Control+EGFP: 42.7±3.6, Bcl11b cKO+EGFP: 39.5±4.6, 20-30 min: Control+EGFP: 52.5±7.6, Bcl11b cKO+EGFP: 24.8±3.2, 30-40 min: Control+EGFP: 50.1±7.3, Bcl11b cKO+EGFP: 20.3±3.7, mean±SEM), consistent with our previous data (De Bruyckere et al., 2018). The loss of LTP was completely reversed upon the re-expression of C1ql2, with the rescue mice exhibiting comparable LTP to controls at all time intervals (Figure 3a-c; 0-10 min: Control+EGFP: 90.4±7.2, Bcl11b cKO+EGFP-2A-C1ql2: 86.5±7.4, 10-20 min: Control+EGFP: 42.7±3.6, Bcl11b cKO+EGFP-2A-C1ql2: 49.4±5.9, 20-30 min: Control+EGFP: 52.5±7.6, Bcl11b cKO+EGFP-2A-C1ql2: 47.2±5.7, 30-40 min: Control+EGFP: 50.1±7.3, Bcl11b cKO+EGFP-2A-C1ql2: 44.9±5.3, mean±SEM). Importantly, this rescue effect was specific to C1ql2 as the overexpression of C1ql3 failed to reverse the Bcl11b cKO phenotype (Figure 3a-c; 0-10 min: Control+EGFP: 90.4±7.2, Bcl11b cKO+EGFP-2A-C1ql3: 104.2±9.9, 10-20 min: Control+EGFP: 42.7±3.6, Bcl11b cKO+EGFP-2A-C1ql3: 44.4±5.7, 20-30 min: Control+EGFP: 52.5±7.6, Bcl11b cKO+EGFP-2A-C1ql3: 29.0±2.3, 30-40 min: Control+EGFP: 50.1±7.3, Bcl11b cKO+EGFP-2A-C1ql3: 22.6±2.4, mean±SEM).

C1ql2 reintroduction in Bcl11b cKO DGN rescues mossy fiber LTP.

a Representative fEPSP traces showing baselines before HFS (black), fEPSP changes 30-40 min after HFS (cyan) and following the application of 3 µM DCG-IV (red). b Time course of fEPSP slopes. The black arrow indicates HFS and the dashed line is the baseline level. c Quantification of fEPSP facilitation at four different time intervals after HFS. Changes in the fEPSP slope are shown as the percentage of the mean baseline fEPSP. Control+EGFP, 7 slices from 6 mice; Bcl11b cKO+EGFP, 8 slices from 5 mice, Bcl11b cKO+EGFP-2A-C1ql3, 8 slices from 6 mice; Bcl11b cKO+EGFP-2A-C1ql2, 6 slices from 4 mice; All data are presented as means; error bars indicate SEM. One-way ANOVA followed by Bonferroni’s PHC for each time interval. 20-30 min: Control+EGFP vs. Bcl11b cKO+EGFP: **p=0.002, and vs. Bcl11b cKO+EGFP-2A-C1ql3: *p=0.011; Bcl11b cKO+EGFP-2A-C1ql2 vs. Bcl11b cKO+EGFP: *p=0.023; 30-40 min: Control+EGFP vs. Bcl11b cKO+EGFP: ***p<0.001, and vs. Bcl11b cKO+EGFP-2A-C1ql3: **p=0.002; Bcl11b cKO+EGFP-2A-C1ql2 vs. Bcl11b cKO+EGFP: *p=0.01 and vs. Bcl11b cKO+EGFP-2A-C1ql3: *p=0.0523; ns, not significant. d Representative fEPSP traces showing baselines before forskolin application (black), fEPSP changes 105-120 min after the start of application (cyan) and following the addition of 3 µM DCG-IV (red). e Time course of fEPSP slopes. The black solid line indicates forskolin perfusion and the dashed line is the baseline level. f Quantification of fEPSP facilitation at two different time intervals after the start of the forskolin application. Changes in fEPSP slope are shown as percentage of the mean baseline fEPSP. 8 slices from 5 mice. All data are presented as means; error bars indicate SEM. Unpaired t-test for both time intervals. 15-30 min: **p=0.005; 105-120 min: **p=0.0025.

MF-LTP is known to be mediated by the second messenger cAMP, which is produced by adenylyl cyclase (AC) in response to Ca2+ influx through voltage-gated Ca2+ channels (Li et al., 2007) and kainate receptors (KAR) (Lauri et al., 2001; Schmitz et al., 2003). To test whether Bcl11b acts on LTP by interfering with presynaptic Ca2+ dynamics, we directly activated the cAMP pathway in slices from control and Bcl11b cKO mice by applying the AC activator forskolin (Weisskopf et al., 1994). Compared to slices from control animals, forskolin-induced LTP in the mutants had a significantly lower peak and remained significantly weaker throughout the recording (Figure 3d-f; 15-30 min: Control: 342.4±28.0, Bcl11b cKO: 232.3±17.7, 105-120 min: Control: 88.0±7.4, Bcl11b cKO: 47.2±8.2, mean±SEM). This suggests that the regulation of MF-LTP by Bcl11b involves the cAMP-dependent signaling pathway.

Knock-down of C1ql2 in dentate granule neurons perturbs synaptic vesicle recruitment and long-term potentiation at the mossy fiber-CA3 synapse

To further corroborate the observation that Bcl11b acts on MFS specifically through C1ql2, we knocked down C1ql2 expression in the DGN of adult WT mice by stereotaxically injecting an AAV carrying an shRNA cassette against C1ql2 (Figure 4a). Quantitative PCR (Figure 4b), western blot analysis (Figure 4c), as well as immunohistochemistry using C1ql2 antibodies on hippocampal tissue (Figure 4d), revealed that the shRNA-mediated KD resulted in a strong reduction of C1ql2 transcripts (Figure 4b; +shNS-EGFP: 1±0.07, +shC1ql2-EGFP: 0.23±0.059, mean±SEM) as well as protein levels (Figure 4c-d), as compared to animals injected with the control AAV. The shRNA-mediated KD of C1ql2 did not affect the expression of C1ql3, demonstrating the specificity of this approach (Supplementary figure 4a-b; +shNS-EGFP: 1±0.09, +shC1ql2-EGFP: 0.986±0.035, mean±SEM). Compared to controls, C1ql2 KD was sufficient to reduce the average synapse score to similar levels as observed in Bcl11b cKO (Figure 4e, Supplementary figure 4c; +shNS-EGFP: 3.38±0.069, +shC1ql2-EGFP: 3.15±0.031, mean±SEM), as well as the number of docked vesicles per 100 nm of AZ profile length (Figure 4f, Supplementary figure 4d-e; +shNS-EGFP: 0.48±0.04, +shC1ql2-EGFP: 0.31±0.02, mean±SEM). At the same time, the length of the AZ and the diameter of the docked vesicles remained unchanged (Supplementary figure 4f-g; AZ length: +shNS-EGFP: 172.96±8.24, +shC1ql2-EGFP: 182.52±4.8, mean±SEM; Vesicle diameter: +shNS-EGFP: 34.28±0.84, +shC1ql2-EGFP: 35.37±0.21, mean±SEM). Moreover, C1ql2 KD did not affect the number of MFB, as quantified by the number of ZnT3+ puncta in the SL of CA3 (Supplementary figure 4h-i; +shNS-EGFP: 1525.319±90.72, +shC1ql2-EGFP: 1547.94±48.51, mean±SEM). To test whether shRNA-mediated KD of C1ql2 expression also affects MF-LTP, we performed LTP recordings in C1ql2 KD and control mice. Compared to controls, slices from C1ql2 KD mice exhibited a significant reduction of LTP at 20-30 and 30-40 min time intervals, similarly to Bcl11b cKO animals (Figure 4g-i; 0-10 min: +shNS-EGFP: 105.0±4.0, +shC1ql2-EGFP: 94.3±4.5, 10-20 min: +shNS-EGFP: 56.3±4.5, +shC1ql2-EGFP: 35.1±2.8, 20-30 min: +shNS-EGFP: 50.2±4.5, +shC1ql2-EGFP: 23.4±3.5, 30-40 min: +shNS-EGFP: 44.6±4.3, +shC1ql2-EGFP: 20.1±4.1, mean±SEM). C1ql2 KD did not affect basal synaptic transmission, as evidenced by the respective input-output curves (Supplementary figure 3c-d; +shNS-EGFP: 2.50±0.16, +shC1ql2-EGFP: 2.45±0.14, mean±SEM). Thus, KD of C1ql2 in WT DGN recapitulates major phenotypes observed upon Bcl11b cKO, supporting that Bcl11b controls SV recruitment and LTP in hippocampal MFS specifically through C1ql2.

KD of C1ql2 in DGN of WT mice impairs SV recruitment and LTP.

a Experimental design to analyze the MFS after AAV-mediated KD of C1ql2 in WT DGN. b Relative C1ql2 mRNA levels in DGN. N=4. All data are presented as means; error bars indicate SEM. Unpaired t-test: ***p=0.0002. c Western blot of mouse hippocampal homogenates. d Immunohistochemistry of C1ql2 (red) and GFP (green) on hippocampal sections. Scale bar: 200 μm. e Average synapse score. N=4. Unpaired t-test. *p=0.025. f Number of docked vesicles per 100 nm AZ profile length. N=3. All data are presented as means; error bars indicate SEM. Points represent the individual examined AZ. Unpaired t-test. *p=0.018. g Representative fEPSP traces showing baselines before HFS (black), fEPSP changes 30-40 min after HFS (cyan) and following the application of 3 µM DCG-IV (red). h Time course of fEPSP slopes. The black arrow indicates HFS and the dashed line the baseline level. i Quantification of fEPSP facilitation at four different time intervals after HFS. Changes in fEPSP slope are shown as percentage of the mean baseline fEPSP. +shNS-EGFP, 6 slices from 6 mice; +shC1ql2-EGFP, 7 slices from 7 mice. All data are presented as means; error bars indicate SEM. Mann-Whitney U-test for each time interval. 10-20 min: **p=0.0012; 20-30 min: **p=0.0023; 30-40 min: **p=0.0023; ns, not significant.

C1ql2-Nrxn3(25b+) interaction recruits presynaptic vesicles in vitro and in vivo

C1ql2 was previously shown to interact with a particular splice variant of Nrxn3β containing exon 25b sequences, Nrxn3(25b+), which was recombinantly expressed in HEK293 cells (Matsuda et al., 2016). To explore whether the Bcl11b/C1ql2-dependent regulation of MFS involves interaction with neuronal Nrxn3, we co-cultured HEK293 cells that secreted myc-tagged C1ql2, to create regions of highly concentrated C1ql2, with primary hippocampal neurons transfected with GFP-Nrxn3α(25b). We used here the extracellularly longer Nrxn3α isoform because it is more strongly expressed in the murine DG compared to Nrxn3β (Uchigashima et al., 2019). C1ql2-dependent recruitment of Nrxn3α(25b+) was quantified by the surface area of HEK293 cells covered by GFP-Nrxn3α(25b+)-positive neuronal profiles (Figure 5a). HEK293 cells secreting C1ql2 had a significantly larger surface area covered by neuronal Nrxn3α(25b+) in comparison to HEK293 cells secreting myc-tag only (Figure 5b; myc-C1ql2: 39.97±3.99, myc-tag: 17.29±2.27, mean±SEM). Using vGlut1 immunoreactivity as a proxy for SV localization (Aoto et al., 2007; Fremeau et al., 2004), we examined in the same system whether C1ql2-secreting HEK293 cells were able to cluster vGlut1 in surrounding GFP-Nrxn3α(25b+)-positive neurons (Figure 5c). Interestingly, vGlut1 accumulation was significantly increased in GFP-Nrxn3α(25b+)-positive neurons contacting C1ql2-secreting HEK293 cells as compared to C1ql2-negative HEK293 cells (Figure 5d; myc-C1ql2: 40.88±3.25, myc-tag: 24.78±4.99, mean±SEM). To specifically analyze whether endogenous Nrxn3 is required for C1ql2-mediated vGlut1 accumulation in neurons, we co-cultured C1ql2-secreting HEK293 cells with primary hippocampal neurons derived from Nrxn123flox/flox mice, in which all three Nrxn genes are floxed (Chen et al., 2017), and which we transfected with either active Cre recombinase or an inactive Cre (Klatt et al., 2021) (Figure 5e). Neurons lacking all Nrxn variants showed a significantly lower accumulation of endogenous vGlut1 when contacting the C1ql2-secreting HEK293 cells compared to control neurons. Strikingly, selective reintroduction of the Nrxn3α(25b+) isoform into the null-mutant Nrxn cKO neurons was sufficient to normalize vGlut1 accumulation in vitro (Figure 5f; inactive Cre: 51.66±5.97, Cre: 27.83±2.83, Cre+ Nrxn3α(25b+): 39.23±4.3, mean±SEM). Collectively, our data strongly suggest that the C1ql2-mediated recruitment of vGlut1-positive SV in hippocampal neurons depends on the presence of Nrxn3α(25b+).

C1ql2-Nrxn3 interaction recruits vGlut1 in vitro.

a Immunocytochemistry of myc-tagged C1ql2, C1ql2.K262E or myc-tag (magenta) expressing HEK293 cells and GFP-Nrxn3α(25b+) (cyan) from contacting hippocampal neurons. Scale bar: 5 μm. b Nrxn3α(25b+) recruitment by differentially transfected HEK293 cells. N=3. All data are presented as means; error bars indicate SEM. One-way ANOVA and Tuckey’s PHC. myc-C1ql2 vs. myc-tag: *p=0.016, and vs. myc-K262E: *p=0.022; ns, not significant. c Immunocytochemistry of myc-tagged C1ql2, C1ql2.K262E or myc-tag (magenta) expressing HEK293 cells and vGlut1 (cyan) from contacting hippocampal neurons. Scale bar: 5 μm. d vGlut1 recruitment by differentially transfected HEK293 cells. N = 3. All data are presented as means; error bars indicate SEM. One-way ANOVA and Tuckey’s PHC. myc-C1ql2 vs. myc-tag: *p=0.04, and vs. myc-K262E: **p=0.007; ns, not significant. e Immunocytochemistry of myc-tagged C1ql2 (magenta) expressing HEK293 cells and vGlut1 (cyan) from contacting control, Nrxn123 KO or Nrxn123 KO with Nrxn3α(25+) rescued hippocampal neurons. Scale bar: 5 μm. f vGlut1 recruitment by HEK293 cells in presence or absence of neuronal Nrxns. N = 3. All data are presented as means; error bars indicate SEM. One-way ANOVA and Tuckey’s PHC. inactive Cre vs. Cre: *p=0.023, and vs. Cre+ Nrxn3α(25+): p=0.21; ns, not significant. g Trimeric structures of C1ql2 (PDB_ID: 4QPY, upper panels) and the variant C1ql2.K262E (lower panels). Residue 262 is the central residue (red, left & middle panels) of a larger area underneath the C1ql2-specific calcium and receptor binding loops (magenta, middle panel). The mutation K262E alters the charge of that surface area negative (yellow-circled area, right panels) and makes it potentially repulsive to bind Nrxn3(25b+).

To explore the relevant epitope that mediates the binding of C1ql2 to Nrxn3(25b+) proteins, we analyzed the solvent accessible electrostatic surface properties of the C1ql-domain trimeric structure of C1ql2 (Ressl et al., 2015) (PDB_ID: 4QPY) and found that a change of lysine262 (K262) to glutamic acid renders a large area underneath the C1ql2-specific calcium and receptor binding loops negative (Figure 5g) and hypothesized that this would repel binding to Nrxn3(25b+). We generated a C1ql2.K262E variant, expressed it in HEK293 cells as before, and tested it for its ability to cluster Nrxn3α(25b+) as well as vGlut1 in contacting primary neurons (Figure 5a-d). In the presence of C1ql2.K262E, recruitment of Nrxn3α(25b+) was significantly lower compared to WT C1ql2 and indistinguishable from myc-tag control levels. (Figure 5a-b; myc-K262E: 18.84±5.15). Moreover, the expression of C1ql2.K262E in HEK293 cells was unable to accumulate vGlut1 in contacting neurons expressing GFP-Nrxn3α(25b+) (Figure 5c-d; myc-K262E: 16.9±1.2, mean±SEM). Together, these results provide in vitro evidence that the clustering of vGlut1 depends on an intact C1ql2-Nrxn3(25b+) interaction and that a single point mutation that creates a negative charge of that surface area underneath the C1ql2-specific calcium and receptor binding loops abolishes this binding activity and, thereby, the regulation of SV clustering.

To validate our identification of K262 as a key residue for the C1ql2-Nrxn3(25b+) interaction in vivo, we expressed C1ql2.K262E in Bcl11b cKO DGN, in which endogenous C1ql2 expression is downregulated by the ablation of Bcl11b (Figure 6a), while Nrxn3 mRNA levels remain unaltered (Supplementary figure 5a; Control+EGFP: 1±0.173, Bcl11b cKO+EGFP: 1.14±0.27, mean±SEM). AAV-mediated introduction of C1ql2.K262E in Bcl11b cKO DGN resulted in strong overall expression of the mutant protein (Figure 6b-c; Control+EGFP: 1±0.42, Bcl11b cKO+EGFP-2A-K262E: 9.68±4.75, mean±SEM). However, the spatial distribution of C1ql2.K262E was notably different from the WT protein in the SL of CA3 where most of the MFS are located (Figure 6d). In the SL of CA3, protein levels of C1ql2.K262E were significantly lower compared to WT C1ql2 as quantified by the integrated fluorescence density (Figure 6d-e; Bcl11b cKO+EGFP-2A-C1ql2: 9.75±0.57 x104, Bcl11b cKO+EGFP-2A-K262E: 5.89±0.55 x104, mean±SEM). The remaining signal of the C1ql2.K262E at the SL was equally distributed and in a punctate form, similar to WT C1ql2. As C1ql3 has been shown to form functional heteromers with C1ql2 at the MFS, we examined the spatial distribution of the C1ql3 protein upon AAV-mediated introduction of C1ql2.K262E for potential expression pattern changes but observed no overt difference (Supplementary figure 5b).

C1ql2-Nrxn3(25b+) interaction is important for C1ql2 localization at the MFS and SV recruitment.

a Experimental design to analyze the MFS after AAV-mediated expression of C1ql2.K262E in Bcl11b cKO DGN. b Western blot and c relative C1ql2.K262E protein levels in mouse hippocampal homogenates. N=3. All data are presented as means; error bars indicate SEM. Mann-Whitney U-test. ns, not significant. d Immunohistochemistry of C1ql2 (red) and GFP (green) in hippocampal sections. Scale bar: 200 μm. Upper panels depict close-ups of C1ql2 staining from the SL of CA3. Scale bar: 15 μm. e Integrated density of C1ql2 fluorescence in the SL of CA3. N=3. All data are presented as means; error bars indicate SEM. Unpaired t-test. *p=0.008. f Electron microscope images of MFS and proximal SV. White bars mark synapse length from postsynaptic side. Scale bar: 100 nm. g Average synapse score. Control+EGFP, Bcl11b cKO+EGFP-2A-K262E: N=3; Bcl11b cKO+EGFP: N=4. Two-way ANOVA and Tuckey’s PHC. Control+EGFP vs. Bcl11b cKO+EGFP: ***p=0.0004, and vs. Bcl11b cKO+EGFP-2A-K262E: ***p=0.0002; ns, not significant. h Number of docked vesicles per 100 nm AZ profile length. Control+EGFP, Bcl11b cKO+EGFP-2A-K262E: N=3; Bcl11b cKO+EGFP: N=4. All data are presented as means; error bars indicate SEM. Points represent the individual examined AZ. Two-way ANOVA and Tuckey’s PHC. Control+EGFP vs. Bcl11b cKO+EGFP: *p=0.0434, and vs. Bcl11b cKO+EGFP-2A-K262E: *p=0.0196; ns, not significant. Data for Control+EGFP and Bcl11b cKO+EGFP-2A-K262E from f-h in this figure are compared with Bcl11b cKO+EGFP data from Figure 2.

To investigate whether the remaining C1ql2.K262E affected the SV recruitment, we determined if C1ql2.K262E was able to recover the average synapse score in the Bcl11b cKO background, and found that the mutant C1ql2 variant did not rescue the SV distribution (Figure 6f-g, Supplementary figure 5c; Control+EGFP: 3.44±0.012, Bcl11b cKO+EGFP: 2.96±0.037, Bcl11b cKO+EGFP-2A-K262E: 2.87±0.043, mean±SEM). Furthermore, the number of docked vesicles per 100 nm of AZ profile length in the MFB of animals receiving the C1ql2.K262E AAV was significantly lower compared to control animals and similar to that of Bcl11b mutants (Figure 6h, Supplementary figure 5f-g; Control+EGFP: 0.41±0.049, Bcl11b cKO+EGFP: 0.24±0.038, Bcl11b cKO+EGFP-2A-K262E: 0.19±0.025, mean±SEM). The length of the AZ and the diameter of docked vesicles were unchanged (Supplementary figure 5d-e; AZ length: Control+EGFP: 167.94±7.35, Bcl11b cKO+EGFP: 161.9±5.56, Bcl11b cKO+EGFP-2A-K262E: 173.51±7.7, mean±SEM; Vesicle diameter: Control+EGFP: 39.06±1.22, Bcl11b cKO+EGFP: 35.18±1.13, Bcl11b cKO+EGFP-2A-K262E: 36.38±2.19, mean±SEM). Unexpectedly, however, C1ql2.K262E was able to rescue the loss of MF-LTP observed in Bcl11b cKO (Supplementary figure 5h-j; 0-10 min: Control+EGFP: 90.4±7.2, Bcl11b cKO+EGFP-2A-K262E: 155.8±30.3, 10-20 min: Control+EGFP: 42.7±3.6, Bcl11b cKO+EGFP-2A-K262E: 68.7±17.3, 20-30 min: Control+EGFP: 52.5±7.6, Bcl11b cKO+EGFP-2A-K262E: 55.9±13.8, 30-40 min: Control+EGFP: 50.1±7.3, Bcl11b cKO+EGFP-2A-K262E: 47.8±9.3, mean±SEM). It has been shown that C1ql2 also interacts with specific postsynaptic KAR subunits (Matsuda et al., 2016). To test whether the C1ql2.K262E variant retained its ability to interact with GluK2, protein extract of HEK293 cells expressing either GluK2-myc-flag/GFP-C1ql2 or GluK2-myc-flag/GFP-C1ql2.K262E was examined by co-immunoprecipitation and revealed that both C1ql2 and C1ql2.K262E had GluK2 bound when precipitated (Supplementary figure 5k). Together, our data suggest that Bcl11b regulates MFS function through divergent C1ql2-dependent downstream signaling pathways: while SV recruitment depends on a direct interaction of C1ql2 with Nrxn3(25b+), C1ql2 appears to regulate MF-LTP through Nrxn3(25b+)-independent mechanisms.

To further explore whether binding to Nrxn3 is required for C1ql2-dependent regulation of SV recruitment, we stereotaxically injected an AAV expressing GFP-tagged Cre or inactive Cre into the DG of 2-month-old Nrxn123flox/flox mice (Figure 7a), which resulted in strong reduction of Nrxn3 mRNA levels in DGN 2 months later. Only mild reduction of Nrxn1 and unchanged expression of Nrxn2 was observed (Figure 7b, Supplementary figure 6a; Nrxn1: +inactive Cre: 1±0.084, +Cre: 0.714±0.037; Nrxn2: +inactive Cre: 1±0.065, +Cre: 0.771±0.071; Nrxn3: +inactive Cre: 1±0.127, +Cre: 0.381±0.09, mean±SEM). Fluorescence intensity of endogenous C1ql2 protein along the MF axons in the SL of CA3 was significantly reduced in Nrxn123 cKO animals compared to Nrxn123flox/flox animals expressing inactive Cre (Figure 7c-d; +inactive Cre: 11.72±1.63, +Cre: 4.71±0.93, mean±SEM). However, C1ql2 mRNA levels in DGN remained unchanged (Supplementary figure 6b; +inactive Cre: 1±0.23, +Cre: 0.8±0.11, mean±SEM), suggesting that overall production of C1ql2 protein was not affected. To control for the specificity of this effect, we also determined the level of C1ql3 expression and found no overt changes in Nrxn123 cKO (Supplementary figure 6c). To exclude that the reduced C1ql2 fluorescence intensity was simply a consequence of an overall loss of MFB, we used ZnT3 as a marker of MFB and found it unchanged in Nrxn123 cKO compared to controls (Figure 7e). Remarkably, disruption of the C1ql2-Nrxn3(25b+) binding by ablation of Nrxn3 in Nrxn123 cKO mutants not only led to reduced C1ql2 fluorescence intensity (Figure 7c-d), but recapitulated the phenotype observed upon Bcl11b ablation or by KD of C1ql2 as evidenced by a large reduction of the average synapse score in Nrxn123 cKO (Figure 7f-g, Supplementary figure 6d; +inactive Cre: 3.11±0.06; +Cre: 2.67±0.074, mean±SEM). Also, similarly to the Bcl11b and C1ql2 mutant phenotypes, we observed the number of docked vesicles per 100 nm of AZ profile length in Nrxn123 cKO to be diminished compared to controls (Figure 7h, Supplementary figure 6e-f; +inactive Cre: 0.404±0.035, +Cre: 0.195±0.02, mean±SEM), whereas the AZ length and the diameter of docked vesicles remained unchanged (Supplementary figure 6g-h; AZ length: +inactive Cre: 193.2±6.88; +Cre: 188.44±11.43, mean±SEM; Vesicle diameter: +inactive Cre: 38.38±0.44; +Cre: 37.12±0.8, mean±SEM). Thus, our results provide evidence that Bcl11b controls MFS organization through C1ql2/Nrxn3(25b+)-dependent signaling, explicating how Bcl11b, a transcription factor with a broad range of functions, can regulate highly specific processes in the brain.

Nrxn KO perturbs C1ql2 localization at the MFS and SV recruitment.

a Experimental design to analyze the MFS after AAV-mediated Nrxn KO. b Relative Nrxn3 mRNA levels. N=4. All data are presented as means; error bars indicate SEM. Unpaired t-test. **p=0.007. c Immunohistochemistry of C1ql2 (red) and GFP (green) in hippocampal sections. Scale bar: 200 μm. Upper panels depict close-ups of C1ql2 staining from the SL of CA3. Scale bar: 15 μm. d Integrated density of C1ql2 fluorescence in the SL of CA3. N=3. All data are presented as means; error bars indicate SEM. Unpaired t-test. *p=0.02. e Immunohistochemistry of ZnT3 (red) and GFP (green) in hippocampal sections. Scale bar: 200 μm. Upper right corner of ZnT3 panels depicts close-ups from the SL of CA3. Scale bar: 15 μm. f Electron microscope images of MFS and proximal SVs. White bars mark synapse length from postsynaptic side. Scale bar: 100 nm. g Average synapse score. N=3. Unpaired t-test. **p=0.009. h Number of docked vesicles per 100 nm AZ profile length. N=3. All data are presented as means; error bars indicate SEM. Points represent the individual examined AZ. Unpaired t-test. **p=0.007.

Discussion

There is emerging evidence that the zinc finger transcription factor Bcl11b is involved in the pathogenesis of neurodevelopmental as well as neuropsychiatric disorders that are frequently associated with synaptic dysfunction. Previous work from our group demonstrated Bcl11b to be essential for synapse function in the mossy fiber circuit of the adult murine hippocampus. The underlying molecular mechanisms downstream of Bcl11b, however, remained elusive. In the present study, we uncover a novel C1ql2-dependent regulatory pathway through which Bcl11b controls the structural as well as functional integrity of hippocampal MFS in adult mice. We show that SV recruitment to the AZ of MFS, as well as the expression of MF-LTP, depend on C1ql2, which is a direct functional target of Bcl11b. Reintroduction of C1ql2 into Bcl11b mutant DGN restores defective SV recruitment and LTP expression. KD of C1ql2 in DGN recapitulates the impaired SV recruitment and loss of LTP observed in Bcl11b mutants. Finally, we show that C1ql2 controls SV recruitment through a direct interaction with presynaptic Nrxn3(25b+), while LTP depends on C1ql2 signals independent of Nrxn3 interaction. Recent studies suggested Nrxn3, as well as C1ql2, to be associated with neuropsychiatric disorders (Hishimoto et al., 2007; Hu et al., 2013; Huggett & Stallings, 2020a, 2020b; Marballi et al., 2022). Our study for the first time identifies a Bcl11b/C1ql2/Nrxn3-dependent signaling pathway in the control of basic structural and functional properties of MFS. Analysis of this regulatory pathway in mice may provide important novel insights into the pathogenesis of neurodevelopmental and neuropsychiatric disorders.

We have previously shown that conditional ablation of Bcl11b in the adult hippocampus leads to structural and functional changes of MFS characterized by an overall reduction in synapse numbers, loss of bouton complexity, misdistribution of SV as well as loss of MF-LTP (De Bruyckere et al., 2018). Here, we found that reintroduction of the synaptic organizer protein C1ql2, which is a direct transcriptional target of Bcl11b and is downregulated in Bcl11b mutant DGN (De Bruyckere et al., 2018), was able to rescue major part of the Bcl11b mutant phenotype at the MFS. Restoring C1ql2 expression in Bcl11b cKO DGN led to a complete rescue of the SV distribution and docking, as well as LTP at the MFS, while synapse numbers and ultrastructural complexity of boutons remained unchanged.

Furthermore, KD of C1ql2 in WT DGN recapitulated the Bcl11b phenotype with impaired SV recruitment and loss of LTP, supporting the specificity of C1ql2 function. MF-LTP, which manifests as a long-term increase in presynaptic vesicle release probability (Pr) (Shahoha et al., 2022), directly depends on the distribution of SV in the proximity of the AZ. Recent studies have shown that the increase in Pr involves the recruitment of new AZ and an increase in the number of docked and tethered vesicles, corresponding to the readily releasable pool of SV (Orlando et al., 2021; Vandael et al., 2020). The perturbed SV recruitment in both Bcl11b cKO and C1ql2 KD mice could thus potentially explain the loss of LTP in both conditions. Indeed, the reintroduction of C1ql2 in Bcl11b cKO DGN specifically rescued SV recruitment and LTP, while synapse numbers and ultrastructural complexity of MFB remained unchanged. Whether the regulation of SV distribution by C1ql2 and the expression of MF-LTP are directly and causally linked remains to be determined. Additional factors have been suggested to contribute to the increase in Pr, including a tighter coupling between Ca2+ channels and SV (Midorikawa & Sakaba, 2017) and the accumulation of Ca2+ channels near release sites (Fukaya et al., 2021). It cannot, therefore, be excluded that C1ql2 regulates MF-LTP through one of these alternative mechanisms. Aiming to narrow in on the nature of the mechanism through which Bcl11b regulates MF-LTP, we used forskolin to induce LTP in Bcl11b cKO. MF-LTP relies on presynaptic mechanisms (Castillo, 2012; Zalutsky & Nicoll, 1990) and is mediated by the second messenger cAMP, which is produced by AC in response to Ca2+ influx through voltage-gated Ca2+ channels (Li et al., 2007) and KARs (Lauri et al., 2001; Schmitz et al., 2003). By using forskolin to directly activate AC we bypassed these initial steps, and still found a reduction of LTP in Bcl11b cKO mice, similarly to HFS. These results strongly suggest that the loss of LTP is caused by a process downstream of the initial presynaptic Ca2+ influx following stimulation. We note that, in the present experiments, we did not observe the decrease in input-output relation in Bcl11b cKO as reported in De Bruyckere et al., 2018. After excluding technical differences, e.g., different methods of data analysis, we conclude that the discrepancy is best explained by differences in the population of presynaptic fibers. In the present study, mossy fiber responses were specifically identified by testing for frequency facilitation and sensitivity to mGluR antagonists, whereas in the previous study, this purification was not done (De Bruyckere et al., 2018). It is not immediately obvious why the reduction in synapse numbers and misdistribution of SV in Bcl11b cKO animals does not affect basal synaptic transmission. While a modest displacement of SV might fail to noticeably influence synaptic transmission due to the low initial Pr at MFS, causing only a fraction of release-ready vesicles to be initially released, the reduction in synapse numbers might indeed be expected to reflect in the input-output relationship. It might be that synapses that are preferentially eliminated in Bcl11b mutants are predominantly silent or have weak coupling strength, such that their loss has only a minimal effect on synaptic transmission. Further investigation is needed to elucidate this apparent discrepancy. Together, our results suggest Bcl11b to be an important synaptic regulator that controls the structure and function of adult MFS through both C1ql2-dependent, as well as –independent transcriptional programs.

C1ql proteins are complement-related factors that are synthesized by the presynapse and secreted into the synaptic cleft. Within the hippocampus, C1ql2 and –3 protein expression overlaps and is highly restricted to DGN (Iijima et al., 2010) and the corresponding mossy fiber system, including MF-CA3 synapses. C1ql2 and –3 were previously suggested to form functional heteromers that can cluster postsynaptic KAR on MFS. Selective deletion of either C1ql2 or –3 in mice was reported to have no overt mutant hippocampal phenotype, suggesting functional compensation for both proteins (Matsuda et al., 2016). Using shRNA-mediated selective KD of C1ql2 in DGN as well as rescue of the Bcl11b mutation by the reintroduction of C1ql2 into mutant DGN, we observed a novel, presynaptic function for C1ql2 in the recruitment of SV and the expression of LTP in MFS. This function was specific to C1ql2 since overexpression of C1ql3 in Bcl11b mutant DGN was unable to rescue the synapse phenotype. Furthermore, another study has identified all four C1ql proteins, including C1ql2, as ligands for the postsynaptic Brain-specific angiogenesis inhibitor 3 (BAI3). Addition of any of the four C1ql proteins to cultured hippocampal neurons led to a loss of excitatory synapses, a function inhibited by the presence of BAI3 (Bolliger et al., 2011). In our study, we show that neither loss of C1ql2 nor overexpression of C1ql2 affects the number of MFS, supporting the notion that synaptic organizers have synapse-specific functions. This highlights the role of C1ql2 as a synaptic organizer and adds a new layer of understanding to its function at the MFS.

Previous in vitro studies suggested that C1ql2 function at the MFS involves interaction with Nrxn3 isoforms containing the splice site 5 25b sequence (SS525b) (Matsuda et al., 2016). Nrxns are synaptic cell adhesion molecules that mediate various synaptic properties (Reissner et al., 2013; Sudhof, 2017), including the recruitment of SV and dense-core vesicles (Dean et al., 2003; Ferdos et al., 2021; Quinn et al., 2017; Rui et al., 2017). This prompted us to analyze, whether C1ql2-dependent SV recruitment in MFS requires a direct interaction with Nrxn3(25b+) in vitro and in vivo. Expression of C1ql2 in HEK293 cells co-cultured with GFP-Nrxn3α(25+)-expressing hippocampal neurons was able to recruit Nrxn3α(25b+) and vGlut1 at contact points, while C1ql2.K262E, a C1ql2 variant with an amino-acid replacement that perturbs the interaction with Nrxn3(25b+), was no longer able to recruit neuronal vGlut1. Furthermore, clustering of vGlut1 by C1ql2-secreting HEK293 cells was reduced in neurons harboring a pan-neurexin mutation, a phenotype that was rescued by the selective reintroduction of Nrxn3α(25b+). Finally, the introduction of C1ql2.K262E in Bcl11b cKO DGN in vivo was unable to rescue SV recruitment, while the silencing of Nrxns in DGN in vivo perturbed SV recruitment to a similar extent as in Bcl11b cKO and C1ql2 KD. Based on these findings, we anticipated the overexpression of C1ql2.K262E in Bcl11b cKO DGN to be unable to rescue MF-LTP. Unexpectedly, the introduction of C1ql2.K262E into Bcl11b cKO fully rescued MF-LTP. This raises the possibility that C1ql2 can influence MF-LTP through additional, yet uncharacterized mechanisms, independent of SV recruitment or direct interaction with Nrxn3(25b+). We cannot exclude, however, that the expression of a mutant C1ql2 variant created an additional gain-of-function effect that circumvented SV recruitment and allowed the rescue of MF-LTP in our experimental system. The latter is supported by the fact that within the first 10 minutes after HFS, fEPSP slopes for C1ql2.K262E were significantly elevated compared to controls, an effect that was not seen after C1ql2 re-expression. Together, our data provide comprehensive experimental evidence that the direct interaction of C1ql2 with Nrxn3(25b+) is essential for SV recruitment at the MFS. Finally, we observed that an abolished interaction between C1ql2 and Nrxn3(25b+) was associated with reduced localization of the C1ql2 protein along the MF tract. This raises the possibility that C1ql2-Nrxn3 interaction might be involved in surface presentation of C1ql2, transportation, or stabilization at the MFS. The C1q domain can form stable, higher-order oligomers (Ressl et al., 2015). Neurexins, on the other hand, are highly mobile outside and inside of synaptic terminals (Klatt et al., 2021; Neupert et al., 2015). Thus, the interaction of C1ql2 with Nrxn3(25b+) may reciprocally augment the accumulation of both proteins at synaptic sites.

Neurexin mRNAs are subjected to extensive alternative splicing that leads to the expression of thousands of isoforms with differential expression patterns (Treutlein et al., 2014; Ullrich et al., 1995) that act in a type-specific manner on synaptic functions (Dai et al., 2019; Schreiner et al., 2014; Traunmuller et al., 2016). Nrxn3 splice variants have been shown to regulate the function and plasticity of glutamatergic and GABAergic synapses through various mechanisms (Aoto et al., 2013; Dai et al., 2019; Lloyd et al., 2023; Trotter et al., 2023). The Nrxn3 splice site SS5 is a major contributor to the high number of Nrxn3 isoforms (Schreiner et al., 2014). One such isoform was recently found to be highly expressed in GABAergic interneurons at the DG, where it regulates dendritic inhibition (Hauser et al., 2022). Our findings on the role of Nrxn3 isoforms containing SS525b in the recruitment of SV at the MFS through interaction with C1ql2 add to the understanding of the synapse-specific mechanisms of action of Nrxns.

Perturbations in synaptic structure and function are major determinants of various neuropsychiatric and neurodevelopmental disorders (Hayashi-Takagi, 2017; Lepeta et al., 2016; Zoghbi & Bear, 2012). Emerging evidence from recent genetic studies suggests such disorders to be linked to various genes encoding for synaptic proteins (Sudhof, 2021; Torres et al., 2017; Wang et al., 2018). Decoding the molecular mechanisms of synaptic organization and stability and their transcriptional regulation would therefore be expected to contribute to the mechanistic understanding of neuropsychiatric and neurodevelopmental disorders. The transcription factor Bcl11b has been linked to neurodevelopmental (Lessel et al., 2018), neurodegenerative (Kunkle et al., 2016; Song et al., 2022) and neuropsychiatric disorders (Whitton et al., 2018; Whitton et al., 2016). BCL11B mutations in humans are associated with neurodevelopmental delay, overall learning deficits as well as impaired speech acquisition and autistic features (Eto et al., 2022; Lessel et al., 2018; Punwani et al., 2016; Yang et al., 2020). Moreover, conditional ablation of Bcl11b selectively in the adult murine hippocampus results in impaired learning and memory behavior (Simon et al., 2016). NRXN3 single-nucleotide polymorphisms (SNP) have been implicated in schizophrenia (Hu et al., 2013) and addiction (Hishimoto et al., 2007), with one recorded SNP located close to SS5 altering the expression of Nrxn3(25b+). Interestingly, recent studies have also associated C1QL2 with schizophrenia (Marballi et al., 2022) as well as cocaine addiction (Huggett & Stallings, 2020a, 2020b). In this study we demonstrate that Bcl11b, through its transcriptional target C1ql2, modulates the synaptic organization of MFS by controlling the recruitment of SV at AZ. This regulatory mechanism depends on a direct interaction of C1ql2 with Nrxn3(25b+). Importantly, SV trafficking and altered release probability have been implicated in neurological and neuropsychiatric disorders (Egbujo et al., 2016; Lepeta et al., 2016; Zhu et al., 2021). Thus, the identification of the Bcl11b/C1ql2/Nrxn3(25b+)-dependent signaling module in this study provides a new entry point for future mechanistic analyses of synaptopathies. Moreover, the existence of such cell type-specific signaling modules reveals how a fundamental transcription factor with diverse functions such as Bcl11b can be implicated in the pathogenesis of brain disorders characterized by synaptic dysfunction.

Materials and Methods

Animals

Bcl11b inducible mutants were generated as previously described (De Bruyckere et al., 2018). Bcl11bflox/flox; CaMKIIa-CreERT2 (Bcl11b cKO) and Bcl11b+/+; CaMKIIa-CreERT2 (control) littermates were used. The Bcl11b mutation was induced by intraperitoneal injection of 2 mg tamoxifen for five consecutive days. C57BL/6JRj mice were obtained from Janvier-Labs. For the pan-neurexin KO, Nrxn123flox mice (Chen et al., 2017) were used. Animals were kept in a 12:12-h light–dark cycle at a constant temperature (22 ± 1 °C) in IVC cages. All mouse experiments were carried out in compliance with the German law and approved by the respective government offices in Tübingen and Karlsruhe, Germany.

Stereotaxic Injections

For the expression of C1ql2 and C1ql3, the DG of 80 days old Bcl11b cKO mice were injected with AAV vectors expressing EGFP-2A-C1ql2 and EGFP-2A-C1ql3, respectively. As control, Bcl11b cKO and control mice were injected with an AAV expressing EGFP. For the KD of C1ql2, the DG of 60-day-old C57BL/6JRj mice were injected with AAV 4x shC1ql2-EGFP, expressing 4 shRNAs against C1ql2 or control AAV 4xshNS-EGFP, expressing 4x non-sense shRNAs. For pan-neurexin KO, the DG of 60 days old Nrxn123flox/flox mice were injected with an AAV expressing EGFP-Cre or a control AAV expressing EGFP-Cre.Y324F, an inactive Cre. All AAVs were produced by the Viral Vector Facility of the Neuroscience Center Zurich on request. The four selected non-sense shRNAs and the four shC1ql2 sequences were checked for and presented with no off-target bindings on the murine exome with up to two mismatches by siRNA-Check (http://projects.insilico.us/SpliceCenter/siRNACheck). The mice were anesthetized with 5% isoflurane and placed in a mouse stereotaxic apparatus. During the entire procedure, anesthesia was maintained by constant administration of 2.2% isoflurane. Eye ointment was applied to prevent eyes from drying. For electrophysiological experiments, mice were subcutaneously injected with buprenorphine hydrochloride (0.1 mg/kg, Temgesic, Indivior) 30 minutes before and 3 hours after each surgery. For all other experiments, Butorphanol (Livisto) and Meloxicam (Boehringer-Ingelheim; 5 μg/g) were injected subcutaneously and the local anesthetic Bupivacaine (Puren; 5 μg/g) was injected subcutaneously at the incision site. After 10 min the head of the mouse was shaved and disinfected and an incision was made in the skin. Targeted injection sites were identified and a small craniotomy was performed for each site. The injector was placed at the individual sites and the viral solution was injected at 100 nL/min, with a 5-10 min recovery before removing the injector. After injections at all sites the incision was sutured and the animal was monitored for recovery from anesthesia, after which it was returned to its home cage. For histological and EM analyses of MFS, AAV were injected at three sites per hemisphere with the following coordinates (Bregma: AP 0; ML: 0; DV:0): AP –2 mm; ML ±1 mm; DV –2 mm. AP –2.5 mm; ML ±1.5 mm; DV –1.8 mm. AP –3.1 mm; ML ±2, DV –2.2 mm. For electrophysiological analyses, AAV were injected at two dorsoventral coordinates per hemisphere: AP –3.0 mm; ML ±3.25 mm; DV –2.4 and –2.8 mm. 200-300 nL of AAV (1e12 vg/mL) were injected in each location.

RNA isolation and quantitative real-time PCR

All procedures were performed in an RNase-free environment. Animals were sacrificed under deep CO2-induced anesthesia, brains were quickly dissected in ice-cold PBS, cryopreserved in 20% sucrose overnight, frozen in OCT compound (Polysciences), and stored at −80 °C. 20 μm thick coronal sections were collected on UV-treated and 0.05% poly-L-lysine coated membrane-covered PEN slides (Zeiss), fixed for 1 min in ice-cold 70% EtOH, incubated for 45 sec in 1% cresyl violet acetate solution (Waldeck) and washed for 1 min each in 70% EtOH and 100% EtOH. Sections were briefly dried on a 37 °C warming plate and immediately processed. The granule cell layer of the DG was isolated by laser capture microdissection using a PALM MicroBeam Rel.4.2 (Zeiss). RNA was isolated from the collected tissue using Rneasy Micro Kit (Qiagen) and reverse transcribed using the SensiFast cDNA Synthesis Kit (Bioline). Quantitative real-time PCR was performed in triplets for each sample using the LightCycler DNA Master SYBR Green I Kit in a LightCycler 480 System (Roche). The relative copy number of Gapdh RNA was used for normalization. Data were analyzed using the comparative CT method (Schmittgen & Livak, 2008).

Western blots

Briefly, hippocampi from freshly removed brains were dissected in ice-cold PBS, collected in Lysis Buffer (50mM Tris pH 7.5, 150mM NaCl, 0.5% sodium deoxycholate, 1% triton-X100, 0.1% SDS) and manually homogenated. Samples were centrifuged for 25 min at 13200 rpm at 4 °C and the supernatant was collected. Protein concentration was calculated with Bradford assay. Protein suspension containing 40 μg of protein was mixed 1:1 with 2x SDS loading dye (62.5mM Tris, 10% Glycerol, 5% β-mercaptoethanol, 80mM SDS, 1.5mM bromophenol blue), boiled at 95 °C for 5 min, separated by SDS-PAGE and electrophoretically transferred onto PVDF membranes (Merck). Membranes were blocked with 5% non-fat milk (Sigma-Aldrich), incubated with mouse anti-β-actin (1:5000; Sigma-Aldrich) and rabbit anti-C1ql2 (1:500; Sigma-Aldrich), followed by Peroxidase-conjugated secondary antibodies (Jackson ImmunoResearch) and developed with Pierce ECL Western Blotting Substrate (ThermoScientific). Signal was detected with ChemiDoc Imaging System (BioRad) and analyzed with Image Lab Software (BioRad). Protein signal was normalized with the signal of β-actin.

Immunohistochemistry and RNA in situ Hybridization

Animals were sacrificed under deep CO2-induced anesthesia, and brains were dissected in ice-cold 1x PBS, and either fixed for 4 h in 4% PFA in PBS at 4 °C and cryopreserved in 20% sucrose in PBS overnight at 4 °C or directly cryopreserved and then frozen in OCT compound (Polysciences). Sections were prepared at 14 μm. The unfixed sections were postfixed with 4% PFA in 1x PBS for 20 min. Heat-induced antigen retrieval in 10 mM citrate buffer (pH 6.0) was performed for fixed sections. Sections were blocked at RT for 1 h in 1x PBS containing 0.1% TritonX-100 and 10% horse serum, and incubated overnight at 4 °C with primary antibodies, followed by a 90 min incubation with secondary antibodies. Sections were counterstained with DAPI (Molecular Probes). The following primary antibodies were used on fixed sections: guinea pig anti-Bcl11b (1:1000; Simon et al., 2012), rabbit anti-C1ql2 (1:1000; Invitrogen), rabbit anti-C1ql2 (1:500; Sigma-Aldrich), chicken anti-GFP (1:2000; Abcam) and rabbit anti-C1ql3 (1:500; Biozol). Primary antibodies used on unfixed sections: mouse anti-vGlut1 (1:100; Synaptic Systems), guinea pig anti-Homer1 (1:250; Synaptic Systems), rabbit anti-C1ql2 (1:1000; Invitrogen) and rabbit anti-ZnT3 (1:200; Synaptic Systems). All fluorescent secondary antibodies were purchased from Jackson ImmunoResearch and used at 1:500 dilution. Hybridizations were performed with DIG-labelled riboprobes on 14 μm thick sections.

Transmission electron microscopy

Animals were sacrificed through CO2-inhalation and immediately perfused transcardially with 0.9% NaCl for 1 min, followed by a fixative solution of 1.5% glutaraldehyde (Carl Roth) and 4% PFA in 0.1M PB pH 7.2 for 13 min. Brains were dissected and postfixed in the fixative solution for 4 h at 4 °C. Ultrathin sections (60 nm) were prepared and stained with lead citrate. Images were acquired using a transmission electron microscope LEO 906 (Zeiss) with a sharp-eye 2k CCD camera and processed with ImageSP (Tröndle). Synapse score (De Bruyckere et al., 2018) was calculated according to the following criteria: 0-5 vesicles above the active zone = 0; 5-20 vesicles = 1; small group of vesicles (≤ 200000 nm2) with distance between density and closest vesicle > 100 nm = 2; small group of vesicles (≤ 200000 nm2) with distance between density and closest vesicle ≤ 100 nm = 3; big group of vesicles (> 200000 nm2) with distance between density and closest vesicle > 100 nm = 4; big group of vesicles (> 200000 nm2) with distance between density and closest vesicle ≤ 100 nm = 5. Synapses from approximately 30 MFB per animal were analyzed. Vesicles with a distance ≤ 5 nm from the plasma membrane were considered docked (Kusick et al., 2022; Vandael et al., 2020). Approximately 100 AZ per animal were analyzed.

Electrophysiological recordings and data analysis

Animals were sacrificed under deep CO2-induced anesthesia at 4 months. Brains were quickly removed and placed in ice-cold modified ACSF containing (in mM) 92 N-methyl-D-glucamine (NMDG), 2.5 KCl, 1.2 NaH2PO4, 30 NaHCO3, 20 HEPES, 25 glucose, 5 Na-ascorbate, 2 thiourea, 3 Na-pyruvate, 10 MgSO4, 0.5 CaCl2, 6 N-acetyl-L-cysteine (NAC), saturated with carbogen gas (95% O2 and 5% CO2, pH 7.4) (Ting et al., 2014). 450 µm thick horizontal slices were cut using a vibratome slicer (Leica) at a defined angle to improve the preservation of mossy fibers (Bischofberger et al., 2006). After cutting, slices were transferred to a “Haas”-type interface chamber (Haas et al., 1979), where they were perfused with carbogen-saturated ACSF containing (in mM) 124 NaCl, 3 KCl, 2.3 CaCl2, 1.8 MgSO4, 10 glucose, 1.25 NaH2PO4, 26 NaHCO3 (pH 7.4 at 34°C) at a rate of 1.5 mL/min at 34 ± 1 °C. Slices were allowed to recover for a minimum of 1 h before the start of recordings.

Recordings were carried out by placing a glass micropipette (tip diameter 3–5 μm) filled with ACSF in the SL of the CA3b area. To induce MF field excitatory post-synaptic potentials (fEPSP), a bipolar electrode (Science Products) was placed within the hilus region of the DG. 0.1 ms pulses were delivered with an Iso-Flex stimulus isolator (AMPI) at 20 sec intervals. Putative mossy fiber signals were preliminarily identified using a 25 Hz train of five pulses. Input-output relationships were obtained by measuring the fiber volley amplitude and fEPSP slope in response to stimulations with intensities ranging from 3 to 40 V. For LTP recordings, stimulation intensity for each slice was adjusted to obtain a slope value of 20% (30% in the case of forskolin (Biomol) experiments) of the maximum fEPSP slope. LTP was induced by three trains of 100 stimulation pulses at 100 Hz (high frequency stimulation, HFS), repeated every 8 sec. 3 μM DCG-IV (Tocris Bioscience) was applied after each experiment, and only recordings displaying >70% reduction in putative MF-fEPSP slopes were used for analysis. fEPSPs were amplified 100x with an EXT 10-2F amplifier (npi electronics). Signals were low-pass filtered at 2 kHz and high-pass filtered at 0.3 Hz, digitized at 20 kHz with an analog-to-digital converter (Cambridge Electronic Design (CED)) and stored for offline analysis using Spike2 (v7) software (CED). Offline data analysis was performed on raw traces using Spike2. Slope values were measured from the linear part of the fEPSP rising phase by manually placing vertical cursors. Changes in fEPSP slopes were calculated as a percentage of the average baseline fEPSP ((average fEPSP slope in a given time interval after HFS – average fEPSP slope before HFS)/ (average fEPSP slope before HFS)).

DNA constructs

For expression, C1ql2 was cloned from mouse cDNA. A 6xHis-myc tag or GFP was attached to the N-terminus and the construct was cloned into the pSecTag2A vector (Invitrogen) in frame with the N-terminal IgK signal sequence. A stop codon was introduced directly after C1ql2. The K262E point mutation was introduced with the Q5 Site-Directed Mutagenesis Kit (New England Biolabs). pSecTag2A was used for control experiments. Rat Nrxn3α(25b+) cDNA (Ushkaryov & Sudhof, 1993) was inserted into an pSyn5 vector with human Synapsin promoter (Neupert et al., 2015) using BamHI and BglII. For the pan-neurexin KO and the control experiments, vectors with NLS-GFP-Cre or NLS-GFP-Cre.Y324F were used (Klatt et al., 2021; Wang et al., 2016). Expression vector for GluK2 was purchased from OriGene. All vectors were validated by sequencing (Eurofins Genomics).

Primary Hippocampal Cultures

Hippocampi were dissected from P0 mice in HBSS media, digested for 15 min with HBSS containing 0.1% Trypsin (Gibco) at 37 °C, dissociated in plating media (MEM supplemented with 0.6% glucose, 10% FBS, 1% penicillin/streptomycin, DnaseI 4 U/mL) and seeded on poly-L-Lysin precoated coverslips placed inside 12-well plates at 1.5×105 cells/mL. After 3 h the plating media was replaced with neuronal growth media (Neurobasal A supplemented with 2% B27, 2mM L-Glutamine, 1% penicillin/streptomycin, 1% N2 and 0.005% NGF). Cultures were kept at 37 °C under 5% CO2 atmosphere. The day of plating was considered as 0 days in vitro (DIV). At DIV3 and DIV7 80% of the medium was exchanged with fresh growth medium. At DIV9 the medium was exchanged with penicillin/streptomycin-free growth medium and at DIV10 neurons were transfected using Lipofectamine 2000 (Invitrogen). Briefly, a total of 200 μL transfection mix per well was prepared by first mixing 100 μL Opti-MEM with 4 μL Lipofectamine 2000 in one tube and 100 μL Opti-MEM with 3 μg DNA in a different tube. After 5 min both volumes were combined and the mixture was incubated for 20 min at RT. The transfection mix was then added dropwise to the neurons. After 3 h of incubation, the medium was exchanged with fresh growth medium.

HEK293 Cell Culture

Human embryonic Kidney (HEK) 293 cells were obtained from ATCC and were maintained in DMEM supplemented with 10% fetal calf serum and 1% penicillin/streptomycin at 37 °C under 5% CO2 atmosphere. Cells were transfected using Lipofectamine 2000 according to the manufacturer’s instructions on the same day the neurons were transfected. Cells were incubated for at least 24 h before being used in co-culture experiments.

Neuronal and HEK293 co-culture and immunostaining

Transfected HEK293 cells were washed, dissociated, and resuspended in neuronal growth medium. 15×103 cells were added in each well containing DIV11 transfected neurons. HEK293 cells were co-cultured with the hippocampal primary neurons 2 days (DIV13 for neurons) before proceeding with immunostaining. Coverslips with cultured neurons and HEK293 cells were first fixed with 4% PFA in 1x PBS for 10 min at 4 °C, then washed 3x with 1 mL 1x PBS and blocked with 1x PBS containing 0.1% Triton X-100 and 10% horse serum for 1 h at RT. Primary antibodies were incubated overnight at 4 °C, followed by a 90 min incubation with secondary antibodies. Cells were counterstained with DAPI (Molecular Probes). The following primary antibodies were used: rabbit anti-myc-tag (1:2000; Abcam), guinea pig anti-vGlut1 (1:250; Synaptic Systems), chicken anti-GFP (1:2000; Abcam). All fluorescent secondary antibodies were purchased from Jackson ImmunoResearch and used at 1:500 dilution. For each condition, 25 cells per experiment were analyzed.

Structural protein modelling

The crystal structure of trimeric C1q-domains of mouse C1ql2 (Ressl et al., 2015) was used to predict a potential electrostatic binding site to splice insert 25 of Nrxn3α. An electrostatic surface map of the trimer was calculated using APBS (Jurrus et al., 2018). The K262E mutation was introduced using FoldX (foldxsuite.crg.eu) and was chosen in order to generate a negatively charged surface that would potentially be repulsive to Nrxn3α binding. Final models were visualized with PyMOL ((pymol.org).

Co-immunoprecipitation

HEK293 cells were transfected using Lipofectamine 3000 according to the manufacturer’s instructions and were incubated for at least 48 h. Cells were harvested and proteins were extracted in lysis buffer containing 25 mM Tris pH 7.4, 150 mM NaCl, 2 mM MgCl2, 1% Igepal, 5% Glycerol, 1x EDTA-free proteinase inhibitor, 0.5 mM DTT and 2.5 U/mL Benzonase. Protein A magnetic beads were washed 2x with PBS including 0.02% Tween-20 and were incubated on a rotating wheel in RT for 2 h with 2 μg of the following antibodies suspended in 200 μL 2x with PBS/0.02% Tween-100: rabbit anti-IgG (Cell Signal), rabbit anti-flag (Sigma-Aldrich) or rabbit anti-C1ql2 (Sigma-Aldrich). Beads were washed 2x with PBS/0.02% Tween-20, resuspended in 50 μL 2x with PBS/0.02% Tween-20 and 40 μg protein extract was added. Beads were incubated o/n on a rotating wheel at 4°C. Beads were thoroughly washed with 2x with PBS/0.02% Tween-20, resuspended in 2x SDS loading dye, and boiled for 10 min at 95°C. Western blot (see above) was performed. Briefly, membranes were blocked with 5% non-fat milk (Sigma-Aldrich) and incubated with mouse anti-flag M2 (1:2000; Sigma-Aldrich). Three independent experiments were carried out.

Image acquisition and analysis

All fluorescent images of sectioned hippocampal tissue were examined on a TCS SP5II confocal microscope (Leica) using LAS-X software and processed with Fiji (Schindelin et al., 2012). Overview images were acquired with a 20x objective. Synapse numbers and ZnT3+ puncta were quantified in the SL imaged with a 40x objective at x2 zoom. C1ql2 fluorescence intensity was quantified in the SL imaged with a 40x objective. Acquisition settings were kept constant for every sample and condition. All fluorescent images of co-cultured HEK293 cells were examined on a TCS SP8 confocal microscope (Leica) using LAS-X software and processed with Fiji. Images were acquired with a 40x objective at 4x zoom. As before, acquisition settings were kept constant for every sample and condition. Images were analyzed by masking transfected HEK293 cells and measuring the area of each mask covered by the chosen stain.

Quantification and statistical analysis

Statistical analysis and graph generation was done using Python 3. If samples met the criteria for normality, we used two-tailed unpaired t-test to compare two groups and one-way ANOVA for more than two groups. For non-normally distributed data Mann-Whitney u-test was used. Two-way ANOVA was used for examining the influence of two different categorical independent variables. If ANOVAs were significant, we used a post hoc Tukey’s multiple-comparisons test to compare groups (structural and expression data) or a post hoc Bonferonni’s comparison test (electrophysiological data). Data are presented as mean±SEM. Significance levels were set as indicated in figures: *p<0.05, **p<0.01, ***p<0.001.

Data availability

All relevant data are available from the authors.